---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format: html
editor: visual
---

# Importing data

Have to see which packages are needed later and add them.

```{r}
#| label: load libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(readr) # Data importation functions
library(stringr) # String and character text
library(dplyr)
library(ggplot2)
library(rstatix)
library(car)
library(emmeans) #Post hoc testing

theme_set(theme_bw())
#just writing this to commit something as it didn't work the first time
```

```{r}
#| label: importing raw data of traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/clean_data/cleaned_data.csv")


#Creating a data frame
cleaned_data <- as.data.frame(cleaned_data)
cleaned_data$calluna_NIN_type <- factor(cleaned_data$calluna_NIN_type)
cleaned_data$age_class <- factor(cleaned_data$age_class)

#Make age class in order pioneer, building, mature


```

```{r}
#| label: Data visualization and overview

summary(cleaned_data)
str(cleaned_data)

# Order relevant variables (useful for plotting and order based analyses)
# age_class sorted from youngest to oldest

cleaned_data$age_class <- factor((cleaned_data$age_class),
                             levels = c("Pioneer", 
                                        "Building", "Mature"))
# Check if changed successfully
levels(cleaned_data$age_class)


# Order NIN types
cleaned_data$calluna_NIN_type <- factor((cleaned_data$calluna_NiN_type),
                                    levels = c("BH", "FH", "TH"))
# Check if changed successfully
levels(cleaned_data$calluna_NiN_type)

## Define colour palette (used with scale fill manual)

NIN <- c("BH" = "#083508", "FH" = "#589758", "TH" = "#889000")
AGE <- c("Pioneer" = "#087E8B","Building" = "#EB5461", "Mature" = "#884000") 


```

Remember to exclude NA values if they confuse the code

# Wet mass(g) visualizations

```{r}
#| label: Wet mass visualizations

# Data frame with mean, median and se

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE), 
            median_wet_mass= median.default(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()), 
            n = n(), .groups = "drop")

#Histograms

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(bins = 30, color = "lightblue", fill = "midnightblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")



#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Wet mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()+ 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)


#Interactive plots


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  theme_minimal()+ 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)



#Density Plot 

ggplot(cleaned_data, aes(x = wet_mass, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
    facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot wet mass", x = "Wet mass") +
  theme_minimal()





```

# Dry mass(g) visualizations

```{r}
#| label: Dry mass visualization

# Data frame with mean and se

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lavender") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")


#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Dry mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal() +
  labs(x = "Age class", y = "Dry mass (g)")

#Interactive plot

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line(linewidth=1) + geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_dry_mass, aes(x = calluna_NiN_type, y = mean_dry_mass, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#Density Plot 

ggplot(cleaned_data, aes(x = dry_mass, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot dry mass (g)", x = "Dry mass (g)") +
  theme_minimal()


```

# Height (cm)

```{r}
#| label: Height visualizations

# Data frame with mean and se

df_sum_height <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(bins = 30, color = "black", fill = "violet") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Height (cm)")


# violin plot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Height (cm)")

# Interactive plot


ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
   theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)

# swapped axis, NiN type on x-axis

ggplot(df_sum_height, aes(x = calluna_NiN_type, y = mean_height_cm, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
    theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)
  
#Density Plot 

ggplot(cleaned_data, aes(x = height_cm, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot height (cm)", x = "Height (cm)") +
  theme_minimal()


```

# Leaf thickness (mm) visualizations

```{r}
#| label: Leaf thickness visualizations

#create table with leaf thickness 1,2,3 combined
data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

# Data frame with mean and se

df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram



ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(bins = 30, color = "black", fill = "forestgreen") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Leaf thickness (mm)", y = "Number of Calluna individuals")





# Boxplot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NiN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)")

# violin plot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)") +
  theme_minimal()


# Interactive plot


ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
   theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_thickness, aes(x = calluna_NiN_type, y = mean_leaf_thickness, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
     theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)

#Density Plot 

ggplot(data_thickness, aes(x = leaf_thickness, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf thickness (mm)", x = "Leaf thickness (cm)") +
  theme_minimal()


```

# Leaf area (mm^2^) visualizations

```{r}
#| label: Leaf area visualizations

# Data frame with mean and se

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(bins = 30, color = "black", fill = "pink") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Leaf area (mm^2))", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (mm^2)")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NiN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal()+
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (mm^2)")




# Interactive plot


ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  ) 
 



# swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_area, aes(x = calluna_NiN_type, y = mean_leaf_area, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  )

#Density Plot 

ggplot(cleaned_data, aes(x = leaf_area, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf area (mm^2)", x = "Leaf area (mm^2)") +
  theme_minimal()

```

# LDMC visualizations

```{r}
#| label: LDMC visualizations

# Data frame with mean and se

df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")


# Histogram

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(bins = 30, color = "black", fill = "orange") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "LDMC", y = "Number of Calluna individuals")


# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal() +
  labs(x = "Age class", y = "LDMC")



# Interactive plot


ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
     theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "LDMC",
    color = "Age class"
  )


#swapped axis, NiN type on x-axis

ggplot(df_sum_LDMC, aes(x = calluna_NiN_type, y = mean_LDMC, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
    theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "LDMC",
    color = "Age class"
  )

#Density Plot 

ggplot(cleaned_data, aes(x = LDMC, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot LDMC", x = "LDMC") +
  theme_minimal()




```

# SLA visualizations

```{r}
#| label: SLA visualizations

# Data frame with mean and se


df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")




# Boxplot


ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "SLA")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal() +
  labs(x = "Age class", y = "SLA")


# Interactive plot


ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Age class",
    y = "SLA",
    color = "Calluna NiN type"
  )

#swapped axis, NiN type on x-axis

ggplot(df_sum_SLA, aes(x = calluna_NiN_type, y = mean_SLA, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "SLA",
    color = "Age class"
  )


#Density Plot 

ggplot(cleaned_data, aes(x = SLA, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot SLA", x = "SLA") +
  theme_minimal()






```

# 

# Testing for normality

The dependent variable should be approximately normally distributed within each group. While ANOVA is relatively robust to moderate violations of normality, especially with larger sample sizes, extreme departures can affect validity.

```{r}
#| label: testing age class and NiN type for normality with shapiro-wilk, qq plots, and residuals

#Find if data meets assumptions for a two-way ANOVA and if transformations are needed by using shapiro-wilk test and qq plot



########################################### Wet mass
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(wet_mass) %>% filter(p<0.05)

#sort by p-values above 0.05, which indicates normality. Thought it would be eaier to filter out which relationships can be used for the ANOVA, or maybe which we should transform before ANOVA?



# QQ plot is generally more informative than the Shapiro test because large datasets often make Shapiro significant even when deviations are minor.

#Combined plot for age class*NiN type, for every combination type, better overview

ggplot(cleaned_data, aes(sample = wet_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()


#Linear model with interaction of age class and NiN type
lm_wet_mass <- lm(wet_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_wet_mass)



#Checking residuals 
#W = 0.98438, p-value = 0.001733, indicates non-normality
qqnorm(residuals(lm_wet_mass))
qqline(residuals(lm_wet_mass), col = "red")
plot(lm_wet_mass, which = 3)

shapiro.test(residuals(lm_wet_mass))





######################################### Dry mass
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(dry_mass) %>% filter(p>0.05)

#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = dry_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()



#Linear model with interaction of age class and NiN type
lm_dry_mass <- lm(dry_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_wet_mass)


#Checking residuals
# W = 0.97974, p-value = 0.0001981 indicates data is not normally distributed, even if it appears so
qqnorm(residuals(lm_dry_mass))
qqline(residuals(lm_dry_mass), col = "red")
plot(lm_dry_mass, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_dry_mass))




########################################## Height
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(height_cm) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = height_cm)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()



#Linear model with interaction of age class and NiN type
lm_height_cm <- lm(height_cm ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_height_cm)




#Checking residuals
#W = 0.98229, p-value = 0.0006273, indicates non-normality
qqnorm(residuals(lm_height_cm))
qqline(residuals(lm_height_cm), col = "red")
plot(lm_height_cm, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_height_cm))





########################################### Leaf area
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_area) %>% filter(p>0.05)



#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = leaf_area)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#Linear model leaf area with interaction of age class and NiN type

lm_leaf_area <- lm(leaf_area ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_leaf_area)



#Checking residuals
#W = 0.98409, p-value = 0.001577, indicates non-normality

qqnorm(residuals(lm_leaf_area))
qqline(residuals(lm_leaf_area), col = "red")
plot(lm_leaf_area, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_leaf_area))






##################################### Leaf thickness
data_thickness %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_thickness) %>% filter(p>0.05)




#Combined plot for age class*NiN type, for every combination type

ggplot(data_thickness, aes(sample = leaf_thickness)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()



#Linear model with interaction of age class and NiN type
lm_leaf_thickness <- lm(leaf_thickness ~ age_class*calluna_NiN_type, data = data_thickness)
summary(lm_leaf_thickness)


#checking residuals
#W = 0.96656, p-value = 6.664e-14, not normally distributed
qqnorm(residuals(lm_leaf_thickness))
qqline(residuals(lm_leaf_thickness), col = "red")
plot(lm_leaf_thickness, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_leaf_thickness))






###################################### SLA
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(SLA) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = SLA)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()


#Linear model with interaction of age class and NiN type
lm_SLA <- lm(SLA ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_SLA)


#checking residuals
#W = 0.96722, p-value = 1.639e-06, not normally distributed
qqnorm(residuals(lm_SLA))
qqline(residuals(lm_SLA), col = "red")
plot(lm_SLA, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_SLA))





#################################### LDMC
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(LDMC) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = LDMC)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#Checking residuals
#W = 0.98146, p-value = 0.0004388, not normally distributed

#Linear model with interaction of age class and NiN type
lm_LDMC <- lm(LDMC ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_LDMC)



#Checking residuals
#W = 0.98146, p-value = 0.0004388, not normally distributed
qqnorm(residuals(lm_LDMC))
qqline(residuals(lm_LDMC), col = "red")
plot(lm_LDMC, which = 3, add.smooth = FALSE)

shapiro.test(residuals(lm_LDMC))

```

# Levenetest for checking variance -not so relevant

Assumption for ANOVA is homogeneity of variance, so it is roughly equal cross all groups. Ensures that the pooled variance estimate used in ANOVA calculations is appropriate.

```{r}
#| label: levene test to check for homogeneity of variance as ANOVA requires roughly equal variance across all groups

#wet mass, result significant, variance is not equal


leveneTest(
  wet_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Dry mass, result non-significant, variance is equal

leveneTest(
  dry_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Height, results significant, variance is not equal

leveneTest(
  height_cm ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf area, results significant, variance is not equal

leveneTest(
  leaf_area ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf thickness, results significant, variance is not equal

leveneTest(
  leaf_thickness ~ interaction(age_class, calluna_NiN_type),
  data = data_thickness
)

#LDMC, results not significant, variance is equal

leveneTest(
  LDMC ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#SLA, results significant, variance is not equal

leveneTest(
  SLA ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)
```

# Account for random effects and data clustering with linear mixed effects models

→ likely fixed effects with interactions: trait \~  NIN\*age class, effect of NiN type on trait also depends on effect of age class, and possibly plant sampling groups

-Plant ID and leaf bulks are grouping factors that should be used as random effects to account for non-independence. To account for “within-individual” grouping. These are nested groups with lower level units contained within higher levels

→ often mixed-effects models as factors with few levels are fixed (NiN, age class) and factors with many levels are random (Plant 1-7 grouping)

\
-Linear mixed effects models, lme4 package, diff model, but same anova, summary etc. 1I plantID, not for height due to low replicate number

```{r}
#| label: lme4 modelling
#| 
```

# ANOVA and Post-hoc testing

All results show significance between age class, NiN and the interaction of the two with ANOVA

Post hoc test with more than two levels, emmeans package

```{r}
#| label: ANOVA

#Wet mass

anova(lm_wet_mass)

(emmeans_wet_mass<- emmeans(lm_wet_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_wet_mass)

#Dry mass

anova(lm_dry_mass)

(emmeans_dry_mass<- emmeans(lm_dry_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_dry_mass)

#Height

anova(lm_height_cm)

(emmeans_height_cm<- emmeans(lm_height_cm, ~ age_class * calluna_NiN_type))
pairs(emmeans_height_cm)

#Leaf area

anova(lm_leaf_area)

(emmeans_leaf_area<- emmeans(lm_leaf_area, ~ age_class * calluna_NiN_type))
pairs(emmeans_leaf_area)

#Leaf thickness

anova(lm_leaf_thickness)

(emmeans_leaf_thickness<- emmeans(lm_leaf_thickness, ~ age_class * calluna_NiN_type))
pairs(emmeans_leaf_thickness)

#LDMC

anova(lm_LDMC)

(emmeans_LDMC<- emmeans(lm_LDMC, ~ age_class * calluna_NiN_type))
pairs(emmeans_LDMC) 



#SLA

anova(lm_SLA)

(emmeans_SLA<- emmeans(lm_SLA, ~ age_class * calluna_NiN_type))
pairs(emmeans_SLA) 
```

# Data transformation -probably not necessary as most data follows a somewhat normal distribution

**-data often natural log transformed and scaled** to fairly compare across traits, we can statistically contrast the different traits side-by-side: makes skewed distributions more normal by pulling extreme outliers inward; allows interpretation in multiplicative steps (1x increase → 10x increase etc.); prevents heteroscedasticity (data artificially widening as mean increases); when comparing multiple traits, scaling makes sure a change of one unit in one trait means the same across all variables

-transform traits individually and do back-transformation

```{r}
#| label: data transformation
```

# Multivariate analysis

Because traits are not independent, we should investigate interactions between them

```{r}
#| label: Multivariate analysis

#manova_mod <- manova(
 # cbind(
  #  dry_mass,
   # height,
    #leaf_area,
    #leaf_thickness,
    #SLA,
    #LDMC
  #) ~ age_class * NiN_type,
  #data = cleaned_data
#)

```

# Presenting results -interactive plots

-add titles eventually

```{r}
#| label: making summary plots for presenting results in a tidy manner


#Excluding NAs from the clalculations of the mean as they prevent results, na.omit()



#Wet mass


ggplot(df_sum_wet_mass, 
       aes(x = calluna_NiN_type, y = mean_wet_mass, colour = age_class,
           ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass)) +
  # this adds the mean
  geom_point(size = 3) +
  # this adds the error bars
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Wet mass (g)") + 
  # use a more professional theme
  theme_bw()

ggplot(df_sum_wet_mass, 
       aes(x = calluna_NiN_type, y = mean_wet_mass, fill = age_class,
           ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass)) +
  # this adds the mean
  geom_col(position = position_dodge()) +
  # this adds the error bars
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  # controlling the appearance
  xlab("NiN type") + ylab("Wet mass (g)")



#Dry mass

pos <- position_dodge(0.15). #to dodge the position of plotted points if overlapping

ggplot(df_sum_dry_mass, 
       aes(x = calluna_NiN_type, y = mean_dry_mass, colour = age_class,
           ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass)) +
  geom_point(size = 3, position = pos) +
  geom_errorbar(width = 0.1, position = pos) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Dry mass (g)") + 
  theme_bw()


#Height


ggplot(df_sum_height, 
       aes(x = calluna_NiN_type, y = mean_height_cm, colour = age_class,
           ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Height (cm)") + 
  theme_bw()


#Leaf area


ggplot(df_sum_leaf_area, 
       aes(x = calluna_NiN_type, y = mean_leaf_area, colour = age_class,
           ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Leaf area (mm^2)") + 
  theme_bw()


#Leaf thickness

ggplot(df_sum_leaf_thickness, 
       aes(x = calluna_NiN_type, y = mean_leaf_thickness, colour = age_class,
           ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Leaf thickness (mm)") + 
  theme_bw()


#LDMC

ggplot(df_sum_LDMC, 
       aes(x = calluna_NiN_type, y = mean_LDMC, colour = age_class,
           ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("LDMC") + 
  theme_bw()


#SLA

ggplot(df_sum_SLA, 
       aes(x = calluna_NiN_type, y = mean_SLA, colour = age_class,
           ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("SLA") + 
  theme_bw()
```
