---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format: html
editor: visual
---

# Importing data

Have to see which packages are needed later and add them.

```{r}
#| label: load libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(readr) # Data importation functions
library(stringr) # String and character text
library(dplyr)
library(ggplot2)
library(rstatix)
library(car)
library(emmeans) #Post hoc testing
library(afex) #linear mixed models
library(lme4) #linear mixed models
library(sdamr) #linear mixed models


theme_set(theme_bw())



```

```{r}
#| label: importing raw data of traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/clean_data/cleaned_data.csv")


#Creating a data frame
cleaned_data <- as.data.frame(cleaned_data)
cleaned_data$calluna_NIN_type <- factor(cleaned_data$calluna_NIN_type)
cleaned_data$age_class <- factor(cleaned_data$age_class)


```

```{r}
#| label: Data visualization and overview

summary(cleaned_data)
str(cleaned_data)

# Order relevant variables (useful for plotting and order based analyses)
# age_class sorted from youngest to oldest

cleaned_data$age_class <- factor((cleaned_data$age_class),
                             levels = c("Pioneer", 
                                        "Building", "Mature"))
# Check if changed successfully
levels(cleaned_data$age_class)


# Order NIN types
cleaned_data$calluna_NIN_type <- factor((cleaned_data$calluna_NiN_type),
                                    levels = c("BH", "FH", "TH"))
# Check if changed successfully
levels(cleaned_data$calluna_NiN_type)

## Define colour palette (used with scale fill manual)

NIN <- c("BH" = "#083508", "FH" = "#589758", "TH" = "#889000")
AGE <- c("Pioneer" = "#087E8B","Building" = "#EB5461", "Mature" = "#884000") 




```

# Wet mass(g) visualizations

```{r}
#| label: Wet mass visualizations

# Data frame with mean, median and se

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE), 
            median_wet_mass= median.default(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()), 
            n = n(), .groups = "drop")

#Histograms

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(bins = 30, color = "lightblue", fill = "midnightblue") +
  facet_grid(calluna_NiN_type ~ age_class) +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")



#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Wet mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)


#Interactive plots


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)

#Separating plant_nr for each factor: age class and NiN

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, group = plant_nr)) +
  # Draw the raw data points for every plant
  geom_point(aes(color = as.factor(plant_nr)), alpha = 0.3) +
  # Draw the predicted random slope line for every single plant
  #geom_line(aes(y = wet_mass, color = as.factor(plant_nr)), size = 0.8) +
  # Split the graph into panels by Calluna NiN Type!
  facet_wrap(~ calluna_NiN_type) +
  labs(title = "Effect of plant nr across All Groups",
       x = "Age Class", 
       y = "Wet Mass",
       color = "Plant Number")



#Density Plot 

ggplot(cleaned_data, aes(x = wet_mass, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
    facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot wet mass", x = "Wet mass")



   


```

# Dry mass(g) visualizations

```{r}
#| label: Dry mass visualization

# Data frame with mean and se

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lavender") +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")


#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Dry mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Dry mass (g)")

#Interactive plot

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_dry_mass, aes(x = calluna_NiN_type, y = mean_dry_mass, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#Density Plot 

ggplot(cleaned_data, aes(x = dry_mass, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot dry mass (g)", x = "Dry mass (g)") 


```

# Height (cm)

```{r}
#| label: Height visualizations

# Data frame with mean and se

df_sum_height <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(bins = 30, color = "black", fill = "violet") +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Height (cm)")


# violin plot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Height (cm)")

# Interactive plot


ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)

# swapped axis, NiN type on x-axis

ggplot(df_sum_height, aes(x = calluna_NiN_type, y = mean_height_cm, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)
  
#Density Plot 

ggplot(cleaned_data, aes(x = height_cm, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot height (cm)", x = "Height (cm)")


```

# Leaf thickness (mm) visualizations

```{r}
#| label: Leaf thickness visualizations

#create table with leaf thickness 1,2,3 combined
data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

# Data frame with mean and se

df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram



ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(bins = 30, color = "black", fill = "forestgreen") +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf thickness (mm)", y = "Number of Calluna individuals")





# Boxplot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NiN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)")

# violin plot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)") 


# Interactive plot


ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_thickness, aes(x = calluna_NiN_type, y = mean_leaf_thickness, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)

#Density Plot 

ggplot(data_thickness, aes(x = leaf_thickness, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf thickness (mm)", x = "Leaf thickness (cm)") 


```

# Leaf area (mm^2^) visualizations

```{r}
#| label: Leaf area visualizations

# Data frame with mean and se

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(bins = 30, color = "black", fill = "pink") +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf area (mm^2))", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (mm^2)")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NiN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (mm^2)")




# Interactive plot


ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  ) 
 



# swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_area, aes(x = calluna_NiN_type, y = mean_leaf_area, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  )

#Density Plot 

ggplot(cleaned_data, aes(x = leaf_area, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf area (mm^2)", x = "Leaf area (mm^2)") 

```

# LDMC visualizations

```{r}
#| label: LDMC visualizations

# Data frame with mean and se

df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")


# Histogram

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(bins = 30, color = "black", fill = "orange") +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "LDMC", y = "Number of Calluna individuals")


# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")



# Interactive plot


ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "LDMC",
    color = "Age class"
  )


#swapped axis, NiN type on x-axis

ggplot(df_sum_LDMC, aes(x = calluna_NiN_type, y = mean_LDMC, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "LDMC",
    color = "Age class"
  )

#Density Plot 

ggplot(cleaned_data, aes(x = LDMC, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot LDMC", x = "LDMC") 




```

# SLA visualizations

```{r}
#| label: SLA visualizations

# Data frame with mean and se


df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")




# Boxplot


ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "SLA")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal() +
  labs(x = "Age class", y = "SLA")


# Interactive plot


ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Age class",
    y = "SLA",
    color = "Calluna NiN type"
  )

#swapped axis, NiN type on x-axis

ggplot(df_sum_SLA, aes(x = calluna_NiN_type, y = mean_SLA, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "SLA",
    color = "Age class"
  )


#Density Plot 

ggplot(cleaned_data, aes(x = SLA, group=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot SLA", x = "SLA") +
  theme_minimal()






```

# 

# Testing for normality

The dependent variable should be approximately normally distributed within each group. While ANOVA is relatively robust to moderate violations of normality, especially with larger sample sizes, extreme departures can affect validity.

-Shapiro-Wilk test reveals non-normality for the linear model, but since we have a large data set n\>50, the test is too sensitive and results might be a false negative, leading to rejection of normal data, when it is in reality fine to use.

-Both ANOVA and mixed models are robust to large datasets when it comes to deviating slightly from normality

```{r}
#| label: testing age class and NiN type for normality with shapiro-wilk, qq plots, and residuals

#Find if data meets assumptions for a two-way ANOVA and if transformations are needed by using shapiro-wilk test and qq plot



########################################### Wet mass
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(wet_mass) %>% filter(p<0.05)

#sort by p-values above 0.05, which indicates normality. Thought it would be eaier to filter out which relationships can be used for the ANOVA, or maybe which we should transform before ANOVA?



# QQ plot is generally more informative than the Shapiro test because large datasets often make Shapiro significant even when deviations are minor.

#Combined qqplot for age class*NiN type, for every combination type, better overview

ggplot(cleaned_data, aes(sample = wet_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Wet mass qq plot")


#Linear model with interaction of age class and NiN type
lm_wet_mass <- lm(wet_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_wet_mass)



#Checking residuals 
#W = 0.98438, p-value = 0.001733, indicates non-normality
shapiro.test(residuals(lm_wet_mass))

qqnorm(residuals(lm_wet_mass), main = "Q-Q Plot of Residuals from Linear Regression of wet mass")
qqline(residuals(lm_wet_mass), col = "red")

#Scale plot
plot(lm_wet_mass, which = 3, add.smooth = TRUE)







######################################### Dry mass
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(dry_mass) %>% filter(p>0.05)

#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = dry_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Dry mass qq plot")



#Linear model with interaction of age class and NiN type
lm_dry_mass <- lm(dry_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_wet_mass)


#Checking residuals
# W = 0.97974, p-value = 0.0001981 indicates data is not normally distributed, even if it appears so
shapiro.test(residuals(lm_dry_mass))

qqnorm(residuals(lm_dry_mass), main = "Q-Q Plot of Residuals from Linear Regression of dry mass")
qqline(residuals(lm_dry_mass), col = "red")

#Scale plot
plot(lm_dry_mass, which = 3, add.smooth = TRUE)






########################################## Height
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(height_cm) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = height_cm)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Height qq plot")



#Linear model with interaction of age class and NiN type
lm_height_cm <- lm(height_cm ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_height_cm)




#Checking residuals
#W = 0.98229, p-value = 0.0006273, indicates non-normality
shapiro.test(residuals(lm_height_cm))

qqnorm(residuals(lm_height_cm), main = "Q-Q Plot of Residuals from Linear Regression of height")
qqline(residuals(lm_height_cm), col = "red") #Plot is slightly heavy tailed, indicating data contains a few more extreme values than would be expected from a normal distribution

#Scale plot
plot(lm_height_cm, which = 3, add.smooth = TRUE)







########################################### Leaf area
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_area) %>% filter(p>0.05)



#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = leaf_area)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf area qq plot")

#Linear model leaf area with interaction of age class and NiN type

lm_leaf_area <- lm(leaf_area ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_leaf_area)



#Checking residuals
#W = 0.98409, p-value = 0.001577, indicates non-normality
shapiro.test(residuals(lm_leaf_area))

qqnorm(residuals(lm_leaf_area), main = "Q-Q Plot of Residuals from Linear Regression of leaf area")
qqline(residuals(lm_leaf_area), col = "red") #Plot is slightly heavy tailed, indicating data contains a few more extreme values than would be expected from a normal distribution

#Scale plot for linear model
plot(lm_leaf_area, which = 3, add.smooth = TRUE, title = "Linear model scale plot") 
  








##################################### Leaf thickness
#Shapiro test for data, lm test more relevant: data_thickness %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_thickness) %>% filter(p>0.05)




#Combined plot for age class*NiN type, for every combination type

ggplot(data_thickness, aes(sample = leaf_thickness)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf thickness qq plot")



#Linear model with interaction of age class and NiN type
lm_leaf_thickness <- lm(leaf_thickness ~ age_class*calluna_NiN_type, data = data_thickness)
summary(lm_leaf_thickness)


#checking residuals
#W = 0.96656, p-value = 6.664e-14, not normally distributed
shapiro.test(residuals(lm_leaf_thickness))

qqnorm(residuals(lm_leaf_thickness), main = "Q-Q Plot of Residuals from Linear Regression of leaf thickness")
qqline(residuals(lm_leaf_thickness), col = "red")

#Scale plot
plot(lm_leaf_thickness, which = 3, add.smooth = TRUE)





###################################### SLA
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(SLA) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = SLA)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "SLA qq plot")


#Linear model with interaction of age class and NiN type
lm_SLA <- lm(SLA ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_SLA)


#checking residuals
#W = 0.96722, p-value = 1.639e-06, not normally distributed
shapiro.test(residuals(lm_SLA))

qqnorm(residuals(lm_SLA), main = "Q-Q Plot of Residuals from Linear Regression of SLA")
qqline(residuals(lm_SLA), col = "red") #Plot is slightly heavy tailed, indicating data contains a few more extreme values than would be expected from a normal distribution

#Scale plot
plot(lm_SLA, which = 3, add.smooth = TRUE)







#################################### LDMC
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(LDMC) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = LDMC)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "LDMC qq plot")


#Linear model with interaction of age class and NiN type
lm_LDMC <- lm(LDMC ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_LDMC)



#Checking residuals
#W = 0.98146, p-value = 0.0004388, not normally distributed
shapiro.test(residuals(lm_LDMC))

qq_LDMC <- qqnorm(residuals(lm_LDMC), main = "Q-Q Plot of Residuals from Linear Regression of LDMC")
qqline(residuals(lm_LDMC), col = "red")

#Scale plot
scale_LDMC <- plot(lm_LDMC, which = 3, add.smooth = TRUE)






#scatter plot --does not work yet

scatter_LDMC<-ggplot(cleaned_data, aes(x = calluna_NiN_type, y = age_class)) +
  geom_point() +
  ggtitle("test") +
  geom_abline(slope = slope_LDMC, intercept = intercept_LDMC, color="red", size=1)
 


# Combine scatter plot and Q-Q plot using patchwork, could make a nice overview
library(patchwork)
qq_LDMC + scatter_LDMC
```

# Levenetest for checking variance -not so relevant

Assumption for ANOVA is homogeneity of variance, so it is roughly equal cross all groups. Ensures that the pooled variance estimate used in ANOVA calculations is appropriate.

```{r}
#| label: levene test to check for homogeneity of variance as ANOVA requires roughly equal variance across all groups

#wet mass, result significant, variance is not equal


leveneTest(
  wet_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Dry mass, result non-significant, variance is equal

leveneTest(
  dry_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Height, results significant, variance is not equal

leveneTest(
  height_cm ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf area, results significant, variance is not equal

leveneTest(
  leaf_area ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf thickness, results significant, variance is not equal

leveneTest(
  leaf_thickness ~ interaction(age_class, calluna_NiN_type),
  data = data_thickness
)

#LDMC, results not significant, variance is equal

leveneTest(
  LDMC ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#SLA, results significant, variance is not equal

leveneTest(
  SLA ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)
```

# Account for random effects and data clustering with lme4

→ likely fixed effects with interactions: trait \~  NIN\*age class, effect of NiN type on trait also depends on effect of age class, and possibly plant sampling groups

-Plant ID (plant nr) is a grouping factor that should be used as random effects to account for non-independence. To account for “within-individual” grouping. These are nested groups with lower level units contained within higher levels

→ often use mixed-effects models as factors with few levels are fixed (NiN, age class) and factors with many levels are random (Plant 1-7 grouping)

\
-Linear mixed effects models, lme4 package, diff model, but same anova, summary etc. 1I plantID, not for height due to low replicate number

-When you move to `lme4` for mixed-effects models, you have two types of normality to check:

- **Residual normality:** Just like ANOVA, the overall errors of the model should look fairly normal on a Q-Q plot.

- **Random effects normality:** The random intercepts or slopes you estimate for your groups (like subjects or schools) should also follow a normal distribution.

-check if continuous variables need rescaling, when model says it is over-parameterized (random effect structure complex)

```{r}
#| label: lme4 modelling

#data=cleaned_data
#dependent variable = plant trait
#fixed factors: NiN and age class
#Grouping factor: Plant_nr nested in age class

###### Wet mass

lme4_wet_mass <- lme4::lmer(wet_mass ~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)

summary(lme4_wet_mass)
plot(lme4_wet_mass)

qq_wet_mass_lme4 <- qqnorm(residuals(lme4_wet_mass), main = "Q-Q Plot of lme Residuals of Wet mass")
qqline(residuals(lme4_wet_mass), col = "red")

#Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean
#R message: Model is nearly unidentifiable: large eigenvalue ratio- Rescale variables? -But, might be due to mass values being so low


#Checking unusual values that may be genuine biological outliers in their group or misplaced decimal points, unit conversion mistakes, impossible values, transcription errors
#Find out if the same rows have unusual values for more traits, suggests real biological variation
#Do the inspected plants have something in common? age class, NiN, same collTime etc. 

r_wet_mass <- resid(lme4_wet_mass, type = "pearson")
idx_wet_mass <- which(abs(r_wet_mass) > 2 * sd(r_wet_mass))
idx_wet_mass

#Which plants drive the greatest residuals

cleaned_data[idx_wet_mass,
             c("plant_nr",
               "wet_mass",
               "age_class",
               "calluna_NiN_type",
               "collection_time")]


###### Dry mass

lme4_dry_mass <- lme4::lmer(dry_mass ~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)

summary(lme4_dry_mass)
plot(lme4_dry_mass)

qq_dry_mass_lme4 <- qqnorm(residuals(lme4_dry_mass), main = "Q-Q Plot of Residuals from lme of Dry mass")
qqline(residuals(lme4_dry_mass), col = "red")

#Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean




r_dry_mass <- resid(lme4_dry_mass, type = "pearson")
idx_dry_mass <- which(abs(r_dry_mass) > 2 * sd(r_dry_mass))
idx_dry_mass

#Which plants drive the greatest residuals

cleaned_data[idx_dry_mass,
             c("plant_nr",
               "dry_mass",
               "age_class",
               "calluna_NiN_type",
               "collection_time")]


###### Height -mixed model for height doesn't make sense as it is equal per plant



###### Leaf area

lme4_leaf_area <- lme4::lmer(leaf_area ~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)

summary(lme4_leaf_area).  #R message: Model is nearly unidentifiable: large eigenvalue ratio, might be due to outliers having smaller/larger values than rest of the data
plot(lme4_leaf_area)

qq_leaf_area_lme4 <- qqnorm(residuals(lme4_leaf_area), main = "Q-Q Plot of Residuals from lme of leaf area")
qqline(residuals(lme4_leaf_area), col = "red")

#Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean

r_leaf_area <- resid(lme4_leaf_area, type = "pearson")
idx_leaf_area <- which(abs(r_leaf_area) > 2 * sd(r_leaf_area))
idx_leaf_area

#Which plants drive the greatest residuals
 
cleaned_data[idx_leaf_area,
             c("plant_nr",
               "bulk_nr_leaves",
               "leaf_area",
               "age_class",
               "calluna_NiN_type",
               "collection_time")]


###### Leaf thickness

lme4_leaf_thickness <- lme4::lmer(leaf_thickness ~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=data_thickness, na.action = na.exclude)

summary(lme4_leaf_thickness) #R message: Model failed to converge: degenerate  Hessian with 1 negative eigenvalues -possibly means model is overfitted to the data or doesn't work well
plot(lme4_leaf_thickness)

qq_leaf_thickness_lme4 <- qqnorm(residuals(lme4_leaf_thickness), main = "Q-Q Plot of Residuals from lme of leaf thickness")
qqline(residuals(lme4_leaf_thickness), col = "red")

r_leaf_thickness <- resid(lme4_leaf_thickness, type = "pearson")
idx_leaf_thickness <- which(abs(r_leaf_thickness) > 2 * sd(r_leaf_thickness))
idx_leaf_thickness

#Which plants drive the greatest residuals

data_thickness[idx_leaf_thickness,
             c("plant_nr",
               "bulk_nr_leaves",
               "leaf_thickness",
               "age_class",
               "calluna_NiN_type",
               "collection_time")]


###### LDMC

lme4_LDMC <- lme4::lmer(LDMC ~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)

summary(lme4_LDMC)  #Model is nearly unidentifiable: large eigenvalue ratio- Rescale variables?
plot(lme4_LDMC)

qq_LDMC_lme4 <- qqnorm(residuals(lme4_LDMC), main = "Q-Q Plot of Residuals from lme of LDMC")
qqline(residuals(lme4_LDMC), col = "red")

r_LDMC <- resid(lme4_LDMC, type = "pearson")
idx_LDMC <- which(abs(r_LDMC) > 2 * sd(r_LDMC))
idx_LDMC

#Which plants drive the greatest residuals

cleaned_data[idx_LDMC,
               c("plant_nr",
                 "bulk_nr_leaves",
                 "LDMC",
                 "age_class",
                 "calluna_NiN_type",
                 "collection_time")]


###### SLA

lme4_SLA <- lme4::lmer(SLA~ age_class*calluna_NiN_type + (1|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)

summary(lme4_SLA) #R message: Model failed to converge: degenerate  Hessian with 1 negative eigenvalues -possibly means model is overfitted to the data or doesn't work well

#Scale plot to visually assess model
plot(lme4_SLA)

#Scale plot sorted by NiN type and age class, could do by individual plant ID too, but not sure what this plot contributes too

data_pred_SLA <- data.frame(
  predicted_SLA = predict(lme4_SLA),
  residual_SLA  = residuals(lme4_SLA),
  age_class     = cleaned_data$age_class,
  plant_nr      = cleaned_data$plant_nr,
  NiN_type      = cleaned_data$calluna_NiN_type
)

data_pred_SLA$Plant_ID <- interaction(
  data_pred_SLA$age_class,
  data_pred_SLA$plant_nr,
  data_pred_SLA$NiN_type,
  sep = "_"
)

ggplot(data_pred_SLA,
       aes(x = predicted_SLA,
           y = residual_SLA,
           colour = NiN_type)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 3)+
  facet_wrap(~age_class)


#Histogram of residuals, check for normal distribution
ggplot(data_pred_SLA,aes(x=residual_SLA)) + geom_histogram(bins=20, color="black")

#qq plot

qq_SLA_lme4 <- qqnorm(residuals(lme4_SLA), main = "Q-Q Plot of Residuals from lme of SLA")
qqline(residuals(lme4_SLA), col = "red")

r_SLA <- resid(lme4_SLA, type = "pearson")
idx_SLA <- which(abs(r_SLA) > 2 * sd(r_SLA))
idx_SLA

#Which plants drive the greatest residuals

cleaned_data[idx_SLA,
             c("plant_nr",
               "bulk_nr_leaves",
               "SLA",
               "age_class",
               "calluna_NiN_type",
               "collection_time")]



############## Does not yet work!


#Estimating lme model with random intercepts and slopes, allowing for the effect of plant grouping to vary across groups, making model more flexible and accurate

modg_SLA <- lme4::lmer(SLA~ age_class*calluna_NiN_type + (1+ age_class*calluna_NiN_type|age_class/plant_nr), data=cleaned_data, na.action = na.exclude)
summary(modg_SLA)

#Model fitting random intercepts and random slopes, but force their correlation to be zero without correlation between random intercept and slope effects, estimates variance of slope and intercepts
#R message: boundary (singular) fit - indicates model is overfitted
modr_SLA <- lme4::lmer(SLA~ age_class*calluna_NiN_type + (1+ age_class*calluna_NiN_type||age_class/plant_nr), data=cleaned_data, na.action = na.exclude)
summary(modr_SLA)


#Likelihood test ration with anova function to compare diff. models 
anova(modr_SLA,modg_SLA)
```

# ANOVA and Post-hoc testing

All results show significance between age class, NiN and the interaction of the two with ANOVA

Post hoc test with more than two levels, emmeans package

```{r}
#| label: ANOVA

#Wet mass

anova(lm_wet_mass)

(emmeans_wet_mass<- emmeans(lm_wet_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_wet_mass)

plot(emmeans_wet_mass)

#Dry mass

anova(lm_dry_mass)

(emmeans_dry_mass<- emmeans(lm_dry_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_dry_mass)

plot(emmeans_dry_mass)

#Height

anova(lm_height_cm)

(emmeans_height_cm<- emmeans(lm_height_cm, ~ age_class * calluna_NiN_type))
pairs(emmeans_height_cm)

#Leaf area

anova(lm_leaf_area)

(emmeans_leaf_area<- emmeans(lm_leaf_area, ~ age_class * calluna_NiN_type))
pairs(emmeans_leaf_area)

#Leaf thickness

anova(lm_leaf_thickness)

(emmeans_leaf_thickness<- emmeans(lm_leaf_thickness, ~ age_class * calluna_NiN_type))
pairs(emmeans_leaf_thickness)

#LDMC

anova(lm_LDMC)

(emmeans_LDMC<- emmeans(lm_LDMC, ~ age_class * calluna_NiN_type))
pairs(emmeans_LDMC) 



#SLA

anova(lm_SLA)

(emmeans_SLA<- emmeans(lm_SLA, ~ age_class * calluna_NiN_type))
pairs(emmeans_SLA) 
```

# Data transformation

**-data often natural log transformed and scaled** to fairly compare across traits, we can statistically contrast the different traits side-by-side: makes skewed distributions more normal by pulling extreme outliers inward; allows interpretation in multiplicative steps (1x increase → 10x increase etc.); prevents heteroscedasticity (data artificially widening as mean increases); when comparing multiple traits, scaling makes sure a change of one unit in one trait means the same across all variables

-transform traits individually and do back-transformation

```{r}
#| label: data transformation

#So far, it looks like it is only necessary to do for the height data

```

# Multivariate analysis

Because traits are not independent, we should investigate interactions between them

```{r}
#| label: Multivariate analysis

#manova_mod <- manova(
 # cbind(
  #  dry_mass,
   # height,
    #leaf_area,
    #leaf_thickness,
    #SLA,
    #LDMC
  #) ~ age_class * NiN_type,
  #data = cleaned_data
#)

```

# Presenting results -interactive plots

-add titles and significance eventually

```{r}
#| label: making summary plots for presenting results in a tidy manner


#Excluding NAs from the clalculations of the mean as they prevent results, na.omit()



#Wet mass


ggplot(df_sum_wet_mass, 
       aes(x = calluna_NiN_type, y = mean_wet_mass, colour = age_class,
           ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass)) +
  # this adds the mean
  geom_point(size = 3) +
  # this adds the error bars
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Wet mass (g)") 


ggplot(df_sum_wet_mass, 
       aes(x = calluna_NiN_type, y = mean_wet_mass, fill = age_class,
           ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass)) +
  # this adds the mean
  geom_col(position = position_dodge()) +
  # this adds the error bars
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  # controlling the appearance
  xlab("NiN type") + ylab("Wet mass (g)")



#Dry mass

#pos <- position_dodge(0.15). #to dodge the position of plotted points if overlapping

ggplot(df_sum_dry_mass, 
       aes(x = calluna_NiN_type, y = mean_dry_mass, colour = age_class,
           ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Dry mass (g)") 

ggplot(df_sum_dry_mass, 
       aes(x = calluna_NiN_type, y = mean_dry_mass, fill = age_class,
           ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("Dry mass (g)")


#Height


ggplot(df_sum_height, 
       aes(x = calluna_NiN_type, y = mean_height_cm, colour = age_class,
           ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Height (cm)") 

ggplot(df_sum_height, 
       aes(x = calluna_NiN_type, y = mean_height_cm, fill = age_class,
           ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("Height (cm)")

#Leaf area


ggplot(df_sum_leaf_area, 
       aes(x = calluna_NiN_type, y = mean_leaf_area, colour = age_class,
           ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Leaf area (mm^2)")

ggplot(df_sum_leaf_area, 
       aes(x = calluna_NiN_type, y = mean_leaf_area, fill = age_class,
           ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("Leaf area (mm^2)")


#Leaf thickness

ggplot(df_sum_leaf_thickness, 
       aes(x = calluna_NiN_type, y = mean_leaf_thickness, colour = age_class,
           ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("Leaf thickness (mm)") 

ggplot(df_sum_leaf_thickness, 
       aes(x = calluna_NiN_type, y = mean_leaf_thickness, fill = age_class,
           ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("Leaf thickness (mm)")



#LDMC

ggplot(df_sum_LDMC, 
       aes(x = calluna_NiN_type, y = mean_LDMC, colour = age_class,
           ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("LDMC") 

ggplot(df_sum_LDMC, 
       aes(x = calluna_NiN_type, y = mean_LDMC, fill = age_class,
           ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("LDMC")


#SLA

ggplot(df_sum_SLA, 
       aes(x = calluna_NiN_type, y = mean_SLA, colour = age_class,
           ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +  
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  xlab("NiN type") + ylab("SLA") 

ggplot(df_sum_SLA, 
       aes(x = calluna_NiN_type, y = mean_SLA, fill = age_class,
           ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(position = position_dodge(0.9), width=.2) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  xlab("NiN type") + ylab("SLA")


#Comparing results of emmeans post-hoc test, for visualizing the magnitude and direction of differences between group means
#A heatmap of estimates does not show: variability within groups, outliers, sample distributions,overlap among observations.
#Negative values → the first group has a lower mean than the second.
#Positive values → the first group has a higher mean than the second.
#Values near 0 → the groups are similar.

pairs(emmeans_SLA)
contrasts_SLA_df <- as.data.frame(pairs(emmeans_SLA))
head(contrasts_SLA_df)

library(tidyr)

contrasts_SLA_df <- contrasts_SLA_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )


ggplot(contrasts_SLA_df,
       aes(calluna_NiN_type, age_class, fill = estimate)) +
  geom_tile() +
  geom_point(
    data = subset(contrasts_SLA_df, p.value < 0.05),
    shape = 8
  )+
  scale_fill_gradient2(
    low = "darkblue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) 




ggplot(contrasts_SLA_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
   scale_fill_gradient2(midpoint = 0) +
  coord_equal()+
  scale_fill_gradient2(
    low = "darkblue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) 
```
