---
title: "DURIN 2025 Plant Functional Traits -Hydraulics"
author: "Sonya R Geange" and "Maiken anabel Ransve"
format: html
editor: visual
---

## Import Data

```{r}
#| label: loading libraries
#| message: false 

# installing / loading packages 

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(janitor) # Has tools for data cleaning 
library(lubridate) # Date and time formatting
library(readr) # Data importation functions
library(stringr) # String and character text
library(dplyr)
```

```{r}
#| label: importing raw data of traits
#| message: false 

# Eventually this will download from the DURIN OSF
# For now, download the excel from google sheets and save in your projects 'raw_data' folder.
# If using projects just need "./filenam.xlsx'
# If not in projects, need full path name

# Read plant functional trait data (excluding leaf area)
raw_data <- read_excel("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/2025 DURIN Plant Functional Traits - Hydraulics.xlsx", 
  sheet = "Data", # Choose the sheet from excel
  na = c("", "NA")) # NAs are actually imported as NAs, not as characters 
```

```{r}
#| label: clean trait dataset to match DURIN formatting
 # Change some column names to snake_case standard

# Change any column names to match DURIN standards - some outlined here (https://github.com/Durin-project/Durin_data)

# Some information will be specific to traits, fluorescence etc.

# using slice() to remove the first three row of example data (ALWAYS CHECK!)

raw_data <- raw_data %>%

# New name then old name of column, including snake case ('_' and lower letters)

rename(site_name = siteID,
age_class = ageClass,
collection_time = CollTime,
paired_with = PairedWith,
burn_year_original = burn_year,
calluna_NiN_type = NiN_type,

# Remove units from name as this should be in the data dictionary

wet_mass = wet_mass_g,
leaf_thickness_1 = leaf_thickness_1_mm,
leaf_thickness_2 = leaf_thickness_2_mm,
leaf_thickness_3 = leaf_thickness_3_mm,
dry_mass = dry_mass_g,
stem_diameter_base = stem_diameter_base_mm,
stem_diameter_mid = stem_diameter_mid_mm,
stem_diameter_tip = stem_diameter_tip_mm) %>%

slice(-(1:1)) # Remove first row as it is an example entry for the dataset (excluding headers)
```

```{r}


# Check variable types (character, numeric etc)
glimpse(raw_data)

# Changing classes of all variables that should be in class factor
raw_data <- raw_data %>%  
  mutate(across(c(month, site_name, species, project_2025, project, age_class, burn_year_original, calluna_NiN_type, plant_nr, leaf_age, sample, paired_with, calluna_shoot_type,scanned), as.factor))


# Changing classes of variables that should be in class dbl/numeric
raw_data <- raw_data %>% 
    mutate(across(c(wet_mass, leaf_thickness_1, leaf_thickness_2, leaf_thickness_3,
                    stem_diameter_base, stem_diameter_mid, stem_diameter_tip, dry_mass), as.numeric))

# Changing classes of all variables that should be in class integers
raw_data <- raw_data %>% 
    mutate(across(c(day, bulk_nr_leaves, leaf_nr), as.integer))

# Recheck data types again
glimpse(raw_data) 


# Order relevant variables (useful for plotting and order based analyses)
# age_class sorted from youngest to oldest

raw_data$age_class <- factor((raw_data$age_class),
                         levels = c("Pioneer", 
                                    "Building", "Mature"))
# Check if changed successfully
levels(raw_data$age_class)


# Order NIN types
raw_data$calluna_NIN_type <- factor((raw_data$calluna_NiN_type),
                         levels = c("BH", "FH", "TH"))
# Check if changed successfully
levels(raw_data$calluna_NiN_type)



# Converting day, month, year column into one column on the format yyyy_mm_dd
# removing the day and year columns, keep month for data cleaning checking for now      

raw_data <- raw_data %>% 
  mutate(date  = ymd(paste(year, as.character(month), day)), .after = 1)  %>% 
  select(-c(day, month, year)) 



glimpse(raw_data) 


# Create a unique plant_ID given age_class, NiN type, and plant_nr
raw_data <- raw_data %>%   # assuming you already created Month/date_day earlier
  mutate(
    unique_plantID = paste(age_class, calluna_NiN_type, plant_nr, sep = "_")
  )


# NEW Columns:
# We need to match the DURIN data standards, which includes adding columns for location

# Create new columns for coordinates 
raw_data$latitude_N         <- as.numeric(NA)
raw_data$longitude_E        <- as.numeric(NA)


# Fill in coordinates

# Pioneer sampled from BH: 60.699655, 5.092279
# Pioneer sampled from FH: 60.699426, 5.091988
# Pioneer sampled from TH: 60.699467, 5.092246
# Building sampled from BH: 60.699005, 5.101941
# Building sampled from FH: 60.698828, 5.102044
# Building sampled from TH: 60.698798, 5.101822
# Mature sampled from BH: 60.697479, 5.096821
# Mature sampled from FH: 60.697325, 5.096801
# Mature sampled from TH: 60.697461, 5.096151

raw_data$latitude_N <- case_when(raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "BH" ~ 60.699655,
raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "FH" ~ 60.699426,
raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "TH" ~ 60.699467,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "BH" ~ 60.699005,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "FH" ~ 60.698828,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "TH" ~ 60.698798,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "BH" ~ 60.697479,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "FH" ~ 60.697325,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "TH" ~ 60.697461)

raw_data$longitude_E <- case_when(raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "BH" ~ 5.092279,
raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "FH" ~ 5.091988,
raw_data$age_class == "Pioneer" & raw_data$calluna_NIN_type == "TH" ~ 5.092246,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "BH" ~ 5.101941,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "FH" ~ 5.102044,
raw_data$age_class == "Building" & raw_data$calluna_NIN_type == "TH" ~ 5.101822,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "BH" ~ 5.096821,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "FH" ~ 5.096801,
raw_data$age_class == "Mature" & raw_data$calluna_NIN_type == "TH" ~ 5.096151)


```

```{r}
#| label: import leaf area dataset tidy and match to trait dataset


# Read in files
data_base_r1 <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/Leaf_scans/DURIN_Calluna_leaf_area_Hydraulics_R1.csv", show_col_types = FALSE) |>
  column_to_rownames(var = "...1")
data_base_r1$source_file <-"DURIN_Calluna_leaf_area_Hydraulics_R1.csv"

data_scaled_r1 <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/Leaf_scans/DURIN_Calluna_leaf_area_Hydraulics_R1_ScaledDifferently.csv") |>
  column_to_rownames(var = "...1")
data_scaled_r1$source_file <-"DURIN_Calluna_leaf_area_Hydraulics_R1_ScaledDifferently.csv"

data_problem_r1 <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/Leaf_scans/DURIN_Calluna_leaf_area_Hydraulics_R1_ScaledDifferently_ProblemImages.csv")|>
  column_to_rownames(var = "...1")
data_problem_r1$source_file <-"DURIN_Calluna_leaf_area_Hydraulics_R1_ScaledDifferently_ProblemImages.csv"

data_base_r2 <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/Leaf_scans/DURIN_Calluna_leaf_area_Hydraulics_R2.csv")|>
  column_to_rownames(var = "...1")
data_base_r2$source_file <-"DURIN_Calluna_leaf_area_Hydraulics_R2.csv"

leaf_area_raw <- bind_rows(data_base_r1, data_scaled_r1, data_problem_r1, data_base_r2)

# Choose correct value per envelope_ID based on the source file, i.e. edited or not
# Create a 'cleaned' file flagging which images were 'base', 'scaled' issues, or scaled issues and 'problem' for scanning
# If base, and no others > keep
# If base, but there's a scaled > remove base, keep scaled
# If problem, should be kept, as should already be unique

leaf_area_clean <- leaf_area_raw %>%
  rename(envelope_ID = ID) %>%
  mutate(file_type = case_when(
    str_detect(source_file, "ScaledDifferently_ProblemImages") ~ "problem",
    str_detect(source_file, "ScaledDifferently") ~ "scaled",
    TRUE ~ "base")) %>%
  group_by(envelope_ID) %>%
  # Create flag columns to identify decision making
  mutate(has_scaled = any(file_type == "scaled"),
         dropped_base_because_scaled = (file_type == "base" & has_scaled))

# Remove the ones with issues
leaf_area_clean <- leaf_area_clean %>%
  filter(!dropped_base_because_scaled | file_type == "problem") %>%
  mutate(leaf_area_scanned_bulk_nr = as.numeric(n),
         leaf_area_sum = as.numeric(leaf_area)) %>%
  select(-c(n, leaf_area, file_type, has_scaled, dropped_base_because_scaled)) %>% 
  ungroup()



# Join the clean leaf area db to the main trait db by the envelope ID
data <- raw_data %>%
  dplyr::mutate(envelope_ID = stringr::str_trim(as.character(envelope_ID))) %>% 
  # Left join, keep values from x, and only matching rows in y (leaf_area_clean)
  dplyr::left_join(leaf_area_clean, by = "envelope_ID")

#To check for duplicate rows in data

any(duplicated(data$envelope_ID))

data$envelope_ID[duplicated(data$envelope_ID)]

#To check for duplicate rows in leaf_area_clean

any(duplicated(leaf_area_clean$envelope_ID))

leaf_area_clean$envelope_ID[duplicated(leaf_area_clean$envelope_ID)]

#To check for duplicate rows in leaf_area_raw -no duplicates here!

any(duplicated(leaf_area_raw$envelope_ID))

#To check for duplicate rows in raw_data -no duplicates here!

any(duplicated(raw_data$envelope_ID))

#remove duplicates AFB1674 and AFZ7040 as they are identical


data <- data %>% distinct(envelope_ID, .keep_all = TRUE)

#Check that the duplicates are removed

data$envelope_ID[duplicated(data$envelope_ID)]

```

# Add in Plant Height and CollTime

```{r}
#| label: create height vegetative column

library(lubridate)
library(readxl)
library(dplyr)
library(hms)

#import metadata
metadata <- read_excel("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/raw_data/DURIN-Hydraulics-Calluna-Sampling-December-2025.xlsx") |>
mutate(age_class = case_when(
age_class == "Pioneer" ~ "Pioneer",
age_class == "Building_Mature" ~ "Building",
age_class == "Mature_Degenerative" ~ "Mature"
))

#create a new column with unique id's to match with the functional traits dataset
metadata <- metadata |>
mutate(unique_plantID = paste(age_class,"_",NiN,"_",plant_nr, sep = ""),
       collection_time = as_hms(time)
)



#updating the colltime column by matching with the sampling dataset
data$collection_time <- metadata$collection_time[match(data$unique_plantID, metadata$unique_plantID)]


#creates a new column with plant height using the height from the metadata dataset

data$height_cm <- metadata$height_cm[match(data$unique_plantID, metadata$unique_plantID)]




```

# Data Cleaning Entire Dataframe

Above was cleaning of column names, formatting of variables etc. Now we need to identify any mistakes in the data entry, outliers etc.

```{r}
#| label: standardize comments and remarks

# Find unique comments in the dataset
unique(data$remark) # currently for SG there are 37 unique comments.

# Or if you'd like to save a dataframe of the remarks
data_remarks <- data %>% 
  filter(!is.na(remark)) # Choose only those rows where there is not a NA in the remarks column, NOTE, this is not unqiue remarks here.

# Remember, sometimes multiple comment strings in the column, and can be seperated by ";" (EXAMPLE ONLY AT THE MOMENT!)
# cleaning up remarks to match envelopes + standardizing 
# standardizing incorrect texts
    # e.g. if remark column says "only 2 leafs at dry mass" change to "only 2 leaves at dry mass".



data <- data %>%  
   mutate(remark = tolower(remark))%>% #evrything is in small letter to remove all capital letter mistakes

  mutate(
    remark = case_when(
      remark == "Plant has not very many good leaves to choose from ; Leaf thickness by ES" ~ ";Plant has not very many good leaves to choose from; Leaf thickness by ES",
      
      remark == "longer shoots mostly only available" ~ "longer shoots only available",
      
      remark == "4 leaves at scan and thickness" ~ "4 leaves at scanning",
      
      remark == "only 4 leaves on the scanning stage" ~ "4 leaves at scanning",
      
      remark == "only 4 leaves at scanning" ~ "4 leaves at scanning",
      
      remark == "leaf went missing at wet mass -now 4" ~ "4 leaves at scanning",
      
      remark == "only 4 leaves at scanning and weighing" ~ "4 leaves at scanning",
      
      remark == ";plant has not very many good leaves to choose from; leaf thickness by es" ~ "plant has not very many good leaves to choose from; leaf thickness by es",
      
      remark == "only 4 leaves at data entry" ~ "4 leaves at data entry",
      
      remark == "leaves are sampled from old short shoots" ~ "leaves sampled from new on old short shoot",
      
       remark == "leaves are sampled from new on old short shoot" ~ "leaves sampled from new on old short shoot",
      
        remark == "only 4 leaves at dry weight; only 4 leaves at data entry" ~ "4 leaves at dry weight; 4 leaves at data entry",
      
       remark == "only 4 leaves at scanning; only 4 leaves at data entry" ~ "4 leaves at scanning; 4 leaves at data entry",
      
      remark == "small leaves; only 4 leaves at scanning and weighing; only 4 leaves at data entry" ~ "small leaves; 4 leaves at scanning; 4 leaves at data entry",
      
         remark == "5 leaves at data entry" ~ "4 leaves at data cleaning",
      
      TRUE ~ remark,
      ))
      
      # this means if the remark doesn't have any other issues flagged above, it just returns the original remark in the cell


# Check comments after the fixes above
unique(data$remark) 

#32 comments ok?
```

# Correcting outliers

```{r}
#|label: Correcting outliers

#Mean dry_mass for AGV1765 is almost 10x higher than other values, has 6dp in datasheet and 4dp on envelope, decided it's a typo, decision is to change the value by adding in an extra 0. Column: dry_mass and 0.057300g

data <- data %>%
  mutate(
    
    # adjust to be which ever trait has the issue
    dry_mass = if_else(
      
      # list the specific envelope ID with the issue and then the specific column value of interest that was specified above 
      envelope_ID == "AGV1765" & dry_mass == 0.057300,
      
      # write the replacement value
      0.00573,
      
      dry_mass))

# Add remark

# Use mutate() and str_c() to add comments to a string
data <- data %>%
  
  # in the 'remark' or the 'comment' column, list the envelopeID from above
  mutate(remark = ifelse(envelope_ID %in% c("AGV1765"),
                         
                         # Add to the existing 'remark' column, a comment but ensure it starts with '; ' first so it's not running in with the prior column, but adds to the end. Here hte comment is 'dry mass error updated'
                         # "; dry mass error, updated"
                         
                         str_c(coalesce(remark, ""), "; dry mass error, updated"), remark)) 

########################################################

#AMP6104 has 10x lower dry mass, comment says "6dp at dry mass", so decision to remove extra 0 behind the comma, dry_mass and 0.001019g


data <- data %>%
  mutate(
    
    dry_mass = if_else(
      
      envelope_ID == "AMP6104" & dry_mass == 0.001019,
      0.01019,
      
      dry_mass))

data <- data %>%
  
  mutate(remark = ifelse(envelope_ID %in% c("AMP6104"),
                         
                         str_c(coalesce(remark, ""), "; dry mass error, updated"), remark)) 


########################################################
#ACI2672 has 3 leaves as bulk nr written in data file and on actual envelope, but on the leaf area scans there are 5 leaves, later it is only found 4 leaves at data cleaning, bulk nr and 3

data <- data %>%
  mutate(
    
    bulk_nr_leaves = if_else(
      
      envelope_ID == "ACI2672" & bulk_nr_leaves == 3,
      5,
      
      bulk_nr_leaves))

data <- data %>%
  
  mutate(remark = ifelse(envelope_ID %in% c("ACI2672"),
                         
                         str_c(coalesce(remark, ""), "; bulk nr error, updated"), remark)) 

######################################################
#AGR3236 has a leaf thickness of 0.780mm written in data sheet, but on envelope it is actually 0.180mm, leaf thickness and 0.780mm


data <- data %>%
  mutate(
    
    leaf_thickness_2 = if_else(
      
      envelope_ID == "AGR3236" & leaf_thickness_2 == 0.780,
      0.180,
      
      leaf_thickness_2))

data <- data %>%
  
  mutate(remark = ifelse(envelope_ID %in% c("AGR3236"),
                         
                         str_c(coalesce(remark, ""), "; leaf thickness error, updated"), remark)) 

######################################################
#AKC8290 has a leaf thickness 2 value that is an outlier, 0.956mm with a z-score of 6.82 and will be changed to NA


data <- data %>%
  mutate(
    leaf_thickness_2 = if_else(
      envelope_ID == "AKC8290" & leaf_thickness_2 == 0.956,
      NA,
      leaf_thickness_2))

data <- data %>%
  mutate(remark = ifelse(envelope_ID %in% c("AKC8290"),
                         str_c(coalesce(remark, ""), "; leaf thickness outlier, omitted"), remark)) 
```

# Calculate per leaf trait values

While the bulk number from the envelope represents the initial sample size, sometimes leaves are lost or gained during the trait sampling process. Need to check comments and the particle count from the leaf area scans to adjust the bulk number leaves for each trait. This needs to be done before calculating per leaf trait values, which can be used for metrics like SLA and LDMC.

Relies on cleaned remarks column from step above

Order of trait measurements is: wet_mass, leaf_area, leaf_thickness, dry_mass - stem diameters don't rely on bulk values.

We then also want to know the remarks for each given trait ideally, so we make seperate columns for these based off the intial remarks column.

```{r}
#| label: calculate bulk nr for each trait

# Before checking for outliers, need to convert traits to 'per leaf' values, else outliers might simply reflect a shift in the leaf numbers for 'absolute' values.

# Note, this is only for wet mass, leaf area, dry mass.
# For thickness, each measurement (1-3) needs to be checked first, then averaged in case one of the replicate samples is an outlier and will skew the average thickness


# Default column is the initial bulk_nr, initial values for following trait derived from prior trait + edits via comments
data$bulk_nr_wet_mass <- data$bulk_nr_leaves # Create a new column 'bulk_nr_wet_mass' that's derived from the overall 'bulk_nr' column

#So far, no comments about wet mass stage

# Modify based on the wet mass comments
#data <- data %>%
#  mutate(bulk_nr_wet_mass = case_when(
    
    
#   grepl("3 leaves at wet mass", remark) ~ 3, #TODO: Update to match EIRIK's sampling strategy and the comments from above
    # Add more statements like above if there are more relevant comments
#    .default = bulk_nr_wet_mass    # Keep original bulk nr value if no comment is applied to row
#  ))



# Create leaf area, rename the particle counts from scanner. More accurate then comments for values less than 4. However, sometimes leaves break into multiple parts, so the particle count may be higher than the leaf count artificially.
data <- data %>% 
  rename(bulk_nr_leaf_area = leaf_area_scanned_bulk_nr)

# Modify based on the wet mass comments
data <- data %>%
  mutate(bulk_nr_leaf_area = case_when(
    # Must be cross-checked against images!! DO NOT UNIVERSALLY APPLY
    grepl("4 leaves at scanning", remark) ~ 4, # if "statement" in the remark column, then change to value outside brackets in bulk_nr_leaf_area
    grepl("4 leaves at scanning; 4 leaves at data entry", remark) ~ 4,
    grepl("small leaves; 4 leaves at scanning; 4 leaves at data entry", remark) ~ 4,
    .default = bulk_nr_leaf_area    # Keep original bulk nr value if no comment is applied to row, and is confirmed by images.
  ))

# If needed, can check count of leaf numbers by the remark column to help identify issues
table(data$remark, data$bulk_nr_leaf_area)

# Find rows where leaf number for leaf area is higher than leaf number for wet mass
# Need to go and check 1) leaf scans, 2) envelopes to see if extra particles are in envelope, 3) remarks column, 4) the excel tab on leaf area checks
bulk_nr_issues <- data %>%
  filter(bulk_nr_leaf_area > bulk_nr_wet_mass) %>% # only choose values where bulk number is higher for area than wet mass
  select(envelope_ID, remark, bulk_nr_leaves, bulk_nr_wet_mass, bulk_nr_leaf_area) # only return these columns back instead of all columns


# Address each outcome by envelope_ID - write the notes here about confirmations/ decisions
# E.g.
# AEI1422 actually has 5 leaves- see comment edit above
# ADM0437 actually has 5 leaves - see comment edit above
# All other envelopes have correct number based on the wet mass column in the images, so correct back to the wet mass values


# Modify based on the comments, checking images etc. 
data <- data %>%
  mutate(bulk_nr_leaf_area = case_when( # modify bulk_nr_leaf_area when the following occurs...
    # Envelope ID's are X, then do Y
    #TODO: THESE ARE EXAMPLES
    grepl("AEI1422", envelope_ID) ~ 5,
    grepl("ADM0437", envelope_ID) ~ 5,    
    grepl("5 leaves at scanning", remark) ~ 5,
    grepl("3 leaves at scanning", remark) ~ 3,
    .default = bulk_nr_wet_mass    # Keep original bulk nr value if no comment is applied to row
  ))




# Note, for thickness the value is not correct by bulk leaf number, however, if there are fewer than three leaves, there may only be one or two leaf thickness measurements to average across. 
data$bulk_nr_thickness <- data$bulk_nr_leaf_area

# Check how many envelopes have fewer than three thickness measurements
bulk_nr_issues_thickness <-  data %>%
  filter(rowSums(across(starts_with("leaf_thickness_"), ~ . > 0)) < 3)

# Are there any issues? If so, list here in commented statements with EnvelopeID's

# Envelope ALL9208 only have 2 thickness measurements, but didn't appear when checking thickness issues in code, is the code adding up the 3 columns of leaf_thickness 1,2 and 3?

#No thickness comments

# Modify based on the thickness comments
data <- data %>%
  mutate(bulk_nr_thickness = case_when(
        #TODO: THESE ARE EXAMPLES
    grepl("3 leaves at thickness", remark) ~ 3,
    grepl("5 leaves at thickness", remark) ~ 5,
    .default = bulk_nr_thickness    # Keep original bulk nr value if no comment is applied to row
  ))


# Create dry mass bulk nr based on thickness
data$bulk_nr_dry_mass <- data$bulk_nr_thickness

# Modify based on the dry mass comments
data <- data %>%
  mutate(bulk_nr_dry_mass = case_when(
        #TODO: THESE ARE EXAMPLES
    grepl("4 leaves at dry weight; 4 leaves at data entry", remark) ~ 4,
    .default = bulk_nr_dry_mass    # Keep original bulk nr value if no comment is applied to row
  ))

```

```{r}
#| label: calculate per leaf values for each trait


data <- data %>% 
  mutate(
    wet_mass = wet_mass / bulk_nr_wet_mass,
    leaf_area = leaf_area_sum / bulk_nr_leaf_area,
    dry_mass = dry_mass / bulk_nr_dry_mass
  ) %>% 
  # Round the decimal places
  mutate(
    wet_mass = round(wet_mass, digits = 5),
    leaf_area = round(leaf_area, digits = 3),
    dry_mass = round(dry_mass, digits =  5)
  )
```

```{r}
#| label: check plant height (vegetative) outliers

# making a scatterplot for vegetative plant height per plant
(data %>% 
       ggplot(aes(x = factor(plant_nr), y = height_cm, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Plant Height - Vegetative (cm)",
    title = "Mean Vegetative Plant Height (cm) by plant, age class and NiN type",
        color = "Plant Number"
  ))


```

```{r}
#| label: check plant height (vegetative) outliers

# making a scatterplot for vegetative plant height per plant
(data %>% 
       ggplot(aes(x = factor(plant_nr), y = height_cm, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Plant Height - Vegetative (cm)",
    title = "Mean Vegetative Plant Height (cm) by plant, age class and NiN type",
        color = "Plant Number"
  ))



```

```{r}
#| label: check wet mass outliers (per leaf)

# Scatterplot for wet mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis
# Check for unit issues and decimal places as well

(data %>% 
       ggplot(aes(x = factor(plant_nr), y = wet_mass, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Wet mass (g)",
    title = "Mean wet mass per leaf by plant, age class and NiN type",
    color = "Plant Number"
  ))


```

```{r}
#| label: check leaf area outliers (per leaf)

# Scatterplot for leaf area mean for trait leaves, plant_nr on x-axis and mean leaf area on y-axis. by plant nr, age class, NiN type 
(data %>% ggplot(aes(x = factor(plant_nr), y = leaf_area, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and NiN type",
    color = "Plant Number"
  ))

# Check source file issues
(data %>% ggplot(aes(x = factor(plant_nr), y = leaf_area, colour = factor(source_file))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and NiN type",
    color = "Source File"
  ))

# Check source file issues
(data %>% ggplot(aes(x = factor(plant_nr), y = leaf_area, colour = factor(source_file))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ source_file) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and Source File",
    color = "Source File"
  ))
```

```{r}
#| label: check leaf thickness outliers (per measurement 1-3 replicates)
# Scatterplot for leaf thickness distribution for trait leaves, not mean, rather separate reps


data_long_thickness <- data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

ggplot(data_long_thickness, #plotting drom long format
       aes(x = factor(plant_nr), y = leaf_thickness, colour = factor(leaf_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf thickness (mm)",
    title = "Leaf thickness replicates by plant, age class and NiN type"
  )

#Check if BUI FH 3, leaf thickness 2, value of 0.956mm is more than 3 SDs from mean

#Creating a unique identifier
data$plant_id <- interaction(
  data$calluna_NiN_type,
  data$age_class,
  data$plant_nr
)


data$leaf_thickness_mean <- rowMeans(data[c("leaf_thickness_1", "leaf_thickness_2", "leaf_thickness_3")], na.rm=TRUE)

mean_thickness_per_plant <- data %>%
  group_by(plant_id) %>%
  summarise(
    plant_mean = mean(leaf_thickness_mean, na.rm = TRUE),
    plant_sd   = sd(leaf_thickness_mean, na.rm = TRUE),
    n = n()
  )

#Z score formula

z <-(0.956-0.4725333)/0.07085337
z

#As z-score is 6.82, the leaf thickness 2 for envelope AKC8290 will be omitted from the cleaned script
```

```{r}
#| label: check dry mass outliers (per leaf)

# Scatterplot for dry mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis
# Check for unit issues and decimal places as well

(data %>% ggplot(aes(x = factor(plant_nr), y = dry_mass, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Dry mass (g)",
    title = "Mean dry mass per leaf by plant, age class and NiN type"
  ))

```

```{r}
#| label: wet mass vs dry mass
# Dry mass should never be more than wet mass
# Values above the dotted 1:1 indicate an issue
# Also check for values that might have unit issues, i.e. 0.001 instead of 0.0001 etc
(data %>% 
   ggplot(aes(x = wet_mass, y = dry_mass)) +
  
  # scatter points
  geom_point(aes(colour = factor(plant_nr)),
             size = 2, alpha = 0.7) +
  
  # 1:1 reference line
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50") +
  
  # fitted regression line
  geom_smooth(method = "lm", se = TRUE, colour = "blue") +
  
  # optional: label
  labs(
    x = "Wet mass per leaf (g)",
    y = "Dry mass per leaf (g)",
    title = "Dry mass vs wet mass per leaf",
    subtitle = "Dashed line = 1:1 reference, blue line = fitted linear trend",
    colour = "Plant Nr"
  ) +
  
  theme_bw())

```

# Calculate SLA and LDMC

```{r}
#| label: calculate SLA and LDMC

# Specific Leaf Area (SLA)
data <- data %>%  
  mutate(
    SLA = ifelse(
      !is.na(leaf_area) & !is.na(dry_mass), # excluding rows with NAs
      leaf_area / dry_mass, # Leaf level metrics
      NA_real_
    )
  )


# Leaf Dry Matter Content (LDMC)
data <- data %>% 
  mutate(
    LDMC = ifelse( # making new ldmc column
      !is.na(dry_mass) & !is.na(wet_mass), # excluding rows with NAs
      dry_mass / wet_mass, # Leaf level metrics
      NA_real_
    )
  )
```

```{r}
#| label: check SLA outliers (per leaf)

# Scatterplot for SLA for each plant_nr on x-axis and weight in g on y-axis
(data %>% 
    ggplot(aes(x = factor(plant_nr), y = SLA, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "SLA",
    title = "Mean SLA per leaf by plant, age class and NiN type",
    color = "Plant Nr"
  ))
```

```{r}
#| label: check LDMC outliers (per leaf)

# Scatterplot for dry mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis

(data %>% 
       ggplot(aes(x = factor(plant_nr), y = LDMC, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ calluna_NIN_type) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf Dry Matter Content (LDMC)",
    title = "Mean LDMC per leaf by plant, age class and NiN type",
    color = "Plant Nr"
  ))
```

## Checking leaf thickness by measurer

```{r}
#| label: Check leaf thickness done by SG vs ES

# Check leaf thickness bias

data_thickness_by <- data %>%
  mutate(measurer = case_when(
    str_detect(remark, "es") ~ "es",
    TRUE ~ "SG"
  ))


# Make sure there are only one envelope ID in the data_long_thickness for the plots underneath

leaf_thickness_by_id <- data_long_thickness %>%
  dplyr::mutate(envelope_ID = stringr::str_trim(as.character(envelope_ID))) %>%
  dplyr::group_by(envelope_ID) %>%
  dplyr::summarise(leaf_thickness = mean(leaf_thickness, na.rm = TRUE), .groups = "drop")

data_thickness_by <- data_thickness_by %>%
  dplyr::left_join(leaf_thickness_by_id, by = "envelope_ID")

#Plot with all thicknesses

 data_thickness_by %>% ggplot(aes(
     x = factor(plant_nr),
     y = leaf_thickness_1,leaf_thickness_2, leaf_thickness_3,
     colour = measurer
   )) +
   geom_point(
     size = 2,
     alpha = 0.8,
     position = position_jitter(width = 0.15)) +
   facet_grid(age_class ~ calluna_NIN_type) +
   theme_bw() +
   labs(
     x = "Plant number",
     y = "Leaf thickness (mm)",
     title = "Leaf thickness by measurer",
     colour = "Measured by"
   )

# Check leaf thickness bias- plot with mean thickness for each sample


#Add column leaf_thickness to data_thickness_by -only oif using mean thickness values, but this plot looked very messy so decided to leave it out

  
#data_thickness_by <- data_thickness_by %>%
#  dplyr::left_join(
#    data_long_thickness %>%
#      dplyr::mutate(envelope_ID = stringr::str_trim(as.character(envelope_ID))) %>%
#      dplyr::select(envelope_ID,leaf_thickness),
#    by = "envelope_ID"
#  )
#
#data_thickness_by %>% ggplot(aes(
#     x = factor(plant_nr),
#     y = leaf_thickness,
#     colour = measurer
#   )) +
#   geom_point(
#     size = 2,
#     alpha = 0.8,
#     position = position_jitter(width = 0.15)) +
#   facet_grid(age_class ~ calluna_NIN_type) +
#   theme_bw() +
#   labs(
#     x = "Plant number",
#     y = "Leaf thickness (mm)",
#     title = "Leaf thickness by measurer",
#    colour = "Measured by"
#   )



```

# Fix the Outliers

```{r}
#| label: check leaf area scans and outliers 

################### Not corrected ######################

#Missing leaf area scans for: ADR7542, ACH0524,AIO5773 





##################### Corrected #######################
#Check mean dry mass for AGV1765, the value is almost 10x the other leaves of the plant
#AMP6104 has 10x lower dry mass, comment says "6dp at dry mass", so might be an accidental extra 0 behind the comma
#Check leaf area scans bulk nr for: ACI2672 (has 3 leaves written in data file, but i suspect it's supposed to be 4 or 5)
#Check leaf thickness_2 for AKC8290, it is quite a high value compared to the other values of the bulk

#Check leaf thickness_2 for AGR3236, quite high value

```

```{r}
#| label: Saving cleaned output


write.csv(data, "/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/clean_data/cleaned_data.csv", row.names = FALSE)

```
