---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format: html
editor: visual
---

# Importing data

Have to see which packages are needed later and add them.

```{r}
#| label: load libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(readr) # Data importation functions
library(stringr) # String and character text
library(dplyr)
library(ggplot2)
library(rstatix)
library(car)

```

```{r}
#| label: importing raw data of traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/clean_data/cleaned_data.csv")


#Creating a data frame
cleaned_data <- as.data.frame(cleaned_data)
cleaned_data$calluna_NIN_type <- factor(cleaned_data$calluna_NIN_type)
cleaned_data$age_class <- factor(cleaned_data$age_class)

#Make age class in order pioneer, building, mature


```

```{r}
#| label: Data visualization and overview

summary(cleaned_data)
str(cleaned_data)

# Order relevant variables (useful for plotting and order based analyses)
# age_class sorted from youngest to oldest

cleaned_data$age_class <- factor((cleaned_data$age_class),
                             levels = c("Pioneer", 
                                        "Building", "Mature"))
# Check if changed successfully
levels(cleaned_data$age_class)


# Order NIN types
cleaned_data$calluna_NIN_type <- factor((cleaned_data$calluna_NiN_type),
                                    levels = c("BH", "FH", "TH"))
# Check if changed successfully
levels(cleaned_data$calluna_NiN_type)

```

Remember to exclude NA values if they confuse the code

# Wet mass(g) visualizations

```{r}
#| label: Wet mass visualizations

# Data frame with mean, median and se

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE), 
            median_wet_mass= median.default(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()), 
            n = n(), .groups = "drop")

#Histograms

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lightblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")



#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Wet mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()+ 
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)


#Interactive plots


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  theme_minimal()+ 
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)



#Single Density Plot --not working yet

ggplot(cleaned_data, aes(x = wet_mass, y=age_class~calluna_NiN_type) +
  geom_density(fill = "skyblue", color = "midnightblue", alpha = 0.6) +
  labs(title = "Density plot wet mass", x = "Wet mass", y = "Density") +
  theme_minimal()





```

# Dry mass(g) visualizations

```{r}
#| label: Dry mass visualization

# Data frame with mean and se

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lavender") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")

#testing for normal distribution with Q-Q plot

qqnorm(cleaned_data$dry_mass)
qqline(cleaned_data$dry_mass)

#testing for normality with Shapiro-Wilk test ---gives a result of non-normal data

shapiro.test(cleaned_data$dry_mass)

#Linear model

lm_dry_mass <- lm(dry_mass ~ calluna_NIN_type * age_class, data = cleaned_data)
summary(lm_dry_mass)

#Checking residuals

qqnorm(residuals(lm_dry_mass))
qqline(residuals(lm_dry_mass))

shapiro.test(residuals(lm_dry_mass))


#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Dry mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()

#Interactive plot

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line(linewidth=1) + geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_dry_mass, aes(x = calluna_NIN_type, y = mean_dry_mass, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme_minimal() + labs(
    x = "Calluna NIN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)



#Plotting fitted values, needs to be changed!!!

newd_dry_mass <- expand.grid(calluna_NIN_type = levels(cleaned_data$calluna_NIN_type),
                    age_class = levels(cleaned_data$age_class))
newd_dry_mass$pred_dry_mass <- predict(lm_dry_mass, newd_dry_mass, se.fit = TRUE)$fit

ggplot(newd_dry_mass, aes(x = age_class, y = pred_dry_mass, color = calluna_NIN_type, group = calluna_NIN_type)) +
   geom_line() + geom_point() + geom_line(linewidth = 1) +
geom_point(size = 3.5)+ theme_minimal()

```

# Height (cm)

```{r}
#| label: Height visualizations

# Data frame with mean and se

df_sum_height <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(bins = 30, color = "black", fill = "violet") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Height (cm)")


# violin plot

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()

# Interactive plot


ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
   theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean height (cm)",
    color = "Age class"
)

# swapped axis, NiN type on x-axis

ggplot(df_sum_height, aes(x = calluna_NIN_type, y = mean_height_cm, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
    theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean height (cm)",
    color = "Age class"
)
  



```

# Leaf thickness (mm) visualizations

```{r}
#| label: Leaf thickness visualizations

#create table with leaf thickness 1,2,3 combined
data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

# Data frame with mean and se

df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram



ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(bins = 30, color = "black", fill = "forestgreen") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Leaf thickness (mm)", y = "Number of Calluna individuals")





# Boxplot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Leaf thickness (mm)")

# violin plot

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


# Interactive plot


ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
   theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_thickness, aes(x = calluna_NIN_type, y = mean_leaf_thickness, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
     theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)




```

# Leaf area (mm^2^) visualizations

```{r}
#| label: Leaf area visualizations

# Data frame with mean and se

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(bins = 30, color = "black", fill = "pink") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Leaf area (mm^2))", y = "Number of Calluna individuals")



# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Leaf area (mm)")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()



# Interactive plot


ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  )

# swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_area, aes(x = calluna_NIN_type, y = mean_leaf_area, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "Mean leaf area (mm^2)",
    color = "Age class"
  )



```

# LDMC visualizations

```{r}
#| label: LDMC visualizations

# Data frame with mean and se

df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")


# Histogram

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(bins = 30, color = "black", fill = "orange") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "LDMC", y = "Number of Calluna individuals")


# Boxplot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "LDMC")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()



# Interactive plot


ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
     theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "LDMC",
    color = "Age class"
  )


#swapped axis, NiN type on x-axis

ggplot(df_sum_LDMC, aes(x = calluna_NIN_type, y = mean_LDMC, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
    theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NIN type",
    y = "LDMC",
    color = "Age class"
  )




```

# SLA visualizations

```{r}
#| label: SLA visualizations

# Data frame with mean and se


df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")




# Boxplot


ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "SLA")

# violin plot

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


# Interactive plot


ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Age class",
    y = "SLA",
    color = "Calluna NiN type"
  )

#swapped axis, NiN type on x-axis

ggplot(df_sum_SLA, aes(x = calluna_NIN_type, y = mean_SLA, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + labs(
    x = "Calluna NiN type",
    y = "SLA",
    color = "Age class"
  )






```

# 

# Testing bootstrap modelling

Does not work at the moment, have to figure out how to write this code and if it is needed

```{r}
#| label: bootstrap modelling

library(boot)

boot_dry_mass <- function(data, indices) {
  d <- cleaned_data[indices, ]
  
  lm_dry_mass
 
  coef(lm_dry_mass)

}

set.seed(123)

boot_result_dry_mass <- boot(
  data = df,
  statistic = boot_dry_mass,
  R = 1000
)


bt <- boot_result_dry_mass$t

apply(bt, 2, quantile, probs = c(0.025, 0.975))


boot_df <- data.frame(
  mean = boot_result_dry_mass$t0,
  lower = apply(boot_result_dry_mass$t, 2, quantile, 0.025),
  upper = apply(boot_result_dry_mass$t, 2, quantile, 0.975)
)



ggplot(boot_df, aes(x = age_class, y = mean, color= calluna_NIN_type)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper))



hist(boot_result_dry_mass$t[,1],
     main = "Bootstrap distribution",
     xlab = "Mean dry mass")

ggplot(boot_df, aes(x = group)) +
  geom_histogram(bins = 10,
                 fill = "lavender",
                 color = "black")
```

# Testing for normality before ANOVA

Attempting to make a qqplot with both age class and NiN + trait

```{r}
#| label: testing age class and NiN type for normality with shapiro-wilk, qq plots, and residuals

#Find if data meets assumptions for a two-way ANOVA and if transformations are needed by using shapiro-wilk test and qq plot



########################################### Wet mass
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(wet_mass)

#sort by p-values above 0.05, which indicates normality--------------> does not work yet, thought it would be eaier to filter out which relationships can be used for the ANOVA
shapiro_test(wet_mass)%>%
  filter(P_value > 0.05)




#QQ plot, separate for age class and NiN type: The QQ plot is generally more informative than the Shapiro test because large datasets often make Shapiro significant even when deviations are minor.


ggplot(cleaned_data, aes(sample = wet_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_wrap(~ age_class) +
  theme_minimal()

ggplot(cleaned_data, aes(sample = wet_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_wrap(~ calluna_NiN_type) +
  theme_minimal()

#Combined plot for age class*NiN type, for every combination type, better overview

ggplot(cleaned_data, aes(sample = wet_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()



#Checking residuals 
#W = 0.98438, p-value = 0.001733, indicates non-normality


lm_wet_mass <- lm(wet_mass ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_wet_mass)

qqnorm(residuals(lm_wet_mass))
qqline(residuals(lm_wet_mass), col = "red")

shapiro.test(residuals(lm_wet_mass))





######################################### Dry mass
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(dry_mass)

#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = dry_mass)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#Checking residuals
# W = 0.97974, p-value = 0.0001981 indicates data is not normally distributed, even if it appears so

lm_dry_mass <- lm(dry_mass ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_wet_mass)

qqnorm(residuals(lm_dry_mass))
qqline(residuals(lm_dry_mass), col = "red")

shapiro.test(residuals(lm_dry_mass))




##########################################Height
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(height_cm)



#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = height_cm)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()


#Checking residuals
#W = 0.98229, p-value = 0.0006273, indicates non-normality

lm_height_cm <- lm(height_cm ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_height_cm)

qqnorm(residuals(lm_height_cm))
qqline(residuals(lm_height_cm), col = "red")

shapiro.test(residuals(lm_height_cm))





###########################################Leaf area
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_area)



#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = leaf_area)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#Checking residuals
#W = 0.98409, p-value = 0.001577, indicates non-normality

lm_leaf_area <- lm(leaf_area ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_leaf_area)

qqnorm(residuals(lm_leaf_area))
qqline(residuals(lm_leaf_area), col = "red")

shapiro.test(residuals(lm_leaf_area))






#####################################Leaf thickness
data_thickness %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_thickness)




#Combined plot for age class*NiN type, for every combination type

ggplot(data_thickness, aes(sample = leaf_thickness)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()


#checking residuals
#W = 0.96656, p-value = 6.664e-14, not normally distributed

lm_leaf_thickness <- lm(leaf_thickness ~ age_class*calluna_NIN_type, data = data_thickness)
summary(lm_leaf_thickness)

qqnorm(residuals(lm_leaf_thickness))
qqline(residuals(lm_leaf_thickness), col = "red")

shapiro.test(residuals(lm_leaf_thickness))






######################################SLA
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(SLA)




#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = SLA)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#checking residuals
#W = 0.96722, p-value = 1.639e-06, not normally distributed

lm_SLA <- lm(SLA ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_SLA)

qqnorm(residuals(lm_SLA))
qqline(residuals(lm_SLA), col = "red")

shapiro.test(residuals(lm_SLA))





####################################LDMC
cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(LDMC)




#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = LDMC)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  theme_minimal()

#Checking residuals
#W = 0.98146, p-value = 0.0004388, not normally distributed

lm_LDMC <- lm(LDMC ~ age_class*calluna_NIN_type, data = cleaned_data)
summary(lm_LDMC)

qqnorm(residuals(lm_LDMC))
qqline(residuals(lm_LDMC), col = "red")

shapiro.test(residuals(lm_LDMC))

```

# Levene test for checking variance

```{r}
#| label: levene test to check for homogeneity of variance as ANOVA requires roughly equal variance across all groups

#wet mass


leveneTest(
  wet_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)
```

# Account for random effects and grouping by PlantID and bulk samples

Need to figure out how

```{r}
#| label: Accountig for grouping effects
```
