---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format:
  html:
    embed-resources: true
editor: visual
---

# Importing data

```{r}
#| label: load-libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(here)
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(ggridges) #Moving density plots
library(rstatix)
library(car)
library(performance) #checking model normality, homogeneity, variance
library(emmeans) #Post hoc testing
library(afex) #linear mixed models
library(lme4) #linear mixed models
library(sdamr) #linear mixed models
library(lmerTest)
library(multcomp) #emmeans contrasts on plots
library(factoextra) #PCA biplot
library(corrplot)
library(vegan) #RDA analysis
library(ggdist) 
library(ggrepel) #RDA plot making
library(ggbeeswarm)
library(ggnewscale) #Visuals of PCA
library(ggsignif) #betadisper display significance
library(patchwork)
library(conflicted)
conflict_prefer_all("dplyr", quiet = TRUE)


theme_set(theme_bw())
theme_update(plot.title.position = "plot")



```

```{r}
#| label: importing-raw-data-of-traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv(here::here("data/clean_data/cleaned_data_copy.csv"))



# Order relevant variables 
# age_class sorted from youngest to oldest, NiN type sorted from wettest to driest
cleaned_data$calluna_NiN_type <- factor(cleaned_data$calluna_NiN_type,  levels = c("BH","FH","TH"))
cleaned_data$age_class <- factor(cleaned_data$age_class, levels = c("Pioneer", 
                                        "Building", "Mature"))


#Add column of mean leaf thickness to cleaned data
cleaned_data$leaf_thickness_mean <- rowMeans(cleaned_data[, c("leaf_thickness_1", "leaf_thickness_2", "leaf_thickness_3")], na.rm=TRUE)
```

```{r}
#| label: Data-visualization-and-overview

summary(cleaned_data)
glimpse(cleaned_data)

## Define colour palette (used with scale fill manual) 

NIN <- c("BH" = "darkslategrey", "FH" = "deepskyblue4", "TH" = "lightskyblue")
AGE <- c("Pioneer" = "#087E8B","Building" = "#EB5461", "Mature" = "#884000") 

```

# Wet mass(g) 

```{r}
#| label: Data-frame-with-mean-median-and-se-histogram-wet-mass

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE), 
            median_wet_mass= median.default(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()), 
            n = n(), .groups = "drop")

#Histograms, general and NiN and age class specific

ggplot(cleaned_data, aes(x = wet_mass)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1) 

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NiN_type ~ age_class) +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")


```

Boxplot to check overlapping error bars and data medians

```{r}
#| label: Boxplot-wet-mass

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = wet_mass, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Wet mass (g)")
```

```{r}
#| label: Interactive-plots-wet-mass


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)
```

```{r}
#| label: Density-Plot-wet-mass

ggplot(cleaned_data, aes(x = wet_mass, y =calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges (alpha = 0.5)+ #To separate the density plots vertically
    facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot wet mass", x = "Wet mass")



```

# Dry mass(g) 

@fig-Data-frame-histogram-dry-mass shows slight right skewedness of data\

```{r}
#| label: fig-Data-frame-histogram-dry-mass

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN and age class specific

ggplot(cleaned_data, aes(x = dry_mass)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1)


ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot-dry-mass

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = dry_mass, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Dry mass (g)")
```

```{r}
#| label: Interactive-plot-dry-mass

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Age class",
    y = "Mean dry mass (g)",
    color = "NiN type"
)

```

```{r}
#| label: Density-Plot-dry-mass

ggplot(cleaned_data, aes(x = dry_mass, y = calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot dry mass (g)", x = "Dry mass (g)") 
```

# Height (cm)

-@Data-frame-histogram-height does not appear normally distributed, histogram looks bimodal, but might be too few datapoints

```{r}
#| label: Data-frame-histogram-height

df_sum_height <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN and age class specific

ggplot(cleaned_data, aes(x = height_cm)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1)

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")
```

-From @Boxplot_height, the variances across age classes have quite a different scatter, MAT in BH and FH have a lot more variation than PIO i.e

-More variation in MAT, but TH shows a different trend than FH and BH again. The BUI for TH is acting differently that the other nature types

-Plants generally taller in BH and FH

```{r}
#| label: Boxplot-height

Boxplot_height <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = height_cm, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Vegetative height (cm)", title="Vegetative height (cm) of Calluna vulgaris across Age class and NiN type")

Boxplot_height

```

```{r}
#| label: Interaction-plot-height


ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Age class",
    y = "Mean height (cm)",
    color = "NiN type"
)


```

```{r}
#| label: Density-plot-height

ggplot(cleaned_data, aes(x = height_cm, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot height (cm)", x = "Height (cm)")


```

# Leaf thickness (mm) 

-@fig-dataframe-histogram-leaf-thickness appears normally distributed, a little right-skewed

```{r}
#| label: fig-dataframe-histogram-leaf-thickness

  
#create table with leaf thickness 1,2,3 combined
data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

# Data frame with mean and se
df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN, age class specific
ggplot(data_thickness, aes(x = leaf_thickness)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf thickness (mm)", y = "Number of Calluna individuals")
```

-leaf thickness quite overlapping in @boxplot_leaf_thickness, thickness highest for PIO in TH, lowest in FH (supports an initial conservative strategy), in BUI thickest leaves in BH, thinnest in TH (more acquisitive), for MAT back to thicker leaves in BH, TH and thinner in FH (conservative)

```{r}
#| label: Boxplot-leaf-thickness

boxplot_leaf_thickness <- ggplot(data_thickness, aes(x = calluna_NiN_type, y = leaf_thickness, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Leaf thickness (mm)")

boxplot_leaf_thickness
```

```{r}
#| label: Interaction-plot-leaf-thickness


interaction_plot_thickness <- ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness, width = 0.1)) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Age class",
    y = "Leaf thickness (mm)",
    color = "NiN type", title= "Mean leaf thickness (mm) of C. vulgaris against NiN type and age classes"
)

interaction_plot_thickness
```

```{r}
#| label: Density-Plot-leaf-thickness

ggplot(data_thickness, aes(x = leaf_thickness, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf thickness (mm)", x = "Leaf thickness (cm)") 


```

# Leaf area (cm^2^) visualizations

-@Data-frame-histogram-leaf-area appears right-skewed, but normally distributed

```{r}
#| label: Data-frame-histogram-leaf-area

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN, age class specific

ggplot(cleaned_data, aes(x = leaf_area)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf area (cm^2))", y = "Number of Calluna individuals")
```

-leaf area for TH vs FH in PIO and BUI might be significant, seen from @Boxplot-leaf-area. Generally greater leaf area from PIO→MAT (acquisitive), different trend in TH again. Greater variation in leaf area between age classes in FH

```{r}
#| label: Boxplot-leaf-area

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_area, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Leaf area (cm^2)")
```

```{r}
#| label: Interaction-plot-leaf-area


ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Age class",
    y = "Mean leaf area (cm^2)",
    color = "NiN type"
  ) 
```

```{r}
#| label: Density-Plot-leaf-area

ggplot(cleaned_data, aes(x = leaf_area, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf area (cm^2)", x = "Leaf area (cm^2)") 

```

# Leaf Dry Matter Content

-@fig-Data-frame-histogram-LDMC appears normally distributed, but left-skewed

```{r}
#| label: fig-Data-frame-histogram-LDMC


#Add mean for each age class -used in the results section
mean_LDMC_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(mean_LDMC_age = mean(LDMC, na.rm = TRUE),
            .groups = "drop")

#Add SE for age class
se_LDMC_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(se_LDMC_age = sd(LDMC, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")

#Add mean for each NiN type
mean_LDMC_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(mean_LDMC_NiN = mean(LDMC, na.rm = TRUE),
            .groups = "drop")

#Add SE for NiN type
se_LDMC_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(se_LDMC_NiN = sd(LDMC, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")



df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")  %>%
  left_join(mean_LDMC_age, by = "age_class") %>%
  left_join(mean_LDMC_NiN, by = "calluna_NiN_type") %>%
  left_join(se_LDMC_age, by = "age_class") %>%
  left_join(se_LDMC_NiN, by = "calluna_NiN_type")



# Histogram

ggplot(cleaned_data, aes(x = LDMC)) + 
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "LDMC", y = "Number of Calluna individuals")
```

\- @Boxplot-LDMC slightly increasing LDMC trend from Pio to Mat (plants become more conservative with age), FH shows a different trend with BUI having the highest LDMC. test significance PIO vs BUI in FH, and BUI: FH vs TH vs BH, FH as the highest LDMC and BH the lowest, TH in the middle (might be overlapping)

```{r}
#| label: Boxplot-LDMC

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")
```

```{r}
#| label: Interaction-plot-LDMC


ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "LDMC",
    color = "Age class"
  )
```

```{r}
#| label: Density-Plot-LDMC

ggplot(cleaned_data, aes(x = LDMC, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot LDMC", x = "LDMC") 




```

# Specific Leaf Area

\- @Data-frame-histogram-SLA appears right-skewed, Pio FH and Bui TH more scattered

```{r}
#| label: Data-frame-histogram-SLA

#Add mean for each age class
mean_SLA_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(mean_SLA_age = mean(SLA, na.rm = TRUE),
            .groups = "drop")

#Add se for age class
se_SLA_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(se_SLA_age = sd(SLA, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")

#Add mean for each NiN type
mean_SLA_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(mean_SLA_NiN = mean(SLA, na.rm = TRUE),
            .groups = "drop")

#Add se for NiN type
se_SLA_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(se_SLA_NiN = sd(SLA, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")

#Dataframe
df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

#Histogram

ggplot(cleaned_data, aes(x = SLA)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")
```

- Seen from @Boxplot-SLA, SLA generally highest for PIO, decreasing to MAT, except BUI in TH. SLA highest in FH for PIO, but quite similar in BH and TH for BUI and MAT. test significance in FH: Pio vs Bui and Mat, and for PIO FH vs TH vs BH

```{r}
#| label: Boxplot-SLA


ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "SLA")
```

```{r}
#| label: Interaction-plot-SLA


ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Age class",
    y = "SLA",
    color = "Calluna NiN type"
  )
```

```{r}
#| label: Density-plot-SLA

ggplot(cleaned_data, aes(x = SLA, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot SLA", x = "SLA") +
  theme_minimal()
```

# Testing for normality

-Shapiro-Wilk test reveals non-normality for the linear model, but since we have a large data set n\>50, the test is too sensitive and results might be a false negative, leading to rejection of normal data, when it is in reality fine to use.

-Both ANOVA and mixed models are robust to large datasets when it comes to deviating slightly from normality

```{r}
#| label: testing-normality-qq-plots

#Find if data meets assumptions for a two-way ANOVA and if transformations are needed by using qq plots
#Combined qqplot for age class*NiN type, for every combination type, better overview

#Wet mass

ggplot(cleaned_data, aes(sample = wet_mass)) +
 stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Wet mass qq plot")


#Dry mass

ggplot(cleaned_data, aes(sample = dry_mass)) +
stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Dry mass qq plot")

#Leaf area

ggplot(cleaned_data, aes(sample = leaf_area)) +
 stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf area qq plot")

#Leaf thickness

ggplot(data_thickness, aes(sample = leaf_thickness)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf thickness qq plot")

#SLA

ggplot(cleaned_data, aes(sample = SLA)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "SLA qq plot")

#LDMC

ggplot(cleaned_data, aes(sample = LDMC)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "LDMC qq plot")
```

-Plot for height is slightly heavy tailed and deviates slightly from the qqline, indicating data contains a few more extreme values than would be expected from a normal distribution. Scale plot also shows variance increases with fitted values

Height should be fitted to a linear model as we have the same values per plant which includes many of the samples

```{r}
#| label: Height-lm-and-testing-normality

ggplot(cleaned_data, aes(sample = height_cm)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Height qq plot")

#Linear model with interaction of age class and NiN type
lm_height_cm <- lm(height_cm ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_height_cm)

#Checking residuals
#W = 0.98229, p-value = 0.0006273, indicates non-normality
shapiro.test(residuals(lm_height_cm))

#qqplot is slightly heavy tailed, indicating data contains a few more extreme values than would be expected from a normal distribution
plot(lm_height_cm, which=2)

#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
plot(lm_height_cm, which = 3, add.smooth = TRUE)

check_model(lm_height_cm)

```

# Account for random effects and data clustering with lmer

→ Fixed effects with interactions: trait \~  NIN\*age class, effect of NiN type on trait also depends on effect of age class, and possibly plant sampling groups

-Plant ID (plant nr) is a grouping factor that should be used as random effects to account for non-independence, “within-individual” grouping. These are nested groups with lower level units contained within higher levels

→ Use mixed-effects model as factors with few levels are fixed (NiN, age class) and factors with many levels are random (Plant 1-7 grouping)

5 leaf bulks ⊂ 5 samples ⊂ Plant 1-7 ⊂ Age class × NiN type

\
-Linear mixed effects models, lme4 package, diff model, but same anova, summary etc. 1I plantID, not for height due to low replicate number

-With `lme4` for mixed-effects models, we check two types of normality:

- **Residual normality:** Just like ANOVA, the overall errors of the model should look fairly normal on a Q-Q plot.

- **Random effects normality:** The random intercepts or slopes you estimate for your groups (like subjects or schools) should also follow a normal distribution.

```{r}
#| label: Mixed-model-wet-mass

#data=cleaned_data
#dependent variable = plant trait
#fixed factors: NiN and age class
#Grouping factor: Plant_nr nested in age class


lme4_wet_mass <- lmerTest::lmer(wet_mass ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_wet_mass)
plot(lme4_wet_mass, which=2) #Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean

qq_wet_mass_lme4 <- qqnorm(residuals(lme4_wet_mass), main = "Q-Q Plot of lme Residuals of Wet mass")
qqline(residuals(lme4_wet_mass), col = "red")

check_model(lme4_wet_mass)
```

```{r}
#| label: Mixed-model-dry-mass

lme4_dry_mass <- lmerTest::lmer(dry_mass ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_dry_mass)
plot(lme4_dry_mass) #Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean

qq_dry_mass_lme4 <- qqnorm(residuals(lme4_dry_mass), main = "Q-Q Plot of Residuals from lme of Dry mass")
qqline(residuals(lme4_dry_mass), col = "red")

check_model(lme4_dry_mass)
```

```{r}
#| label: Mixed-model-leaf-area

lme4_leaf_area <- lmerTest::lmer(leaf_area ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data)

summary(lme4_leaf_area)  
plot(lme4_leaf_area)

check_model(lme4_leaf_area)
```

```{r}
#| label: Mixed-model-leaf-thickness

lme4_leaf_thickness <- lmerTest::lmer(leaf_thickness_mean ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_leaf_thickness) 
plot(lme4_leaf_thickness)

check_model(lme4_leaf_thickness)
```

```{r}
#| label: Mixed-model-LDMC

lme4_LDMC <- lmerTest::lmer(LDMC ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_LDMC)
plot(lme4_LDMC)

check_model(lme4_LDMC)
```

```{r}
#| label: Mixed-model-SLA

lme4_SLA <- lmerTest::lmer(SLA~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_SLA) 

plot(lme4_SLA) #Scale plot suggests mild heteroscedasticity with slight fanning effect increasing with mean

check_model(lme4_SLA)

```

# ANOVA and Emmeans for height

-H0: there are no significant differences between the interaction of age class and NiN type on given trait

-Post hoc test with more than two levels, emmeans package: Everything the **emmeans** package does is an interpretation of the model that is fitted to the data, the model has to be a good fit before proceeding with this step.

-t-ratio tells you how many Standard Errors (SE) your group ratio is away from the null hypothesis,

**Significant:** A t-ratio **greater than 2** or **less than -2** is generally statistically significant (assuming a standard 5% error rate).

**Not Significant:** A t-ratio close to 0 (between -2 and 2) means the groups are too similar to be sure of any real difference

estimate column: means the average difference in i.e for height in cm between groups

```{r}
#| label: ANOVA-emmeans-height

#Height

cat("######## Height (cm)\n\n")

anova(lm_height_cm)

(emmeans_height_cm<- emmeans(lm_height_cm, ~ age_class * calluna_NiN_type))
pairs(emmeans_height_cm, simple="each", adjust = "tukey") #Tukey adjustment to prevent false positives

#Grouping individual means by letters for visualization
cld(emmeans_height_cm, Letters=letters)
```

# ANOVA and emmeans for mixed models

```{r}
#| label: ANOVA-emmeans-wet-mass

#Wet mass

cat("######## Wet Mass (g)\n\n")

anova(lme4_wet_mass)

(emmeans_lme4_wet_mass<- emmeans(lme4_wet_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_wet_mass, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
```

```{r}
#| label: ANOVA-emmeans-dry-mass

cat("######## Dry Mass (g)\n\n")

anova(lme4_dry_mass)

(emmeans_lme4_dry_mass<- emmeans(lme4_dry_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_dry_mass, simple="each", adjust = "tukey") #Tukey adjustment to prevent false positives
```

```{r}
#| label: ANOVA-emmeans-leaf-area

cat("######## Leaf area (mm^2)\n\n")

anova(lme4_leaf_area)

(emmeans_lme4_leaf_area<- emmeans(lme4_leaf_area, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_leaf_area, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
```

```{r}
#| label: ANOVA-emmeans-leaf-thickness

cat("######## Leaf thickness (mm)\n\n")

anova(lme4_leaf_thickness)

(emmeans_lme4_leaf_thickness<- emmeans(lme4_leaf_thickness, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_leaf_thickness, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
```

```{r}
#| label: ANOVA-emmeans-LDMC

cat("######## LDMC\n\n")

anova(lme4_LDMC)

(emmeans_lme4_LDMC<- emmeans(lme4_LDMC, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_LDMC, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives

#Plots to visualize emmeans output
emmip(emmeans_lme4_LDMC, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_LDMC, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other
```

```{r}
#| label: ANOVA-emmeans-SLA

cat("######## SLA\n\n")

anova(lme4_SLA)

(emmeans_lme4_SLA<- emmeans(lme4_SLA, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_SLA, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
```

# Multivariate analysis and PCA

-Because traits are not independent, we should investigate interactions between them

-Use PCA to reduce dimensionality and identify major axes of trait covariation across NiN types and age classes (remember Diaz et al. 2016 method)

-Because leaf traits often covary (e.g., LES), check for high multicollinearity. Highly correlated variables ((r \> 0.85) or (0.90)) may skew PCA loadings toward specific traits –\> i.e remove wet mass, dry mass

```{r}
#| label: Correlations-between-traits

#Look at trait pairs scatterplot to evaluate if pearson is more appropriate than spearman, relationships are widely scattered, but more linear so pearson is better

traits_compiled <- cleaned_data[, c(
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC"
)]

traits_compiled <- na.omit(traits_compiled)
pairs(traits_compiled)

corr <- cor(cleaned_data[, c(
  "dry_mass",
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC"
)], method = "pearson",
use = "pairwise.complete.obs")

round(corr, 2)
summary(corr)

#Testing correlations between traits as SLA and LDMC are derived from variables already in the model ---appears only wet and dry mass are highly correlated --decision to omit dry mass from pca and rda as they are intermediate measurements use to calculate LDMC and SLA, which are common functional traits
#NAs are omitted, wet and dry mass excluded




```

**-data often scaled** to fairly compare across traits, we can statistically contrast the different traits side-by-side: makes skewed distributions more normal by pulling extreme outliers inward; allows interpretation in multiplicative steps (1x increase → 10x increase etc.); prevents heteroscedasticity (data artificially widening as mean increases); when comparing multiple traits, scaling makes sure a change of one unit in one trait means the same across all variables

```{r}
#| label: scaling trait values for glm model


cleaned_data <- cleaned_data %>%
  mutate(across(c(SLA, LDMC, leaf_area, leaf_thickness_mean, height_cm),
                ~ as.numeric(scale(.)),
                .names = "{.col}_scaled"))

#Test if scaling successful
#vector of column names

cols <- c("SLA_scaled", "LDMC_scaled", "leaf_area_scaled",
          "leaf_thickness_mean_scaled", "height_cm_scaled")

#should be close to 0, which it is here
cleaned_data %>%
  summarise(across(all_of(cols), mean, na.rm = TRUE))

#should be close to 1, which it is here
cleaned_data %>%
  summarise(across(all_of(cols), sd, na.rm = TRUE))

```

```{r}
#| label: Trait-correlation-network-and-plot

#Correlation plot naming

colnames(corr) <- c(
  "Dry mass",
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC"
)

rownames(corr) <- c(
  "Dry mass",
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC"
)


corrplot_export <- corrplot(corr,
         method = "circle",
         type = "lower",
         order = "hclust",
         diag = FALSE,
         col = colorRampPalette(c("#B2182B", "white", "#2166AC"))(200),
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.9,
         cl.cex = 0.8,
         addCoef.col = "black",
         number.cex = 0.7,
         mar = c(0,0,1,0)) +
  scale_size_area(max_size = 15) +
  geom_text(aes(label = sprintf("%.2f", corr)),
            size = 3,
            colour = "black") +
  mtext("Pearson correlation (r)",
        side = 4,
        line = 3,
        cex = 0.9)
```

```{r}
#| label: Dataframes-PCA

traits_NiN_age <- cleaned_data[, c(
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC",
  "age_class",
  "calluna_NiN_type"
)]

colnames(traits_NiN_age) <- c(
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC",
  "age_class",
  "calluna_NiN_type"
)

#Omit NA values
traits_NiN_age <- na.omit(traits_NiN_age)
```

```{r}
#| label: PCA-by-vegan-package

#improve visuals

custom_labels <- data.frame(
  Labels_traits = c("Leaf area", "SLA", "Vegetative height", "Leaf thickness", "LDMC"),
  x = c(1.5,   3.6,  -2.9,  -2.4,  -1.5),  # ◄ Desired X positions on the plot
  y = c(2.8,  -0.3,  -1.4,   2.1,   1.5)   # ◄ Desired Y positions on the plot
)

# Run the PCA using rda()
# scale = TRUE standardizes the traits (equivalent to scale. = TRUE in prcomp)
pca_vegan <- rda(traits_compiled, scale = TRUE)
summary(pca_vegan)


# Get the trait loadings (equivalent to pca$rotation)
# In vegan, these are called 'species' scores
round(scores(pca_vegan, display = "species", choices = 1:2, scaling = 0), 3)

# Extract site scores and combine with your metadata (equivalent to pca$x)
pca_scores <- scores(pca_vegan, display = "sites", choices = 1:2)

scores <- cbind(
  traits_NiN_age,
  PC1 = pca_scores[, "PC1"],
  PC2 = pca_scores[, "PC2"]
)

# Trait loadings (variable vectors)
loadings <- as.data.frame(
  scores(
    pca_vegan,
    display = "species",
    choices = 1:2,
    scaling = 0
  )
)

loadings$Variable <- rownames(loadings)
loadings

# Calculate % variation for the plot axes
# eigenvals() extracts the eigenvalues from the vegan model
eigenvalues <- summary(eigenvals(pca_vegan))
pc1_percent <- round(eigenvalues["Proportion Explained", "PC1"] * 100, 1)
pc2_percent <- round(eigenvalues["Proportion Explained", "PC2"] * 100, 1)
```

```{r}
#| label: PCA-vegan-plot


#Visualize PCA_vegan results
#Can already see from PC1 that mature and pioneer separate in multivariate trait space, don't overlap much, whereas building is intermediate and overlaps with both

ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = age_class,
    shape = calluna_NiN_type
  )
) +
  scale_color_manual(values = AGE) +
  geom_point(size = 3) +
  stat_ellipse()+
  labs(
    x = paste0("PC1 (", pc1_percent, "%)"), 
    y = paste0("PC2 (", pc2_percent, "%)")  
  ) 



```

```{r}
#| label: pca-trait-loadings-plot

#Dataframe with site and species scores
site_scores <- as.data.frame(scores(pca_vegan, display = "sites"))
species_scores <- as.data.frame(scores(pca_vegan, display = "species"))
species_scores$Species <- rownames(species_scores)

# Calculate explained variance for axis labels
scaling_info <- summary(pca_vegan)$cont$importance
pc1_var <- round(scaling_info[2, 1] * 100, 1)
pc2_var <- round(scaling_info[2, 2] * 100, 1)

ggplot() +
  # Draw species vectors (arrows)
  geom_segment(data = species_scores,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "darkred", alpha = 0.7) +
  # Smart species labels that don't overlap
  geom_text_repel(data = species_scores, 
                  aes(x = PC1, y = PC2, label = Species),
                  color = "darkred", size = 3, fontface = "italic") +
  # Draw site points
  geom_point(data = site_scores, 
             aes(x = PC1, y = PC2), 
             color = "royalblue", size = 2.5, alpha = 0.8) +
  # Dynamic axis titles with % variance explained
  labs(x = paste0("PC1 (", pc1_var, "%)"), 
       y = paste0("PC2 (", pc2_var, "%)"),
       title = "Principal Component Analysis (PCA)") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  coord_equal()


```

```{r}
#| label: PCA-plot-coloured-by-age-class

library(grid)


## Scale arrows to match score space


arrow_scale <- min(
  diff(range(scores$PC1))/diff(range(loadings$PC1)),
  diff(range(scores$PC2))/diff(range(loadings$PC2))
) * 0.75

loadings <- loadings %>%
  mutate(
    PC1 = PC1 * arrow_scale,
    PC2 = PC2 * arrow_scale
  )

#Labels traits into loading dataframe

new_labels <- c(
  height_cm = "Height",
  leaf_area = "Area",
  leaf_thickness_mean = "Thickness",
  SLA = "SLA",
  LDMC = "LDMC"
)

loadings$Variable <- ifelse(
  loadings$Variable %in% names(new_labels),
  new_labels[loadings$Variable],
  loadings$Variable
)


#Offset arrow labels
loadings <- loadings %>%
  mutate(
    label_x = PC1 * 1.08,
    label_y = PC2 * 1.2
  )

#Add interaction titles to scores dataframe
scores$group <- interaction(scores$age_class,
                            scores$calluna_NiN_type, sep = " × ")

pca_age_color <- ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = age_class,
    shape = calluna_NiN_type
  )
) +

geom_hline(yintercept = 0,
           colour = "grey80",
           linetype = 2) +

geom_vline(xintercept = 0,
           colour = "grey80",
           linetype = 2) +

stat_ellipse(
  aes(group = age_class,
      colour = age_class),
  level = 0.68,
  linewidth = 0.7,
  show.legend = FALSE,
  alpha = 0.5
) +

geom_point(
  size = 2.2,
  alpha = 0.65
) +

geom_segment(
  data = loadings,
  aes(
    x = 0,
    y = 0,
    xend = PC1,
    yend = PC2
  ),
  inherit.aes = FALSE,
  colour = "grey25",
  linewidth = 0.5,
  arrow = arrow(length = unit(0.25,"cm"))
) +

geom_text_repel(
  data = loadings,
  aes(
    x = label_x,
    y = label_y,
    label = Variable
  ),
  inherit.aes = FALSE,
  colour = "grey20",
  size = 4,
  point.padding = 0.3,
  box.padding = 0.5,
  max.overlaps = Inf, 
  segment.color = "grey80"
) +

scale_colour_manual(values = AGE) +

scale_shape_manual(values = c(
  BH = 16,
  FH = 17,
  TH = 15
)) +

coord_equal() +

labs(
  title = "PCA - Plant Traits",
  x = paste0("PC1 (", pc1_percent, "%)"),
  y = paste0("PC2 (", pc2_percent, "%)"),
  colour = "Age class",
  shape = "NiN type"
) +

theme_classic(base_size = 13) +

theme(
  plot.title = element_text(
    face = "plain",
    hjust = 0.5,
    size = 18
  ),

  axis.title = element_text(
    face = "plain",
    size = 14
  ),

  axis.text = element_text(size = 12),

  legend.title = element_text(
    face = "plain",
    size = 13
  ),

  legend.text = element_text(size = 12),

  legend.position = "right"
) +
  facet_wrap(~calluna_NiN_type)

pca_age_color

#Saving output

ggsave(
 "PCA_plot_age.png",
 width=10,
 height=8,
 dpi=300
)

```

```{r}
#| label: PCA-plot-coloured-by-NiN-type


#Offset arrow labels
loadings <- loadings %>%
  mutate(
    label_x2 = PC1 * 1.09,
    label_y2 = PC2 * 1.085
  )

pca_NiN_color <- ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = calluna_NiN_type,
    fill = calluna_NiN_type
  )
) +

geom_hline(yintercept = 0,
           colour = "grey80",
           linetype = 2) +

geom_vline(xintercept = 0,
           colour = "grey80",
           linetype = 2
) +

  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_point(
    size = 2.2,
    alpha = 0.5
  ) +

geom_segment(
  data = loadings,
  aes(
    x = 0,
    y = 0,
    xend = PC1,
    yend = PC2
  ),
  inherit.aes = FALSE,
  colour = "grey25",
  linewidth = 0.5,
  arrow = arrow(length = unit(0.25,"cm"))
) +

geom_text_repel(
  data = loadings,
  aes(
    x = label_x2,
    y = label_y2,
    label = Variable
  ),
  inherit.aes = FALSE,
  colour = "grey20",
  size = 4,
  point.padding = 0.5,
  box.padding = 0.35,
  max.overlaps = Inf, 
  segment.color = NA
) +

scale_colour_manual(values = NIN) +
  scale_fill_manual(values = NIN) +

coord_equal() +

labs(
  title = "PCA - Plant Traits",
  x = paste0("PC1 (", pc1_percent, "%)"),
  y = paste0("PC2 (", pc2_percent, "%)"),
  colour = "NiN type",
  shape = "Age class"
) +

theme_classic(base_size = 13) +

theme(
  plot.title = element_text(
    face = "plain",
    hjust = 0.5,
    size = 18
  ),

  axis.title = element_text(
    face = "plain",
    size = 14
  ),

  axis.text = element_text(size = 12),

  legend.title = element_text(
    face = "plain",
    size = 13
  ),

  legend.text = element_text(size = 12),

  legend.position = "right"
) +
  guides(fill = "none") +
  facet_wrap(~age_class)

pca_NiN_color

#Saving output

pca_NiN_color <- pca_NiN_color +
  theme(plot.title = element_blank())

ggsave(
 "pca_NiN_color.png",
 width=16,
 height=12,
 dpi=300
)

```

```{r}
#| label: betadisper-anova-permutest-TukeyHSD-NiN

#Test if FH has a stronger multivariate separation that BH and TH 
#The FH group exhibited the largest confidence ellipse, suggesting greater multivariate dispersion in functional trait composition than BH and TH, whereas TH showed the most compact distribution


# Scale and calculate distances
dist_mat <- dist(scale(traits_compiled))
show(dist_mat)

# Check dimensions
nrow(traits_compiled_complete)
nrow(traits_NiN_age_complete)
attr(dist_mat, "Size")
length(traits_NiN_age_complete$calluna_NiN_type)


#betadisper test NiN
bd_NiN <- betadisper(dist_mat, traits_NiN_age$calluna_NiN_type) 

anova(bd_NiN)
permutest(bd_NiN)
TukeyHSD(bd_NiN) #Results significant for BH-FH and TH-FH, but not for TH-BH, indicate FH differed significantly in dispersion from the other two NiN types

#Plot to show dispersion
plot(bd_NiN, kind = "boxplot", xlab = "NiN type", ylab = "Distance to centroid")



```

```{r}
#| label: betadisper-anova-permutest-TukeyHSD-Age

#betadisper test AGE
bd_age <- betadisper(dist_mat, traits_NiN_age$age_class)

anova(bd_age)
permutest(bd_age, pairwise = TRUE, permutations = 999) #Give unadjusted p-values
TukeyHSD(bd_age) #Adjusts for multiple comparisons

# Plot to show dispersion
plot(bd_age, kind = "boxplot", xlab = "Age class", ylab = "Distance to centroid")

#Result meets assumption of homogeneity, but shows no significant difference in the multivariate dispersion in functional trait composition across age class. Hypothesis disproved: trait variance does not significantly decrease with age. 


```

```{r}
#| label: betadisper-anova-permutest-TukeyHSD-NiN

#Testing interaction, creating a merged interaction factor allows to test whether the multivariate spread of plant traits shifts across different age stages within or across NiN types


# Creating a combined interaction factor
traits_NiN_age$interaction_group <- interaction(
  traits_NiN_age$calluna_NiN_type, 
  traits_NiN_age$age_class, 
  sep = "_"
)

# betadisper model on the interaction
bd_interaction <- betadisper(dist_mat, traits_NiN_age$interaction_group)

# Global significance tests
anova(bd_interaction)
permutest(bd_interaction, pairwise = TRUE, permutations = 999)

# Post-hoc comparisons (highly recommended for interactions)
tukey_interaction <- TukeyHSD(bd_interaction)
print(tukey_interaction)

# Plot ordered by the combined groups
plot_bd_interaction <- plot(bd_interaction, kind = "boxplot", 
     xlab = "NiN Type × Age Class", 
     ylab = "Distance to centroid",
     las = 2) # Rotates x-axis labels for readability


```

```{r}
#| label: boxplot-betadisper

#Boxplot with base r to visualize dispersion differences

Boxplot_bd_interaction <- boxplot(
  bd_interaction,
  xlab = "NiN Type × Age Class",
  ylab = "Distance to centroid",
  col = "grey85",
  border = "grey30",
  las = 2,
  cex.axis = 0.85,
  cex.lab = 1.1,
  lwd = 1.2,
  outline = TRUE,
  frame = FALSE
)

Boxplot_bd_interaction

```

```{r}
#| label: boxplot-betadisper-facet-by-age

#Boxplot with ggplot and significance

bd_interaction_data <- data.frame(
  distance = bd_interaction$distances,
  calluna_NiN_type = traits_NiN_age$calluna_NiN_type,
  age_class = traits_NiN_age$age_class
)

# Create interaction group
bd_interaction_data$group <- interaction(
  bd_interaction_data$calluna_NiN_type,
  bd_interaction_data$age_class,
  sep = " "
)

# Set order wettest to driest, BH to TH
bd_interaction_data$group <- factor(
  bd_interaction_data$group,
  levels = c(
    "BH Pioneer", "BH Building", "BH Mature",
    "FH Pioneer", "FH Building", "FH Mature",
    "TH Pioneer", "TH Building", "TH Mature"
  )
)


#plot NIN, only shows 2 out of 4 interactions

bd_interaction_plot_NiN <- ggplot(
  bd_interaction_data,
  aes(x = calluna_NiN_type, y = distance, fill = calluna_NiN_type)
) +
  geom_boxplot(
    width = 0.65,
    outlier.shape = 16
  ) +
  # Pioneer: BH vs FH
  geom_signif(
    data = subset(
      bd_interaction_data,
      age_class == "Pioneer"
    ),
    comparisons = list(
      c("BH", "FH")
    ),
    annotations = "*",
    y_position = 5,
    tip_length = 0.01,
    textsize = 5
  ) +
  
  # Mature: BH vs TH
  geom_signif(
    data = subset(
      bd_interaction_data,
      age_class == "Mature"
    ),
    comparisons = list(
      c("BH", "TH")
    ),
    annotations = "*",
    y_position = 5,
    tip_length = 0.01,
    textsize = 5
  ) +
  
  facet_wrap(
    ~ age_class,
    nrow = 1
  ) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "NiN Type",
    y = "Distance to centroid"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "plain",
      size = 12
    ),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    plot.margin = margin(5, 5, 5, 5)
  )

bd_interaction_plot_NiN

#Plot by age

bd_interaction_plot_age <- ggplot(
  bd_interaction_data,
  aes(
    x = age_class,
    y = distance,
    fill = age_class
  )
) +
  geom_boxplot(
    width = 0.65,
    outlier.shape = 16
  ) +
  
  # FH: Pioneer vs Building
  geom_signif(
    data = subset(
      bd_interaction_data,
      calluna_NiN_type == "FH"
    ),
    comparisons = list(
      c("Pioneer", "Building")
    ),
    annotations = "*",
    y_position = 4.9,
    tip_length = 0.01,
    textsize = 5
  ) +
  
  # TH: Building vs Mature
  geom_signif(
    data = subset(
      bd_interaction_data,
      calluna_NiN_type == "TH"
    ),
    comparisons = list(
      c("Building", "Mature")
    ),
    annotations = "*",
    y_position = 4.9,
    tip_length = 0.01,
    textsize = 5
  ) +
  
  facet_wrap(
    ~ calluna_NiN_type,
    nrow = 1
  ) +
  
  scale_fill_manual(
    values = AGE
  ) +
  
  labs(
    x = "Age Class",
    y = "Distance to centroid"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "plain",
      size = 12
    ),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    plot.margin = margin(5, 5, 5, 5)
  )

bd_interaction_plot_age


bd_combined <- bd_interaction_plot_age / bd_interaction_plot_NiN +
  plot_annotation(
    tag_levels = "A"
  )

bd_combined
```

```{r}
#| label: RDA-analysis-anova


rda <- rda(traits_compiled ~ calluna_NiN_type * age_class, data = traits_NiN_age)
summary(rda)
plot(rda)

RsquareAdj(rda) #Total variance explained by rda

#Significance test
anova(rda, permutations = 999) #Nothing to do with ANOVA, pseudo F-value, but the closest you get to ANOVA with ordination

anova(rda, by ="axis") #Axis significance
anova(rda, by = "terms") #significance by age class and Nin

#Look at multicollinearity, >2 is considered high, variables are fine
sqrt(vif.cca(rda))

#Extract RDA scores

# Site scores
sites_rda <- as.data.frame(scores(rda, display = "sites"))
sites_rda$age_class <- traits_NiN_age$age_class
sites_rda$calluna_NiN_type <- traits_NiN_age$calluna_NiN_type

# Species scores
scores_rda <- cbind(
  traits_NiN_age,
  RDA1 = sites_rda[, "RDA1"],
  RDA2 = sites_rda[, "RDA2"]
)

# Trait loadings (variable vectors)
loadings_rda <- as.data.frame(
  scores(
    rda,
    display = "species",
    choices = 1:2,
    scaling = 0
  )
)

loadings_rda$Variable <- rownames(loadings_rda)

#Naming variable rows
new_labels <- c(
  "Vegetative height" = "Height",
  "Leaf area" = "Area",
  "Leaf thickness" = "Thickness",
  SLA = "SLA",
  LDMC = "LDMC"
)

loadings_rda$Variable <- ifelse(
  loadings_rda$Variable %in% names(new_labels),
  new_labels[loadings_rda$Variable],
  loadings_rda$Variable
)

loadings_rda

# Environmental arrows
env_arrows <- as.data.frame(scores(rda, display = "bp"))
env_arrows$variable <- rownames(env_arrows)

#Percent variance explained
eig <- eigenvals(rda)

pc1_rda <- round(100 * eig[1] / sum(eig), 2)
pc2_rda <- round(100 * eig[2] / sum(eig), 2)


#Offset arrow labels
loadings_rda <- loadings_rda %>%
  mutate(
    label_x3 = RDA1 * 1.25,
    label_y3 = RDA2 * 1.25)
```

```{r}
#| label: plot-rda-facet-by-age



 rda_plot <- ggplot(sites_rda, aes(RDA1, RDA2, colour = calluna_NiN_type, fill = calluna_NiN_type)) +
  
  geom_point(size = 1.5, alpha = 0.85) +
  
  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_segment(
    data = loadings_rda,
    aes(x = 0, y = 0,
        xend = RDA1*2,
        yend = RDA2*2),
    linewidth = 0.7,
    colour = "grey20",
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")
   )
  ) +
  geom_text_repel(
    data = loadings_rda,
    aes(
      x = label_x3,
      y = label_y3,
      label = Variable
    ),
    force = 12,
    inherit.aes = FALSE,
    colour = "grey20",
    size = 4,
    point.padding = 0.3,
    box.padding = 4,
    max.overlaps = Inf, 
    segment.color = "grey40",
    segment.size = 0.4,
    min.segment.length = 0
  ) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right",
    strip.text = element_text(face = "plain")
  ) +
    coord_equal() +
  geom_hline(yintercept = 0,
             linetype = 2,
             colour = "grey70") +

geom_vline(xintercept = 0,
           linetype = 2,
           colour = "grey70") +
    labs(
      title = "RDA - Plant Traits",
      x = paste0("RDA1 (", pc1_rda, "%)"),
      y = paste0("RDA2 (", pc2_rda, "%)"),
      colour = "NiN type"
    ) +
  guides(fill = "none") +
  facet_wrap(~age_class)
  
  rda_plot

#Arrows are offset by 2 * multiplication for aesthetics!

#Saving output
ggsave(
  "rda_plot.png",
  width=12,
  height=10,
  dpi=300
)
```

```{r}
#| label: plot-rda-facet-by-age-and-NiN

#Interaction column
sites_rda$facet_group <- interaction(
  sites_rda$calluna_NiN_type,
  sites_rda$age_class
)

ggplot(sites_rda, aes(RDA1, RDA2, colour = calluna_NiN_type, fill = calluna_NiN_type)) +
  
  geom_point(size = 1.5, alpha = 0.85) +
  
  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_segment(
    data = loadings_rda,
    aes(x = 0, y = 0,
        xend = RDA1*2,
        yend = RDA2*2),
    linewidth = 0.7,
    colour = "grey20",
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")
    )
  ) +
  geom_text_repel(
    data = loadings_rda,
    aes(
      x = label_x3,
      y = label_y3,
      label = Variable
    ),
    force = 12,
    inherit.aes = FALSE,
    colour = "grey20",
    size = 4,
    point.padding = 0.3,
    box.padding = 4,
    max.overlaps = Inf, 
    segment.color = "grey40",
    segment.size = 0.4,
    min.segment.length = 0
  ) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right",
    strip.text = element_text(face = "plain")
  ) +
  coord_equal() +
  geom_hline(yintercept = 0,
             linetype = 2,
             colour = "grey70") +
  
  geom_vline(xintercept = 0,
             linetype = 2,
             colour = "grey70") +
  labs(
    title = "RDA - Plant Traits",
    x = paste0("RDA1 (", pc1_rda, "%)"),
    y = paste0("RDA2 (", pc2_rda, "%)"),
    colour = "NiN type"
  ) +
  guides(fill = "none") +
  facet_wrap(~facet_group)



```

# Presenting results wet mass

```{r}
#| label: Plots-to-visualize-wet-mass


emmip(emmeans_lme4_wet_mass, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_wet_mass, comparisons=TRUE) #These red arrows are specially calculated so that any overlap means "not significant" and no overlap means "significant."



#dataframe with multicomp, adding significance groups from emmeans
cld_wet_mass <- cld(
  emmeans_lme4_wet_mass,
  adjust = "tukey",
  Letters = letters
)

multicomp_wet_mass_df <- left_join(
  df_sum_wet_mass,
  cld_wet_mass %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_wet_mass)
contrasts_wet_mass_df <- as.data.frame(pairs(emmeans_lme4_wet_mass))
head(contrasts_wet_mass_df)

contrasts_wet_mass_df <- contrasts_wet_mass_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: Raincloud-plot-half-violins-jittered-points-and-EMMs-wet-mass

#Place letters above the tallest datapoint in each age class


cld_wet_mass <- cld_wet_mass %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.0075,
      age_class == "Building" ~ 0.0075,
      age_class == "Mature"   ~ 0.0075
    )
  )

ggplot(cleaned_data,
       aes(x = age_class,
           y = wet_mass,
           fill = calluna_NiN_type,
           colour = calluna_NiN_type)) +
  
  # Half violin ("cloud")
  stat_halfeye(
    adjust = 0.5,
    width = 0.4,
    justification = -0.25,   # moves violin left
    point_colour = NA,
    alpha = 0.15,
    position = position_dodge(width = 0.8)
  ) +
  
  # Raw points ("rain")
  geom_quasirandom(
    dodge.width = 0.8,
    width = 0.12,
    alpha = 0.4,
    size = 1
  ) +
  
  # EMMs
  geom_pointrange(
    data = cld_wet_mass,
    aes(
      x = age_class,
      y = emmean,
      ymin = lower.CL,
      ymax = upper.CL,
      group = calluna_NiN_type
    ),
    colour = "black",
    size = 0.7,
    fatten = 2,
    position = position_dodge(width = 0.8),
    inherit.aes = FALSE
  ) +
  
  # CLD letters
  geom_text(
    data = cld_wet_mass,
    aes(
      x = age_class,
      y = label_y,
      label = trimws(.group),
      group = calluna_NiN_type
    ),
    colour = "black",
    fontface = "bold",
    position = position_dodge(width = 0.8),
    inherit.aes = FALSE
  ) +
  
  scale_fill_manual(values = NIN) +
  scale_colour_manual(values = NIN) +
  
  labs(
    x = "Age class",
    y = "Wet mass (g)",
    title = "Mean leaf wet mass (g) of C. vulgaris across age class and NiN types",
    colour = "NiN type",
    fill = "NiN type"
  ) +
  
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )+
  guides(fill = "none")
```

Best visuals so far, easy overview, less noise

```{r}
#| label: Jittered-plot-facet-by-age-wet-mass

results_ggp_wet_mass <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = wet_mass, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_wet_mass, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x=NULL, y="Wet mass (g)", title="Leaf wet mass (g) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    legend.justification = "center"
  ) +
facet_wrap(~ age_class)

results_ggp_wet_mass

results_export_wet_mass <- results_ggp_wet_mass +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_wet_mass.png", plot = results_export_wet_mass, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Boxplot-with-raw datapoints-significance-lines-and-letters-wet-mass



# 1. Finn maks-verdien for wet_mass for å styre høyden på brakettene automatisk
max_y <- max(cleaned_data$wet_mass, na.rm = TRUE)

# 2. Data frame with manual star mapping based on your p-values
signif_wet_mass <- data.frame(
  xmin = c(
    2 - 0.33, # Building: BH
    2,        # Building: FH
    1,        # FH: Pioneer to Mature
    1 + 0.33, # TH: Pioneer to Building
    2 + 0.33  # TH: Building to Mature
  ),
  xmax = c(
    2 + 0.33, # Building: TH
    2 + 0.33, # Building: TH
    3,        # FH: Mature
    2 + 0.33, # Building: TH
    3 + 0.33  # Mature: TH
  ),
  # Height tiers to stack them neatly above the letters
  y_position = c(
    max_y * 1.30, 
    max_y * 1.45, 
    max_y * 1.60, 
    max_y * 1.75, 
    max_y * 1.90  
  ),
  # UPDATED: P-values mapped directly to star ratings
  annotation = c(
    "***", # p = 0.0005 (is < 0.001)
    "***", # p = 0.0002 (is < 0.001)
    "*",   # p = 0.0109 (is < 0.05)
    "***", # p = 0.0003 (is < 0.001)
    "***"  # p = 0.0007 (is < 0.001)
  ),
  group = 1:5 
)

# 3. Build the plot
ggp2_wet_mass <- ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color = calluna_NiN_type)) +
  geom_quasirandom(
    dodge.width = 1,
    alpha = 0.5,
    size = 1,
    width = 0.1
  ) +
  geom_boxplot(
    aes(
      fill = calluna_NiN_type,
      group = interaction(age_class, calluna_NiN_type)
    ),
    color = "black",
    alpha = 0.40,
    width = 0.6,
    position = position_dodge(width = 1)
  ) +
  geom_text(
    data = cld_wet_mass,
    aes(
      y = label_y,
      label = trimws(.group),
      group = calluna_NiN_type
    ),
    position = position_dodge(width = 1),
    vjust = -0.5,
    color = "black",
    fontface = "bold",
    show.legend = FALSE
  ) +
  
  # Signif Bracket Layer
  ggsignif::geom_signif(
    data = signif_wet_mass,
    aes(
      xmin = xmin, 
      xmax = xmax, 
      y_position = y_position, 
      annotations = annotation,
      group = group
    ),
    manual = TRUE,
    inherit.aes = FALSE, 
    color = "black",
    tip_length = 0.02,
    # Increased text size slightly since stars are smaller than text string expressions
    textsize = 4.2, 
    vjust = 0.4 # Centers stars vertically right on top of the bracket line
  ) +
  
  scale_color_manual(values = NIN, name = "NiN type") +
  scale_fill_manual(values = NIN, name = "NiN type") +
  labs(
    x = "Age class", 
    y = "Mean leaf wet mass (g)", 
    title = expression("Mean leaf wet mass (g) of " ~ italic("C. vulgaris") ~ " across age class and NiN types"), 
    color = "NiN type"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.10))
  ) +
  theme_classic() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  geom_segment(
    aes(x = 1.5, xend = 1.5,
        y = 0, yend = 0.0072),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )+
  geom_segment(
    aes(x = 2.5, xend = 2.5,
        y = 0, yend = 0.0072),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )

ggp2_wet_mass

# Save with specific dimensions and high DPI
ggsave("ggp2_wet_mass.png", plot = ggp2_wet_mass, width = 7, height = 4, dpi = 300)

```

# Presenting results height

```{r}
#| label: Visualize-height-emmeans-outputs


emmip(emmeans_height_cm, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_height_cm, comparisons=TRUE) #These red arrows are specially calculated so that any overlap means "not significant" and no overlap means "significant."

#dataframe with multicomp, adding significance groups from emmeans
cld_height <- cld(
  emmeans_height_cm,
  adjust = "tukey",
  Letters = letters
)

multicomp_height_df <- left_join(
  df_sum_height,
  cld_height %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
contrasts_height_df <- as.data.frame(pairs(emmeans_height_cm))
head(contrasts_height_df)

contrasts_height_df <- contrasts_height_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )




```

```{r}
#| label: boxplot-jitter-height

#Place letters above the tallest datapoint in each age class

cld_height <- cld_height %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 54,
      age_class == "Building" ~ 54,
      age_class == "Mature"   ~ 54
    )
  )

ggp_height <- ggplot(
    cleaned_data,
    aes(
      x = age_class,
      y = height_cm,
      color = calluna_NiN_type,
      group = calluna_NiN_type,
      fill = calluna_NiN_type,
    )
  ) +
  stat_unique(
  geom = "point",
  position = position_dodge(width = 1),
  alpha = 0.8, 
  size = 1.8
) +
    geom_boxplot(
      aes(
        group = interaction(age_class, calluna_NiN_type)
      ),
      color = "black",
      alpha = 0.35,
      width = 0.5,
      position = position_dodge(width = 1)
    ) +
    geom_text(
      data = cld_height,
      aes(
        y = label_y,
        label = trimws(.group),
        group = calluna_NiN_type
      ),
      position = position_dodge(width = 1),
      vjust = -0.5,
      color = "black",
      fontface = "bold",
      show.legend = FALSE
    ) +
    scale_color_manual(values = NIN, name = "NiN type") +
    scale_fill_manual(values = NIN, name = "NiN type") +
    labs(x = "Age class", y = "Vegetative height (cm)", title = "Vegetative height (cm) of" ~ italic("C. vulgaris") ~ "across age class and NiN type", color = "NiN type") +
    scale_y_continuous(
      expand = expansion(mult = c(0.02, 0.15))
    ) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )


ggp_height

# Save with specific dimensions and high DPI
ggsave("ggp_height.png", plot = ggp_height, width = 7, height = 4, dpi = 300)
  

```

For the report:

```{r}
#| label: pointrange-EMM-results-plot-height


results_ggp_height <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = height_cm, color = calluna_NiN_type)) +
  stat_unique(
    geom = "point",
    position = position_dodge(width = 1),
    alpha = 0.4, 
    size = 1.8
  ) +
  geom_pointrange(data=cld_height, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y="Vegetative height (cm)", title="Vegetative height (cm) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_height

results_export_height <- results_ggp_height +
  theme(plot.title = element_blank())


# Save with specific dimensions and high DPI
ggsave("results_export_height.png", plot = results_export_height, width = 7, height = 4, dpi = 300)
```

# Presenting results dry mass

```{r}
#| label: Visualizing-dry-mass-emmeans-outputs


emmip(emmeans_lme4_dry_mass, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_dry_mass, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other

#dataframe with multicomp, adding significance groups from emmeans
cld_dry_mass <- cld(
  emmeans_lme4_dry_mass,
  adjust = "tukey",
  Letters = letters
)

multicomp_dry_mass_df <- left_join(
  df_sum_dry_mass,
  cld_dry_mass %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_dry_mass)
contrasts_dry_mass_df <- as.data.frame(pairs(emmeans_lme4_dry_mass))
head(contrasts_dry_mass_df)

contrasts_dry_mass_df <- contrasts_dry_mass_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )



```

This is used in the report:

```{r}
#| label: quasirandom-jitter-plot-dry-mass



results_ggp_dry_mass <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = dry_mass, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_dry_mass, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x=NULL, y="Dry mass (g)", title="Leaf dry mass (g) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_dry_mass

results_export_dry_mass <- results_ggp_dry_mass +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_dry_mass.png", plot = results_export_dry_mass, width = 7, height = 4, dpi = 300)
```

# Presenting results leaf area

```{r}
#| label: Visualize-leaf-area-emmeans-outputs

emmip(emmeans_lme4_leaf_area, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_leaf_area, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other

#dataframe with multicomp, adding significance groups from emmeans
cld_leaf_area <- cld(
  emmeans_lme4_leaf_area,
  adjust = "tukey",
  Letters = letters
)

multicomp_leaf_area_df <- left_join(
  df_sum_leaf_area,
  cld_leaf_area %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_leaf_area)
contrasts_leaf_area_df <- as.data.frame(pairs(emmeans_lme4_leaf_area))
head(contrasts_leaf_area_df)

contrasts_leaf_area_df <- contrasts_leaf_area_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: quasirandom-jitter-leaf-area

#Place letters above the tallest datapoint in each age class

cld_leaf_area <- cld_leaf_area %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.20,
      age_class == "Building" ~ 0.20,
      age_class == "Mature"   ~ 0.20
    )
  )



results_ggp_leaf_area <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_area, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_leaf_area, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x= NULL, y = expression(Leaf ~ area ~ (cm^2)), color = "NiN type", title=expression("Leaf area" ~ (cm^2) ~ "of" ~ italic("C. vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_leaf_area

results_export_leaf_area <- results_ggp_leaf_area +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_leaf_area.png", plot = results_export_leaf_area, width = 7, height = 4, dpi = 300)
```

# Presenting results leaf thickness

```{r}
#| label: Plots-visualizing-emmeans-output-leaf-thickness


emmip(emmeans_lme4_leaf_thickness, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_leaf_thickness, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_leaf_thickness <- cld(
  emmeans_lme4_leaf_thickness,
  adjust = "tukey",
  Letters = letters
)

multicomp_leaf_thickness_df <- left_join(
  df_sum_leaf_thickness,
  cld_leaf_thickness %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_leaf_thickness)
contrasts_leaf_thickness_df <- as.data.frame(pairs(emmeans_lme4_leaf_thickness))
head(contrasts_leaf_thickness_df)

contrasts_leaf_thickness_df <- contrasts_leaf_thickness_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

This is in the report:

```{r}
#| label: quasirandom-jitter-leaf-thickness



results_ggp_leaf_thickness <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_thickness_mean, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_leaf_thickness, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = "Leaf thickness (mm)", color = "NiN type", title = "Mean leaf thickness (mm) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.7),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_leaf_thickness

results_export_leaf_thickness <- results_ggp_leaf_thickness +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_leaf_thickness.png", plot = results_export_leaf_thickness, width = 7, height = 4, dpi = 300)
```

# Presenting results SLA

```{r}
#| label: Plots-visualize-emmeans-output-SLA

emmip(emmeans_lme4_SLA, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_SLA, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_SLA <- cld(
  emmeans_lme4_SLA,
  adjust = "tukey",
  Letters = letters
)

multicomp_SLA_df <- left_join(
  df_sum_SLA,
  cld_SLA %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_SLA)
contrasts_SLA_df <- as.data.frame(pairs(emmeans_lme4_SLA))
head(contrasts_SLA_df)

contrasts_SLA_df <- contrasts_SLA_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: quasirandom-jitter-plot-SLA



results_ggp_SLA <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = SLA, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_SLA, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = expression(SLA ~ (cm^2 ~ g^-1)), color = "NiN type", title = expression("Specific leaf area" ~ (cm^2 ~ g^-1) ~ "of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.95),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_SLA

results_export_SLA <- results_ggp_SLA +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_SLA.png", plot = results_export_SLA, width = 7, height = 4, dpi = 300)
```

# Presenting results LDMC

```{r}
#| label: Plots-visualizing-emmeans-output-LDMC


emmip(emmeans_lme4_LDMC, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_LDMC, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_LDMC <- cld(
  emmeans_lme4_LDMC,
  adjust = "tukey",
  Letters = letters
)

multicomp_LDMC_df <- left_join(
  df_sum_LDMC,
  cld_LDMC %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_LDMC)
contrasts_LDMC_df <- as.data.frame(pairs(emmeans_lme4_LDMC))
head(contrasts_LDMC_df)

contrasts_LDMC_df <- contrasts_LDMC_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

In the report:

```{r}
#| label: quasirandom-jitter-plot-LDMC



results_ggp_LDMC <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = LDMC, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_LDMC, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = expression(LDMC ~ (g ~ g^-1)), color = "NiN type", title = expression("Leaf dry matter content" ~ (g ~ g^-1) ~ "of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_LDMC

results_export_LDMC <- results_ggp_LDMC +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_LDMC.png", plot = results_export_LDMC, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: SLA-LDMC-relationship


ggplot(cleaned_data,
       aes(x = SLA, y = LDMC)) +
  
  # scatter points
  geom_point(aes(color = calluna_NiN_type),
             size = 1.5, alpha = 0.4) +
  
  # fitted regression line by age class
  geom_smooth(
    aes(colour = calluna_NiN_type),
    method = "lm",
    se = TRUE,
    linewidth = 1.3,
    colour = "black"
  ) +
  
  # optional: label
  labs(
    x = "SLA",
    y = "LDMC",
    title = "SLA vs LDMC",
    color = "NiN type"
  ) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )
```
