---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format:
  html:
    embed-resources: true
editor: visual
---

# Importing data

```{r}
#| label: load-libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(here)
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(ggridges) #Moving density plots
library(rstatix)
library(car)
library(performance) #checking model normality, homogeneity, variance
library(emmeans) #Post hoc testing
library(afex) #linear mixed models
library(lme4) #linear mixed models
library(sdamr) #linear mixed models
library(lmerTest)
library(multcomp) #emmeans contrasts on plots
library(factoextra) #PCA biplot
library(corrplot)
library(vegan) #RDA analysis
library(ggdist) 
library(ggrepel) #RDA plot making
library(ggbeeswarm)
library(ggnewscale) #Visuals of PCA
library(pairwiseAdonis) #pairwise PERMANOVA # To install pairwiseAdonis: remotes::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")



theme_set(theme_bw())
theme_update(plot.title.position = "plot")



```

```{r}
#| label: importing-raw-data-of-traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv(here::here("data/clean_data/cleaned_data_copy.csv"))



# Order relevant variables (useful for plotting and order based analyses)
# age_class sorted from youngest to oldest, NiN type sorted from wettest to driest
cleaned_data$calluna_NiN_type <- factor(cleaned_data$calluna_NiN_type,  levels = c("BH","FH","TH"))
cleaned_data$age_class <- factor(cleaned_data$age_class, levels = c("Pioneer", 
                                        "Building", "Mature"))


#Add column of mean leaf thickness to cleaned data
cleaned_data$leaf_thickness_mean <- rowMeans(cleaned_data[, c("leaf_thickness_1", "leaf_thickness_2", "leaf_thickness_3")], na.rm=TRUE)



```

```{r}
#| label: Data-visualization-and-overview

summary(cleaned_data)
glimpse(cleaned_data)


## Define colour palette (used with scale fill manual) 

NIN <- c("BH" = "darkslategrey", "FH" = "deepskyblue4", "TH" = "lightskyblue")
AGE <- c("Pioneer" = "#087E8B","Building" = "#EB5461", "Mature" = "#884000") 




```

# Wet mass(g) visualizations

-Wet mass appears decreasing from Pio to Mat for BH and FH, but for TH the wet mass for Mat has increased to the level of Pio after decreasing in BUI, test for significance between BUI: FH vs TH

```{r}
#| label: Data-frame-with-mean-median-and-se-histogram-wet-mass

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE), 
            median_wet_mass= median.default(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()), 
            n = n(), .groups = "drop")

#Histograms, general and NiN and age class specific

ggplot(cleaned_data, aes(x = wet_mass)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1) 

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NiN_type ~ age_class) +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")


```

Boxplot to check overlapping error bars and data medians

```{r}
#| label: Boxplot-wet-mass

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Wet mass (g)")
```

```{r}
#| label: Violin-plot-wet-mass

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)
```

```{r}
#| label: Interactive-plots-wet-mass


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NIN type",
    y = "Mean wet mass (g)",
    color = "Age class"
)
```

```{r}
#| label: Separating-plant_nr-for-each-factor-age-class-and-NiN-wet-mass

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, group = plant_nr)) +
  # Draw the raw data points for every plant
  geom_point(aes(color = as.factor(plant_nr)), alpha = 0.3) +
  # Draw the predicted random slope line for every single plant
  #geom_line(aes(y = wet_mass, color = as.factor(plant_nr)), size = 0.8) +
  # Split the graph into panels by Calluna NiN Type!
  facet_wrap(~ calluna_NiN_type) +
  labs(title = "Effect of plant nr across All Groups",
       x = "Age Class", 
       y = "Wet Mass",
       color = "Plant Number")
```

```{r}
#| label: Density-Plot-wet-mass

ggplot(cleaned_data, aes(x = wet_mass, y =calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges (alpha = 0.5)+ #To separate the density plots vertically
    facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot wet mass", x = "Wet mass")



```

# Dry mass(g) visualizations

-General histogram: appears slightly right-skewed

-similar trend to wet mass, but looks like in FH the BUI dry mass increases, then declines for Mat and for TH it decreases for BUI, then increases to Mat @fig-Data-frame-histogram-dry-mass shows ...\

```{r}
#| label: fig-Data-frame-histogram-dry-mass

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN and age class specific

ggplot(cleaned_data, aes(x = dry_mass)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1)


ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot dry mass

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Dry mass (g)")

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = dry_mass, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Dry mass (g)")
```

```{r}
#| label: Violin plot dry mass

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Dry mass (g)")
```

```{r}
#| label: Interactive plot dry mass

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)

#swapped axis, NiN type on x-axis

ggplot(df_sum_dry_mass, aes(x = calluna_NiN_type, y = mean_dry_mass, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)
```

```{r}
#| label: Density Plot dry mass

ggplot(cleaned_data, aes(x = dry_mass, y = calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot dry mass (g)", x = "Dry mass (g)") 



```

# Height (cm)

-does not appear normally distributed, histogram looks bimodal, but might be too few datapoints

-For height, more variation in MAT, but TH shows a different trend than FH and BH again. The BUI for TH is acting differently that the other nature types

-Plants generally taller in BH and FH, might be significant between PIO: BH vs FH and TH and for MAT: BH, FH vs TH

-From the boxplot, the variances across age classes have quite a different scatter, MAT in BH and FH have a lot more variation than PIO i.e

```{r}
#| label: Data frame + histogram height

df_sum_height <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN and age class specific

ggplot(cleaned_data, aes(x = height_cm)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1)

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot height

Boxplot_height <- ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Vegetative height (cm)", title="Vegetative height (cm) of Calluna vulgaris across Age class and NiN type")

Boxplot_height

# Save with specific dimensions and high DPI
ggsave("Boxplot_height.png", plot = Boxplot_height, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: violin plot height

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Height (cm)")
```

```{r}
#| label: Interactive plot height


ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)

# swapped axis, NiN type on x-axis

ggplot(df_sum_height, aes(x = calluna_NiN_type, y = mean_height_cm, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "Mean height (cm)",
    color = "Age class"
)
```

```{r}
#| label: Density Plot height

ggplot(cleaned_data, aes(x = height_cm, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot height (cm)", x = "Height (cm)")


```

# Leaf thickness (mm) visualizations

-General: appears normally distributed, a little right-skewed

-leaf thickness quite overlapping, might be significant between BUI (consider outliers here are pulling the mean)

-thickness highest for PIO in TH, lowest in FH (supports an initial conservative strategy), in BUI thickest leaves in BH, thinnest in TH (more acquisitive), for MAT back to thicker leaves in BH, TH and thinner in FH (conservative)

\

```{r}
#| label: Dataframe + histogram leaf thickness

  
#create table with leaf thickness 1,2,3 combined

data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

# Data frame with mean and se

df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN, age class specific

ggplot(data_thickness, aes(x = leaf_thickness)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf thickness (mm)", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot leaf thickness

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NiN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)")


boxplot_leaf_thickness <- ggplot(data_thickness, aes(x = calluna_NiN_type, y = leaf_thickness, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Leaf thickness (mm)")

boxplot_leaf_thickness

# Save with specific dimensions and high DPI
ggsave("boxplot_leaf_thickness.png", plot = boxplot_leaf_thickness, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: violin plot leaf thickness

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "Leaf thickness (mm)")
```

```{r}
#| label: Interactive plot leaf thickness


interaction_plot_thickness <- ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness, width = 0.1)) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Age class",
    y = "Leaf thickness (mm)",
    color = "NiN type", title= "Mean leaf thickness (mm) of C. vulgaris against NiN type and age classes"
)

interaction_plot_thickness

# Save with specific dimensions and high DPI
ggsave("interaction_plot_thickness.png", plot = interaction_plot_thickness, width = 7, height = 4, dpi = 300)


#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_thickness, aes(x = calluna_NiN_type, y = mean_leaf_thickness, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "NiN type",
    y = "Mean leaf thickness (mm)",
    color = "Age class"
)
```

```{r}
#| label: Density Plot leaf thickness

ggplot(data_thickness, aes(x = leaf_thickness, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf thickness (mm)", x = "Leaf thickness (cm)") 


```

# Leaf area (cm^2^) visualizations

-general: appears right-skewed, but normally distributed

-leaf area for TH vs FH in PIO and BUI might be significant, generally greater leaf area from PIO→MAT (acquisitive), different trend in TH again. Greater variation in leaf area between age classes in FH

\

```{r}
#| label: Data frame + histogram leaf area

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram, general and NiN, age class specific

ggplot(cleaned_data, aes(x = leaf_area)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "Leaf area (cm^2))", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot leaf area

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (cm^2)")

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_area, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "Leaf area (cm^2)")
```

```{r}
#| label: violin plot leaf area

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NiN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(title = "Leaf area of Calluna in different age classes and NiN types", x = "Age class", y = "Leaf area (cm^2)")
```

```{r}
#| label: Interactive plot leaf area


ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (cm^2)",
    color = "Age class"
  ) 
 

# swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_area, aes(x = calluna_NiN_type, y = mean_leaf_area, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NiN type",
    y = "Mean leaf area (cm^2)",
    color = "Age class"
  )
```

```{r}
#| label: Density Plot leaf area

ggplot(cleaned_data, aes(x = leaf_area, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot leaf area (cm^2)", x = "Leaf area (cm^2)") 

```

# LDMC visualizations

-General: appears normally distributed, but left-skewed

-Slightly increasing LDMC trend from Pio to Mat (plants become more conservative with age), FH shows a different trend with BUI having the highest LDMC. test significance PIO vs BUI in FH, and BUI: FH vs TH vs BH, FH as the highest LDMC and BH the lowest, TH in the middle (might be overlapping)

\

```{r}
#| label: Data frame + histogram LDMC


#Add mean for each age class
mean_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(mean_LDMC_age = mean(LDMC, na.rm = TRUE),
            .groups = "drop")

#Add se for age class
se_age <- cleaned_data %>%
  group_by(age_class) %>%
  summarise(se_LDMC_age = sd(LDMC, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")

#Add mean for each NiN type
mean_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(mean_LDMC_NiN = mean(LDMC, na.rm = TRUE),
            .groups = "drop")

#Add se for NiN type
se_NiN <- cleaned_data %>%
  group_by(calluna_NiN_type) %>%
  summarise(se_LDMC_Nin = sd(LDMC, na.rm = TRUE)/sqrt(n()), n = n(),
            .groups = "drop")



df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")  %>%
  left_join(mean_age, by = "age_class") %>%
  left_join(mean_NiN, by = "calluna_NiN_type") %>%
  left_join(se_age, by = "age_class") %>%
  left_join(se_NiN, by = "calluna_NiN_type")



# Histogram

ggplot(cleaned_data, aes(x = LDMC)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  labs(x = "LDMC", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot LDMC

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")

ggplot(cleaned_data, aes(x = calluna_NiN_type, y = LDMC, fill = calluna_NiN_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(x = "NiN type", y = "LDMC")
```

```{r}
#| label: violin plot LDMC

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "LDMC")
```

```{r}
#| label: Interactive plot LDMC


ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(
    x = "Calluna NiN type",
    y = "LDMC",
    color = "Age class"
  )


#swapped axis, NiN type on x-axis

ggplot(df_sum_LDMC, aes(x = calluna_NiN_type, y = mean_LDMC, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(
    x = "Calluna NIN type",
    y = "LDMC",
    color = "Age class"
  )
```

```{r}
#| label: Density Plot LDMC

ggplot(cleaned_data, aes(x = LDMC, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot LDMC", x = "LDMC") 




```

# SLA visualizations

-General: appears right-skewed, Pio FH and Bui TH more scattered

-SLA generally highest for PIO, decreasing to MAT, except BUI in TH. SLA highest in FH for PIO, but quite similar in BH and TH for BUI and MAT. test significance in FH: Pio vs Bui and Mat, and for PIO FH vs TH vs BH

```{r}
#| label: Data frame + histogram SLA


df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NiN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = SLA)) + 
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 0.8) 

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(aes(bins=30, y = after_stat(density)), fill = "lavender", color = "black") +
  geom_density(color = "red", linewidth = 0.8) +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")
```

```{r}
#| label: Boxplot SLA


ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  labs(x = "Age class", y = "SLA")
```

```{r}
#| label: violin plot SLA

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE) +
  theme_minimal() +
  labs(x = "Age class", y = "SLA")
```

```{r}
#| label: Interactive plot SLA


ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NiN_type, color = calluna_NiN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN)  +
  labs(
    x = "Age class",
    y = "SLA",
    color = "Calluna NiN type"
  )

#swapped axis, NiN type on x-axis

ggplot(df_sum_SLA, aes(x = calluna_NiN_type, y = mean_SLA, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
   theme (
    text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) + 
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = AGE)  +
  labs(
    x = "Calluna NiN type",
    y = "SLA",
    color = "Age class"
  )
```

```{r}
#| label: Density Plot SLA

ggplot(cleaned_data, aes(x = SLA, y=calluna_NiN_type, fill=calluna_NiN_type)) +
  geom_density_ridges(alpha = 0.5) +
  facet_wrap(~age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  labs(title = "Density plot SLA", x = "SLA") +
  theme_minimal()






```

# Testing for normality

The dependent variable should be approximately normally distributed within each group. While ANOVA is relatively robust to moderate violations of normality, especially with larger sample sizes, extreme departures can affect validity.

-Shapiro-Wilk test reveals non-normality for the linear model, but since we have a large data set n\>50, the test is too sensitive and results might be a false negative, leading to rejection of normal data, when it is in reality fine to use.

-Both ANOVA and mixed models are robust to large datasets when it comes to deviating slightly from normality

-Plot for height is slightly heavy tailed and deviates slightly from the qqline, indicating data contains a few more extreme values than would be expected from a normal distribution. Scale plot also shows variance increases with fitted values

```{r}
#| label: testing age class and NiN type for normality with shapiro-wilk, qq plots, and residuals

#Find if data meets assumptions for a two-way ANOVA and if transformations are needed by using shapiro-wilk test and qq plot



########################################### Wet mass
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(wet_mass) %>% filter(p<0.05)

#sort by p-values above 0.05, which indicates normality. Thought it would be eaier to filter out which relationships can be used for the ANOVA, or maybe which we should transform before ANOVA?



# QQ plot is generally more informative than the Shapiro test because large datasets often make Shapiro significant even when deviations are minor.

#Combined qqplot for age class*NiN type, for every combination type, better overview

ggplot(cleaned_data, aes(sample = wet_mass)) +
 stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Wet mass qq plot")


#Linear model with interaction of age class and NiN type
#lm_wet_mass <- lm(wet_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
#summary(lm_wet_mass)



#Checking residuals 
#W = 0.98438, p-value = 0.001733, indicates non-normality
#shapiro.test(residuals(lm_wet_mass))

#qqplot
#plot(lm_wet_mass, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_wet_mass, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_wet_mass, which=1)






######################################### Dry mass
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(dry_mass) %>% filter(p>0.05)

#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = dry_mass)) +
stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Dry mass qq plot")



#Linear model with interaction of age class and NiN type
#lm_dry_mass <- lm(dry_mass ~ age_class*calluna_NiN_type, data = cleaned_data)
#summary(lm_wet_mass)


#Checking residuals
# W = 0.97974, p-value = 0.0001981 indicates data is not normally distributed, even if it appears so
#shapiro.test(residuals(lm_dry_mass))

#qqplot
#plot(lm_dry_mass, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_dry_mass, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_dry_mass, which=1)


########################################### Leaf area
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_area) %>% filter(p>0.05)



#Combined qq plot for age class*NiN type, every combination type for data before model

ggplot(cleaned_data, aes(sample = leaf_area)) +
 stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf area qq plot")

#Linear model leaf area with interaction of age class and NiN type

#lm_leaf_area <- lm(leaf_area ~ age_class*calluna_NiN_type, data = cleaned_data)
#summary(lm_leaf_area)



#Checking residuals
#W = 0.98409, p-value = 0.001577, indicates non-normality
#shapiro.test(residuals(lm_leaf_area))

#qqplot
#plot(lm_leaf_area, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_leaf_area, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_leaf_area, which=1)




##################################### Leaf thickness
#Shapiro test for data, lm test more relevant: data_thickness %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(leaf_thickness) %>% filter(p>0.05)




#Combined plot for age class*NiN type, for every combination type

ggplot(data_thickness, aes(sample = leaf_thickness)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Leaf thickness qq plot")



#Linear model with interaction of age class and NiN type
#lm_leaf_thickness <- lm(leaf_thickness ~ age_class*calluna_NiN_type, data = data_thickness)
#summary(lm_leaf_thickness)


#checking residuals
#W = 0.96656, p-value = 6.664e-14, not normally distributed
#shapiro.test(residuals(lm_leaf_thickness))

#qqplot
#plot(lm_leaf_thickness, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_leaf_thickness, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_leaf_thickness, which=1)





###################################### SLA
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(SLA) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = SLA)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "SLA qq plot")


#Linear model with interaction of age class and NiN type
#lm_SLA <- lm(SLA ~ age_class*calluna_NiN_type, data = cleaned_data)
#summary(lm_SLA)


#checking residuals
#W = 0.96722, p-value = 1.639e-06, not normally distributed
#shapiro.test(residuals(lm_SLA))

#qqplot
#plot(lm_SLA, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_SLA, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_SLA, which=1)






#################################### LDMC
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(LDMC) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = LDMC)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "LDMC qq plot")


#Linear model with interaction of age class and NiN type
#lm_LDMC <- lm(LDMC ~ age_class*calluna_NiN_type, data = cleaned_data)
#summary(lm_LDMC)



#Checking residuals
#W = 0.98146, p-value = 0.0004388, not normally distributed
#shapiro.test(residuals(lm_LDMC))

#qqplot
#plot(lm_LDMC, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
#plot(lm_LDMC, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
#plot(lm_LDMC, which=1)

```

Height should be fitted to a linear model as we have the same values per plant which includes many of the samples

```{r}
#| label: Height linear model and testing for normality
#Shapiro test for data, lm test more relevant: cleaned_data %>% group_by(calluna_NiN_type, age_class) %>% shapiro_test(height_cm) %>% filter(p>0.05)


#Combined plot for age class*NiN type, for every combination type

ggplot(cleaned_data, aes(sample = height_cm)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  facet_grid(age_class ~ calluna_NiN_type) +
  labs(title = "Height qq plot")



#Linear model with interaction of age class and NiN type
lm_height_cm <- lm(height_cm ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(lm_height_cm)




#Checking residuals
#W = 0.98229, p-value = 0.0006273, indicates non-normality
shapiro.test(residuals(lm_height_cm))

#qqplot is slightly heavy tailed, indicating data contains a few more extreme values than would be expected from a normal distribution
plot(lm_height_cm, which=2)
#Scale plot checking homoscedasticity of variance, constant variance, points should spread out equally around x-axis and line relatively straight
plot(lm_height_cm, which = 3, add.smooth = TRUE)
#Check residuals vs fitted, linearity of model
plot(lm_height_cm, which=1)

check_model(lm_height_cm)

```

# Levenetest for checking variance -not so relevant

Assumption for ANOVA is homogeneity of variance, so it is roughly equal cross all groups. Ensures that the pooled variance estimate used in ANOVA calculations is appropriate.

```{r}
#| label: levene test to check for homogeneity of variance as ANOVA requires roughly equal variance across all groups

#wet mass, result significant, variance is not equal


leveneTest(
  wet_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Dry mass, result non-significant, variance is equal


leveneTest(
  dry_mass ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Height, results significant, variance is not equal

leveneTest(
  height_cm ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf area, results significant, variance is not equal

leveneTest(
  leaf_area ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#Leaf thickness, results significant, variance is not equal

leveneTest(
  leaf_thickness ~ interaction(age_class, calluna_NiN_type),
  data = data_thickness
)

#LDMC, results not significant, variance is equal

leveneTest(
  LDMC ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)

#SLA, results significant, variance is not equal

leveneTest(
  SLA ~ interaction(age_class, calluna_NiN_type),
  data = cleaned_data
)
```

# Account for random effects and data clustering with lmer

→ Fixed effects with interactions: trait \~  NIN\*age class, effect of NiN type on trait also depends on effect of age class, and possibly plant sampling groups

-Plant ID (plant nr) is a grouping factor that should be used as random effects to account for non-independence. To account for “within-individual” grouping. These are nested groups with lower level units contained within higher levels

→ often use mixed-effects models as factors with few levels are fixed (NiN, age class) and factors with many levels are random (Plant 1-7 grouping)

5 leaf bulks ⊂ 5 samples ⊂ Plant 1-7 ⊂ Age class × NiN type

\
-Linear mixed effects models, lme4 package, diff model, but same anova, summary etc. 1I plantID, not for height due to low replicate number

-When you move to `lme4` for mixed-effects models, you have two types of normality to check:

- **Residual normality:** Just like ANOVA, the overall errors of the model should look fairly normal on a Q-Q plot.

- **Random effects normality:** The random intercepts or slopes you estimate for your groups (like subjects or schools) should also follow a normal distribution.

```{r}
#| label: Mixed model wet mass

#data=cleaned_data
#dependent variable = plant trait
#fixed factors: NiN and age class
#Grouping factor: Plant_nr nested in age class


lme4_wet_mass <- lmerTest::lmer(wet_mass ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_wet_mass)
plot(lme4_wet_mass, which=2)

qq_wet_mass_lme4 <- qqnorm(residuals(lme4_wet_mass), main = "Q-Q Plot of lme Residuals of Wet mass")
qqline(residuals(lme4_wet_mass), col = "red")

check_model(lme4_wet_mass)

#Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean


```

```{r}
#| label: Mixed model dry mass

lme4_dry_mass <- lmerTest::lmer(dry_mass ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_dry_mass)
plot(lme4_dry_mass)

qq_dry_mass_lme4 <- qqnorm(residuals(lme4_dry_mass), main = "Q-Q Plot of Residuals from lme of Dry mass")
qqline(residuals(lme4_dry_mass), col = "red")

check_model(lme4_dry_mass)
#Scale plot suggests mild heteroscedasticity with variance increasing slighlty with the mean




```

```{r}
#| label: Mixed model leaf area

lme4_leaf_area <- lmerTest::lmer(leaf_area ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data)

update(lme4_dry_mass, formula = leaf_area ~ .)

summary(lme4_leaf_area)  
plot(lme4_leaf_area)

qq_leaf_area_lme4 <- qqnorm(residuals(lme4_leaf_area), main = "Q-Q Plot of Residuals from lme of leaf area")
qqline(residuals(lme4_leaf_area), col = "red")


check_model(lme4_leaf_area)
```

```{r}
#| label: Mixed model leaf thickness

lme4_leaf_thickness <- lmerTest::lmer(leaf_thickness_mean ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_leaf_thickness) 
plot(lme4_leaf_thickness)

qq_leaf_thickness_lme4 <- qqnorm(residuals(lme4_leaf_thickness), main = "Q-Q Plot of Residuals from lme of leaf thickness")
qqline(residuals(lme4_leaf_thickness), col = "red")

check_model(lme4_leaf_thickness)

#Scale plot sorted by NiN type and age class, to see in which group residuals have greater variation

data_pred_thickness <- data.frame(
  predicted_leaf_thickness = predict(lme4_leaf_thickness),
  residual_leaf_thickness  = residuals(lme4_leaf_thickness),
  age_class     = cleaned_data$age_class,
  plant_nr      = cleaned_data$plant_nr,
  NiN_type      = cleaned_data$calluna_NiN_type
)

data_pred_thickness$Plant_ID <- interaction(
  data_pred_thickness$age_class,
  data_pred_thickness$plant_nr,
  data_pred_thickness$NiN_type,
  sep = "_"
)

ggplot(data_pred_thickness,
       aes(x = predicted_leaf_thickness,
           y = residual_leaf_thickness,
           colour = NiN_type)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 3)+
  facet_wrap(~age_class)
```

```{r}
#| label: Mixed model LDMC

lme4_LDMC <- lmerTest::lmer(LDMC ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_LDMC)  
plot(lme4_LDMC)

qq_LDMC_lme4 <- qqnorm(residuals(lme4_LDMC), main = "Q-Q Plot of Residuals from lme of LDMC")
qqline(residuals(lme4_LDMC), col = "red")

check_model(lme4_LDMC)
```

```{r}
#| label: Mixed model SLA

lme4_SLA <- lmerTest::lmer(SLA~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

summary(lme4_SLA) 

#Scale plot to visually assess model
plot(lme4_SLA)

#Scale plot suggests mild heteroscedasticity with slight fanning effect increasing with mean


#qq plot

qq_SLA_lme4 <- qqnorm(residuals(lme4_SLA), main = "Q-Q Plot of Residuals from lme of SLA")
qqline(residuals(lme4_SLA), col = "red")
check_model(lme4_SLA)




```

# lm ANOVA and Post-hoc -only for height

Null hypothesis: there are no significant differences between the interaction of age class and NiN type on given trait

p-value should be lower than 0.05 to reject the H0

Post hoc test with more than two levels, emmeans package: Everything the **emmeans** package does is an interpretation of the model that you fitted to the data, the model has to be a good fit before proceeding with this step.

Interpret emmeans: emmean column means the adjusted average for each trait in the same units

t-ratio tells you how many Standard Errors (SE) your group ratio is away from the null hypothesis, **Significant:** A t-ratio **greater than 2** or **less than -2** is generally statistically significant (assuming a standard 5% error rate).

**Not Significant:** A t-ratio close to 0 (between -2 and 2) means the groups are too similar to be sure of any real difference

estimate column: means the average difference in i.e for height in cm between groups

```{r}
#| label: lm ANOVA

#Wet mass

#```{r, results='asis'}
#cat("######## Wet Mass (g)\n\n")

#anova(lm_wet_mass)

#(emmeans_wet_mass<- emmeans(lm_wet_mass, ~ age_class * calluna_NiN_type))
#pairs(emmeans_wet_mass)

#plot(emmeans_wet_mass)

#Dry mass

#```{r, results='asis'}
#cat("######## Dry Mass (g)\n\n")

#anova(lm_dry_mass)

#(emmeans_dry_mass<- emmeans(lm_dry_mass, ~ age_class * calluna_NiN_type))
#pairs(emmeans_dry_mass)



#Height

cat("######## Height (cm)\n\n")

anova(lm_height_cm)

(emmeans_height_cm<- emmeans(lm_height_cm, ~ age_class * calluna_NiN_type))
pairs(emmeans_height_cm, simple="each", adjust = "tukey") #Tukey adjustment to prevent false positives

#Grouping individual means by letters for visualization
cld(emmeans_height_cm, Letters=letters)




#Leaf area 

#```{r, results='asis'}
#cat("######## Leaf area (mm^2)\n\n")

#anova(lm_leaf_area)

#(emmeans_leaf_area<- emmeans(lm_leaf_area, ~ age_class * calluna_NiN_type))
#pairs(emmeans_leaf_area)

#Leaf thickness

#```{r, results='asis'}
#cat("######## Leaf thickness (mm)\n\n")

#anova(lm_leaf_thickness)

#(emmeans_leaf_thickness<- emmeans(lm_leaf_thickness, ~ age_class * calluna_NiN_type))
#pairs(emmeans_leaf_thickness)

#LDMC

#```{r, results='asis'}
#cat("######## LDMC\n\n")

#anova(lm_LDMC)

#(emmeans_LDMC<- emmeans(lm_LDMC, ~ age_class * calluna_NiN_type))
#pairs(emmeans_LDMC) 



#SLA

#```{r, results='asis'}
#cat("######## SLA\n\n")

#anova(lm_SLA)

#(emmeans_SLA<- emmeans(lm_SLA, ~ age_class * calluna_NiN_type))
#pairs(emmeans_SLA) 
```

# lmer ANOVA and post-hoc testing

TO DO: find out if this is the proper format for testing with lme4 model, anova does not give a Pr value as the degrees of freedom in a mixed model are. ot accurate, so the p-value cannot be precise enough

```{r}
#| label: ANOVA and emmeans wet mass

#Wet mass

cat("######## Wet Mass (g)\n\n")

anova(lme4_wet_mass)

(emmeans_lme4_wet_mass<- emmeans(lme4_wet_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_wet_mass, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_wet_mass, Letters=letters)


```

```{r}
#| label: ANOVA and emmeans dry mass

cat("######## Dry Mass (g)\n\n")

anova(lme4_dry_mass)

(emmeans_lme4_dry_mass<- emmeans(lme4_dry_mass, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_dry_mass, simple="each", adjust = "tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_dry_mass, Letters=letters) #Testing cld, but have later decided to omit these results from the analysis as they don't show all the significant effects from the pairwise emmeans post hoc
```

```{r}
#| label: ANOVA and emmeans leaf area

cat("######## Leaf area (mm^2)\n\n")

anova(lme4_leaf_area)

(emmeans_lme4_leaf_area<- emmeans(lme4_leaf_area, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_leaf_area, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_leaf_area, Letters=letters)
```

```{r}
#| label: ANOVA and emmeans leaf thickness

cat("######## Leaf thickness (mm)\n\n")

anova(lme4_leaf_thickness)

(emmeans_lme4_leaf_thickness<- emmeans(lme4_leaf_thickness, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_leaf_thickness, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_leaf_thickness, Letters=letters)
```

```{r}
#| label: ANOVA and emmeans LDMC

cat("######## LDMC\n\n")

anova(lme4_LDMC)

(emmeans_lme4_LDMC<- emmeans(lme4_LDMC, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_LDMC, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_LDMC, Letters=letters)

#Plots to visualize emmeans output
emmip(emmeans_lme4_LDMC, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_LDMC, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other
```

```{r}
#| label: ANOVA and emmeans SLA

cat("######## SLA\n\n")

anova(lme4_SLA)

(emmeans_lme4_SLA<- emmeans(lme4_SLA, ~ age_class * calluna_NiN_type))
pairs(emmeans_lme4_SLA, simple="each", adjust="tukey") #Tukey adjustment to prevent false positives
cld(emmeans_lme4_SLA, Letters=letters)

```

# Data transformation?

**-data often natural log transformed and scaled** to fairly compare across traits, we can statistically contrast the different traits side-by-side: makes skewed distributions more normal by pulling extreme outliers inward; allows interpretation in multiplicative steps (1x increase → 10x increase etc.); prevents heteroscedasticity (data artificially widening as mean increases); when comparing multiple traits, scaling makes sure a change of one unit in one trait means the same across all variables

-Never use *p*-values to judge the success of a transformation! We use diagnostic plots to make that assessment. What we hope is that we can transform the response variable so that it conforms, at least approximately, to the assumptions of the statistical model

-transform traits individually and do back-transformation

-Plot for height is slightly heavy tailed and deviates slightly from the qqline, indicating data contains a few more extreme values than would be expected from a normal distribution. Scale plot also shows variance increases with fitted values —\> try transformation

```{r}
#| label: data transformation


height_log_mod <- lm(log(height_cm) ~ age_class*calluna_NiN_type, data = cleaned_data)
summary(height_log_mod)

plot(lm_height_cm, which = 2)
plot(lm_height_cm, which = 3)
plot(height_log_mod, which = 2) #new probability plot looks a little better, but not perfect, tails are outside
plot(height_log_mod, which = 3) #Constant variation assumption fulfilled

check_model(lm_height_cm)
check_model(height_log_mod)

anova(height_log_mod)
emmeans_height_log_mod <- emmeans(height_log_mod, ~ age_class * calluna_NiN_type)
pairs(emmeans_height_log_mod, , type="response") #Add type="response" for contrasting ratios, one group is i.e 0.830 bigger than the other, a ratio of 1 would mean the groups are identical

#the data looks more right-skewed after log transformation
#Does not look like the transformation did much for the siginificance of emmeans results or the variation of the residuals, decision to stay with original height data
```

```{r}
#| label: Should leaf area be transformed? The data is slightly skewed, seen in the histograms


#Original mixed model for comparison
check_model(lme4_leaf_area)
plot(lme4_leaf_area)
qq_leaf_area_lme4 <- qqnorm(residuals(lme4_leaf_area), main = "Q-Q Plot of Residuals from lme of leaf area")
qqline(residuals(lme4_leaf_area), col = "red")

#Log transformed
log_leaf_area <- lmerTest::lmer(log(leaf_area) ~ age_class*calluna_NiN_type + (1|plant_id), data=cleaned_data, na.action = na.exclude)

plot(log_leaf_area)
qq_leaf_area_log <- qqnorm(residuals(log_leaf_area), main = "Q-Q Plot of Residuals from log transformed leaf area")
qqline(residuals(log_leaf_area), col = "red")
check_model(log_leaf_area)


#Residuals are slightly better fitted to line when log transformed, fanning out effect is reduced (less fanning out of the spread as predicted values increase, variance is more constant with log transformation), but the effect 
#QQplots show quite similar fit along the qqline, decision to stay with non-log transformed data

```

# Multivariate analysis and PCA

-Because traits are not independent, we should investigate interactions between them

-The primary goal of MANOVA is to determine whether the means of the dependent variables differ significantly across groups while considering the interrelationships between the variables

-MANOVA tests whether groups differ in the **multivariate combination** of traits rather than running many separate ANOVAs, it tests if overall trait syndrome differs among age classes, NiN types and the interaction

-Use PCA to reduce dimensionality and identify major axes of trait covariation across NiN types and age classes (remember Diaz et al. 2016 method)

-Because leaf traits often covary (e.g., LES), check for high multicollinearity. Highly correlated variables ((r \> 0.85) or (0.90)) may skew PCA loadings toward specific traits –\> i.e remove wet mass, dry mass

```{r}
#| label: Testing correlations between traits

#Look at trait pairs scatterplot to evaluate if pearson is more appropriate than spearman, relationships are widely scattered, but more linear so pearson is better

traits_compiled <- cleaned_data[, c(
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC"
)]

traits_compiled <- na.omit(traits_compiled)


pairs(traits_compiled)

corr <- cor(cleaned_data[, c(
  "dry_mass",
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC"
)], method = "pearson",
use = "pairwise.complete.obs")

round(corr, 2)
summary(corr)

#Testing correlations between traits as SLA and LDMC are derived from variables already in the model ---appears only wet and dry mass are highly correlated --decision to omit dry mass from MANOVA and pca as they are intermediate measurements use to calculate LDMC and SLA, which are common functional traits
#NAs are omitted, wet and dry mass excluded




```

```{r}
#| label: scaling trait values for glm model


cleaned_data <- cleaned_data %>%
  mutate(across(c(SLA, LDMC, leaf_area, leaf_thickness_mean, height_cm),
                ~ as.numeric(scale(.)),
                .names = "{.col}_scaled"))

#Test if scaling successful
#vector of column names

cols <- c("SLA_scaled", "LDMC_scaled", "leaf_area_scaled",
          "leaf_thickness_mean_scaled", "height_cm_scaled")

#should be close to 0, which it is here
cleaned_data %>%
  summarise(across(all_of(cols), mean, na.rm = TRUE))

#should be close to 1, which it is here
cleaned_data %>%
  summarise(across(all_of(cols), sd, na.rm = TRUE))

```

```{r}
#| label: Trait correlation network and MANOVA



#Correlation plot other layout


colnames(corr) <- c(
  "Dry mass",
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC"
)

rownames(corr) <- c(
  "Dry mass",
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC"
)


corrplot_export <- corrplot(corr,
         method = "circle",
         type = "lower",
         order = "hclust",
         diag = FALSE,
         col = colorRampPalette(c("#B2182B", "white", "#2166AC"))(200),
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.9,
         cl.cex = 0.8,
         addCoef.col = "black",
         number.cex = 0.7,
         mar = c(0,0,1,0)) +
  scale_size_area(max_size = 15) +
  geom_text(aes(label = sprintf("%.2f", corr)),
            size = 3,
            colour = "black") +
  mtext("Pearson correlation (r)",
        side = 4,
        line = 3,
        cex = 0.9)


#Manova test

manova_mod <- manova(
  cbind(
    height_cm,
    leaf_area,
    leaf_thickness_mean,
    SLA,
    LDMC
  ) ~ age_class * calluna_NIN_type,
  data = cleaned_data
)

summary(manova_mod)
summary.aov(manova_mod)
summary(manova_mod, test = "Pillai")

#Emmeans post-hos test for Manova??
```

```{r}
#| label: Dataframes for PCA


traits_NiN_age <- cleaned_data[, c(
  "height_cm",
  "leaf_area",
  "leaf_thickness_mean",
  "SLA",
  "LDMC",
  "age_class",
  "calluna_NiN_type"
)]

colnames(traits_NiN_age) <- c(
  "Height",
  "Leaf area",
  "Leaf thickness",
  "SLA",
  "LDMC",
  "age_class",
  "calluna_NiN_type"
)



traits_NiN_age <- na.omit(traits_NiN_age)
```

```{r}
#| label: PCA by prcomp

pca_prcomp <- prcomp(traits_compiled, scale. = TRUE)
summary(pca_prcomp)

scores_prcomp <- cbind(
 traits_NiN_age,
 PC1_prcomp = pca_prcomp$x[,1],
 PC2_prcomp = pca_prcomp$x[,2]
 )

#Calculate % for the plot axis
 pca_importance_prcomp <- summary(pca_prcomp)$importance
 pc1_percent_prcomp <- round(pca_importance_prcomp[2, 1] * 100, 1)
 pc2_percent_prcomp <- round(pca_importance_prcomp[2, 2] * 100, 1)

```

```{r}
#| label: PCA by vegan package

#improve visuals

custom_labels <- data.frame(
  Labels_traits = c("Leaf area", "SLA", "Vegetative height", "Leaf thickness", "LDMC"),
  x = c(1.5,   3.6,  -2.9,  -2.4,  -1.5),  # ◄ Desired X positions on the plot
  y = c(2.8,  -0.3,  -1.4,   2.1,   1.5)   # ◄ Desired Y positions on the plot
)


# 1. Run the PCA using rda()
# scale = TRUE standardizes the traits (equivalent to scale. = TRUE in prcomp)
pca_vegan <- rda(traits_compiled, scale = TRUE)
summary(pca_vegan)


# 3. Get the trait loadings (equivalent to pca$rotation)
# In vegan, these are called 'species' scores
round(scores(pca_vegan, display = "species", choices = 1:2, scaling = 0), 3)

# 4. Extract site scores and combine with your metadata (equivalent to pca$x)
pca_scores <- scores(pca_vegan, display = "sites", choices = 1:2)

scores <- cbind(
  traits_NiN_age,
  PC1 = pca_scores[, "PC1"],
  PC2 = pca_scores[, "PC2"]
)

# Trait loadings (variable vectors)
loadings <- as.data.frame(
  scores(
    pca_vegan,
    display = "species",
    choices = 1:2,
    scaling = 0
  )
)

loadings$Variable <- rownames(loadings)
loadings

# 5. Calculate % variation for the plot axes
# eigenvals() extracts the eigenvalues from the vegan model
eigenvalues <- summary(eigenvals(pca_vegan))
pc1_percent <- round(eigenvalues["Proportion Explained", "PC1"] * 100, 1)
pc2_percent <- round(eigenvalues["Proportion Explained", "PC2"] * 100, 1)
```

```{r}
#| label: PCA_vegan visualize results


#Visualize PCA_vegan results
#Can already see from PC1 that mature and pioneer separate in multivariate trait space, don't overlap much, whereas building is intermediate and overlaps with both

ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = age_class,
    shape = calluna_NiN_type
  )
) +
  scale_color_manual(values = AGE) +
  geom_point(size = 3) +
  stat_ellipse()+
  labs(
    x = paste0("PC1 (", pc1_percent, "%)"), 
    y = paste0("PC2 (", pc2_percent, "%)")  
  ) 






####Other plot

site_scores <- as.data.frame(scores(pca_vegan, display = "sites"))
species_scores <- as.data.frame(scores(pca_vegan, display = "species"))
species_scores$Species <- rownames(species_scores)

# Calculate explained variance for axis labels
scaling_info <- summary(pca_vegan)$cont$importance
pc1_var <- round(scaling_info[2, 1] * 100, 1)
pc2_var <- round(scaling_info[2, 2] * 100, 1)

ggplot() +
  # Draw species vectors (arrows)
  geom_segment(data = species_scores,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "darkred", alpha = 0.7) +
  # Smart species labels that don't overlap
  geom_text_repel(data = species_scores, 
                  aes(x = PC1, y = PC2, label = Species),
                  color = "darkred", size = 3, fontface = "italic") +
  # Draw site points
  geom_point(data = site_scores, 
             aes(x = PC1, y = PC2), 
             color = "royalblue", size = 2.5, alpha = 0.8) +
  # Dynamic axis titles with % variance explained
  labs(x = paste0("PC1 (", pc1_var, "%)"), 
       y = paste0("PC2 (", pc2_var, "%)"),
       title = "Principal Component Analysis (PCA)") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  coord_equal()


```

```{r}
#| label: PCA-plot-coloured-by-age-class

library(grid)


## Scale arrows to match score space


arrow_scale <- min(
  diff(range(scores$PC1))/diff(range(loadings$PC1)),
  diff(range(scores$PC2))/diff(range(loadings$PC2))
) * 0.75

loadings <- loadings %>%
  mutate(
    PC1 = PC1 * arrow_scale,
    PC2 = PC2 * arrow_scale
  )

#Labels traits into loading dataframe


new_labels <- c(
  height_cm = "Height",
  leaf_area = "Area",
  leaf_thickness_mean = "Thickness",
  SLA = "SLA",
  LDMC = "LDMC"
)

loadings$Variable <- ifelse(
  loadings$Variable %in% names(new_labels),
  new_labels[loadings$Variable],
  loadings$Variable
)


#Offset arrow labels
loadings <- loadings %>%
  mutate(
    label_x = PC1 * 1.08,
    label_y = PC2 * 1.2
  )

#Add interaction titles to scores dataframe
scores$group <- interaction(scores$age_class,
                            scores$calluna_NiN_type, sep = " × ")

pca_age_color <- ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = age_class,
    shape = calluna_NiN_type
  )
) +

geom_hline(yintercept = 0,
           colour = "grey80",
           linetype = 2) +

geom_vline(xintercept = 0,
           colour = "grey80",
           linetype = 2) +

stat_ellipse(
  aes(group = age_class,
      colour = age_class),
  level = 0.68,
  linewidth = 0.7,
  show.legend = FALSE,
  alpha = 0.5
) +

geom_point(
  size = 2.2,
  alpha = 0.65
) +

geom_segment(
  data = loadings,
  aes(
    x = 0,
    y = 0,
    xend = PC1,
    yend = PC2
  ),
  inherit.aes = FALSE,
  colour = "grey25",
  linewidth = 0.5,
  arrow = arrow(length = unit(0.25,"cm"))
) +

geom_text_repel(
  data = loadings,
  aes(
    x = label_x,
    y = label_y,
    label = Variable
  ),
  inherit.aes = FALSE,
  colour = "grey20",
  size = 4,
  point.padding = 0.3,
  box.padding = 0.5,
  max.overlaps = Inf, 
  segment.color = "grey80"
) +

scale_colour_manual(values = AGE) +

scale_shape_manual(values = c(
  BH = 16,
  FH = 17,
  TH = 15
)) +

coord_equal() +

labs(
  title = "PCA - Plant Traits",
  x = paste0("PC1 (", pc1_percent, "%)"),
  y = paste0("PC2 (", pc2_percent, "%)"),
  colour = "Age class",
  shape = "NiN type"
) +

theme_classic(base_size = 13) +

theme(
  plot.title = element_text(
    face = "plain",
    hjust = 0.5,
    size = 18
  ),

  axis.title = element_text(
    face = "plain",
    size = 14
  ),

  axis.text = element_text(size = 12),

  legend.title = element_text(
    face = "plain",
    size = 13
  ),

  legend.text = element_text(size = 12),

  legend.position = "right"
) +
  facet_wrap(~calluna_NiN_type)

pca_age_color

#Saving output

ggsave(
 "PCA_plot_age.png",
 width=10,
 height=8,
 dpi=300
)

```

```{r}
#| label: PCA-plot-coloured-by-NiN-type


#Offset arrow labels
loadings <- loadings %>%
  mutate(
    label_x2 = PC1 * 1.09,
    label_y2 = PC2 * 1.085
  )

pca_NiN_color <- ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = calluna_NiN_type,
    fill = calluna_NiN_type
  )
) +

geom_hline(yintercept = 0,
           colour = "grey80",
           linetype = 2) +

geom_vline(xintercept = 0,
           colour = "grey80",
           linetype = 2
) +

  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_point(
    size = 2.2,
    alpha = 0.5
  ) +

geom_segment(
  data = loadings,
  aes(
    x = 0,
    y = 0,
    xend = PC1,
    yend = PC2
  ),
  inherit.aes = FALSE,
  colour = "grey25",
  linewidth = 0.5,
  arrow = arrow(length = unit(0.25,"cm"))
) +

geom_text_repel(
  data = loadings,
  aes(
    x = label_x2,
    y = label_y2,
    label = Variable
  ),
  inherit.aes = FALSE,
  colour = "grey20",
  size = 4,
  point.padding = 0.5,
  box.padding = 0.35,
  max.overlaps = Inf, 
  segment.color = NA
) +

scale_colour_manual(values = NIN) +
  scale_fill_manual(values = NIN) +

coord_equal() +

labs(
  title = "PCA - Plant Traits",
  x = paste0("PC1 (", pc1_percent, "%)"),
  y = paste0("PC2 (", pc2_percent, "%)"),
  colour = "NiN type",
  shape = "Age class"
) +

theme_classic(base_size = 13) +

theme(
  plot.title = element_text(
    face = "plain",
    hjust = 0.5,
    size = 18
  ),

  axis.title = element_text(
    face = "plain",
    size = 14
  ),

  axis.text = element_text(size = 12),

  legend.title = element_text(
    face = "plain",
    size = 13
  ),

  legend.text = element_text(size = 12),

  legend.position = "right"
) +
  guides(fill = "none") +
  facet_wrap(~age_class)

pca_NiN_color

#Saving output

pca_NiN_color <- pca_NiN_color +
  theme(plot.title = element_blank())

ggsave(
 "pca_NiN_color.png",
 width=16,
 height=12,
 dpi=300
)

```

```{r}
#| label: betadisper-anova-permutest-TukeyHSD

#Test if FH has a stronger multivariate separation that BH and TH 
#The FH group exhibited the largest confidence ellipse, suggesting greater multivariate dispersion in functional trait composition than BH and TH, whereas TH showed the most compact distribution

dist_mat <- dist(scale(traits_compiled))

bd_NiN <- betadisper(dist_mat, traits_NiN_age$calluna_NiN_type) 

anova(bd_NiN)
permutest(bd_NiN)
TukeyHSD(bd_NiN) #Results significant for BH-FH and TH-FH, but not for TH-BH, indicate FH differed significantly in dispersion from the other two NiN types
plot(bd_NiN, kind = "boxplot", xlab = "NiN type", ylab = "Distance to centroid")




#Hypothesis: trait variance may decrease with age if mature plants converge on a conservative strategy -test with betadisper
bd_age <- betadisper(dist_mat, traits_NiN_age$age_class)

anova(bd_age)
permutest(bd_age, pairwise = TRUE, permutations = 999) #Give unadjusted p-values
TukeyHSD(bd_age) #Adjusts for multiple comparisons

# Boxplot from betadisper object
plot(bd_age, kind = "boxplot", xlab = "Age class", ylab = "Distance to centroid")

#Result meets assumption of homogeneity, but shows no significant difference in the multivariate dispersion in functional trait composition across age class. Hypothesis disproved: trait variance does not significantly decrease with age. If PERMANOVA is run, ssumptions of homogenous dispersions are met

```

```{r}
#| label: PERMANOVA-test-pairwiseAdonis
#Use euclidean as functional traits are continuous hysical measurements
#The age class PERMANOVA tells you if traits vary across ages, completely ignoring NiN types.
#The NiN type PERMANOVA tells you if traits vary across habitat types, completely ignoring age.

permanova_age_class <- adonis2(dist_mat ~ age_class, 
                            data = traits_NiN_age, 
                            method = "euclidean",     
                            permutations = 999)

print(permanova_age_class)

#Show significant results, but not the direction of the shift. It means the overall phenotype (the combination of all 5 functional traits) of the species significantly shifts across age class. Need to investigate age class gradient. 

permanova_NiN <- adonis2(dist_mat ~ calluna_NiN_type, 
                            data = traits_NiN_age, 
                            method = "euclidean",     
                            permutations = 999)

print(permanova_NiN)


# Run a two-way PERMANOVA with an interaction term to include interaction of age class and NiN type
#age_class: Does trait variation change across ages overall?
#NiN_type: Does trait variation change across NiN types overall?
#age_class:NiN_type (The Interaction): Is the effect of age different depending on the NiN type?

permanova_age_NiN <- adonis2(dist_mat ~ age_class * calluna_NiN_type, 
                           data = traits_NiN_age, 
                           method = "euclidean", 
                           permutations = 999,
                           by = "margin")

print(permanova_age_NiN)

#Pairwise PERMANOVA post hoc-test to check if the centroids of the Pioneer and Mature ellipses are significantly different from each other

# Run pairwise comparisons on your original distance matrix for age class
pairwise_age <- pairwise.adonis2(dist_mat ~ age_class, data = traits_NiN_age)

print(pairwise_age)


# NiN type
pairwise_NiN <- pairwise.adonis2(dist_mat ~ calluna_NiN_type, data = traits_NiN_age)

print(pairwise_NiN)


#Double check PERMANOVA was carried out correctly, row names equal
all(rownames(cleaned_data) == rownames(dist_mat))

```

Got feedback that PERMANOVA and pairwise adonis might not be the best test to evaluate statistical differences in centroid means, therefore an RDA with ANOVA is carried out. For now, i've done the rda on the interaction between age class and NiN type to answer the research question , but could probably perform separate ones as well, as i've done with the PCA plots. The pca analysis does not include the interaction term, but the RDA does, so it is probably a more accurate analysis.

```{r}
#| label: RDA-analysis-anova


rda <- rda(traits_compiled ~ calluna_NiN_type * age_class, data = traits_NiN_age)

summary(rda)
plot(rda)

RsquareAdj(rda) #Total variance explained by rda

anova(rda, permutations = 999) #Nothing to do with ANOVA, pseudo F-value, but the closest you get to ANOVA with ordination

anova(rda, by ="axis") #Axis significance

anova(rda, by = "terms") #significance by age class and Nin

#Look at multicollinearity, >2 is considered high, variables are fine
sqrt(vif.cca(rda))

#Extract RDA scores

# Site scores
sites_rda <- as.data.frame(scores(rda, display = "sites"))
sites_rda$age_class <- traits_NiN_age$age_class
sites_rda$calluna_NiN_type <- traits_NiN_age$calluna_NiN_type

# Species scores
scores_rda <- cbind(
  traits_NiN_age,
  RDA1 = sites_rda[, "RDA1"],
  RDA2 = sites_rda[, "RDA2"]
)

# Trait loadings (variable vectors)
loadings_rda <- as.data.frame(
  scores(
    rda,
    display = "species",
    choices = 1:2,
    scaling = 0
  )
)

loadings_rda$Variable <- rownames(loadings_rda)

#Naming variable rows
new_labels <- c(
  "Vegetative height" = "Height",
  "Leaf area" = "Area",
  "Leaf thickness" = "Thickness",
  SLA = "SLA",
  LDMC = "LDMC"
)

loadings_rda$Variable <- ifelse(
  loadings_rda$Variable %in% names(new_labels),
  new_labels[loadings_rda$Variable],
  loadings_rda$Variable
)

loadings_rda

# Environmental arrows
env_arrows <- as.data.frame(scores(rda, display = "bp"))
env_arrows$variable <- rownames(env_arrows)

#Percent variance explained
eig <- eigenvals(rda)

pc1_rda <- round(100 * eig[1] / sum(eig), 2)
pc2_rda <- round(100 * eig[2] / sum(eig), 2)


#Offset arrow labels
loadings_rda <- loadings_rda %>%
  mutate(
    label_x3 = RDA1 * 1.25,
    label_y3 = RDA2 * 1.25)
```

```{r}
#| label: plot-rda-facet-by-age



 rda_plot <- ggplot(sites_rda, aes(RDA1, RDA2, colour = calluna_NiN_type, fill = calluna_NiN_type)) +
  
  geom_point(size = 1.5, alpha = 0.85) +
  
  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_segment(
    data = loadings_rda,
    aes(x = 0, y = 0,
        xend = RDA1*2,
        yend = RDA2*2),
    linewidth = 0.7,
    colour = "grey20",
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")
   )
  ) +
  geom_text_repel(
    data = loadings_rda,
    aes(
      x = label_x3,
      y = label_y3,
      label = Variable
    ),
    force = 12,
    inherit.aes = FALSE,
    colour = "grey20",
    size = 4,
    point.padding = 0.3,
    box.padding = 4,
    max.overlaps = Inf, 
    segment.color = "grey40",
    segment.size = 0.4,
    min.segment.length = 0
  ) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right",
    strip.text = element_text(face = "plain")
  ) +
    coord_equal() +
  geom_hline(yintercept = 0,
             linetype = 2,
             colour = "grey70") +

geom_vline(xintercept = 0,
           linetype = 2,
           colour = "grey70") +
    labs(
      title = "RDA - Plant Traits",
      x = paste0("RDA1 (", pc1_rda, "%)"),
      y = paste0("RDA2 (", pc2_rda, "%)"),
      colour = "NiN type"
    ) +
  guides(fill = "none") +
  facet_wrap(~age_class)
  
  rda_plot

#Remember to mention arrows are offset by 2 * multiplication for aesthetics!

#Saving output

ggsave(
  "rda_plot.png",
  width=12,
  height=10,
  dpi=300
)

```

```{r}
#| label: plot-rda-facet-by-age-ellipselines


ggplot(sites_rda, aes(RDA1, RDA2, colour = calluna_NiN_type, fill = calluna_NiN_type)) +
  
  geom_point(size = 1.5, alpha = 0.85) +
  
  stat_ellipse(
    aes(group = calluna_NiN_type,
        colour = calluna_NiN_type),
    alpha = 0.5,
    level = 0.95,
    linewidth = 0.7
  ) +
  
  geom_segment(
    data = loadings_rda,
    aes(x = 0, y = 0,
        xend = RDA1*1.5,
        yend = RDA2*1.5),
    linewidth = 0.7,
    colour = "grey20",
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")
    )
  ) +
  geom_text_repel(
    data = loadings_rda,
    aes(
      x = label_x3,
      y = label_y3,
      label = Variable
    ),
    force = 12,
    inherit.aes = FALSE,
    colour = "grey20",
    size = 4,
    point.padding = 0.3,
    box.padding = 4,
    max.overlaps = Inf, 
    segment.color = "grey40",
    segment.size = 0.4,
    min.segment.length = 0
  ) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right",
    strip.text = element_text(face = "plain")
  ) +
  coord_equal() +
  geom_hline(yintercept = 0,
             linetype = 2,
             colour = "grey70") +
  
  geom_vline(xintercept = 0,
             linetype = 2,
             colour = "grey70") +
  labs(
    title = "RDA - Plant Traits",
    x = paste0("RDA1 (", pc1_rda, "%)"),
    y = paste0("RDA2 (", pc2_rda, "%)"),
    colour = "NiN type"
  ) +
  guides(fill = "none") +
  facet_wrap(~age_class)



```

```{r}
#| label: plot-rda-facet-by-age-and-NiN

#Interaction column
sites_rda$facet_group <- interaction(
  sites_rda$calluna_NiN_type,
  sites_rda$age_class
)

ggplot(sites_rda, aes(RDA1, RDA2, colour = calluna_NiN_type, fill = calluna_NiN_type)) +
  
  geom_point(size = 1.5, alpha = 0.85) +
  
  stat_ellipse(
    geom = "polygon",
    alpha = 0.2,
    level = 0.95,
    colour = NA
  ) +
  
  geom_segment(
    data = loadings_rda,
    aes(x = 0, y = 0,
        xend = RDA1*2,
        yend = RDA2*2),
    linewidth = 0.7,
    colour = "grey20",
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")
    )
  ) +
  geom_text_repel(
    data = loadings_rda,
    aes(
      x = label_x3,
      y = label_y3,
      label = Variable
    ),
    force = 12,
    inherit.aes = FALSE,
    colour = "grey20",
    size = 4,
    point.padding = 0.3,
    box.padding = 4,
    max.overlaps = Inf, 
    segment.color = "grey40",
    segment.size = 0.4,
    min.segment.length = 0
  ) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right",
    strip.text = element_text(face = "plain")
  ) +
  coord_equal() +
  geom_hline(yintercept = 0,
             linetype = 2,
             colour = "grey70") +
  
  geom_vline(xintercept = 0,
             linetype = 2,
             colour = "grey70") +
  labs(
    title = "RDA - Plant Traits",
    x = paste0("RDA1 (", pc1_rda, "%)"),
    y = paste0("RDA2 (", pc2_rda, "%)"),
    colour = "NiN type"
  ) +
  guides(fill = "none") +
  facet_wrap(~facet_group)



```

There is quite a big overlap of NiN types, showing they have less of a distinct effect on the traits and age class slightly stringer effect, especially seen with FH pioneer Calluna

```{r}
#| label: PCA separated by facet wrap 



ggplot(
  scores,
  aes(
    PC1,
    PC2,
    colour = calluna_NiN_type
  )
) +
  scale_color_manual(values = NIN) +
  geom_point(size = 1.8, alpha = 0.9) +
  stat_ellipse()+
  facet_wrap(~ age_class, ncol = 3)+
labs(
  x = paste0("PC1 (", pc1_percent, "%)"), 
  y = paste0("PC2 (", pc2_percent, "%)")  
)
```

```{r}
#| label: pca plots for each individual trait

#To visualize and and explain how age class and NiN inyteract for each trait
#color-code your points by both factors or use facets. A great way to do this visually is to plot separate PCA panels for each NiN type, showing the successional trajectory (Pioneer → Building → Mature) within each habitat. If the trajectories point in different directions or have different lengths, that is your significant interaction in action

ggplot(traits_NiN_age, aes(x = age_class, y = `Leaf area`, color = calluna_NiN_type, group = calluna_NiN_type)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  theme_minimal() +
  scale_color_manual(values = NIN)
  labs(title = "Interaction Plot for leaf area", x = "Age Class", y = "Mean Trait Value")
  
  #This plot shows exactly what the individual plots by trait show further up in the script


```

```{r}
#| label: PCA biplot

#Use prcomp pca analysis for these plots as the factoextra is compatible with this format, but the results are mathematically identical for the vegan pca, they yield the exact same coordinates, percentages, and arrows


fviz_pca_biplot(
  pca_prcomp,
  habillage = traits_NiN_age$age_class,
  addEllipses = TRUE,
  label="var",
  repel=TRUE
)+
scale_color_manual(values = AGE) 


#By NiN type

fviz_pca_biplot(
  pca_prcomp,
  habillage = traits_NiN_age$calluna_NiN_type,
  addEllipses = TRUE,
  label="var",
  repel=TRUE,
  alpha = 0.35
)+
  scale_color_manual(values = NIN) +
  stat_ellipse(level = 0.68)
```

```{r}
#| label: PCA biplots with different visuals



biplot_NiN <- fviz_pca_biplot(
  pca_prcomp,
  habillage = traits_NiN_age$calluna_NiN_type,
  addEllipses = TRUE,
  ellipse.level = 0.68,
  ellipse.type = "norm", 
  label = "none",
  repel = TRUE,
  # ◄ This pushes the labels further outward away from the arrow tips/points
  ggrepel.options = list(point.padding = 0.8, box.padding = 0.5, max.overlaps = Inf),
  alpha.var = 0.35,
  geom.ind = "point",
  col.var = "gray20",
  labelsize = 4
) + geom_text(
  data = custom_labels, 
  aes(x = x, y = y, label = Variable), 
  color = "gray20", 
  size = 4,
  fontface = "plain",
  inherit.aes = FALSE   # Prevents it from trying to map to the groups
) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = rep("transparent", length(unique(traits_NiN_age$calluna_NiN_type)))) + 
  scale_shape_manual(values = c(16, 16, 16)) +
  labs(
    title = "PCA Biplot - Plant Traits",
    color = "NiN Type",
    x = "PC1 (40.1%)",
    y = "PC2 (23.2%)"
  ) +
  guides(fill = "none", shape = "none") 

biplot_NiN

#The ellipses overlap substantially, suggesting group differences exist but are not complete separation.
#FH nature type has the greatest spread of data, variation
#leaf_area and SLA point strongly to the right → associated with Pioneer samples
#height points left → associated with Mature samples
#LDMC and leaf_thickness_mean point upper-left and are positively correlated
#SLA and LDMC, leaf thickness point in roughly opposite directions which supports research on aqcuisative vs conservative growth strategies


biplot_age <- fviz_pca_biplot(
  pca_prcomp,
  habillage = traits_NiN_age$age_class,
  addEllipses = TRUE,
  ellipse.level = 0.68,
  ellipse.type = "norm", 
  label = "none",
  repel = TRUE,
  # ◄ This pushes the labels further outward away from the arrow tips/points
  ggrepel.options = list(point.padding = 0.8, box.padding = 0.5, max.overlaps = Inf),
  alpha.var = 0.35,
  geom.ind = "point",
  col.var = "gray20",
  labelsize = 4
) + geom_text(
  data = custom_labels, 
  aes(x = x, y = y, label = Variable), 
  color = "gray20", 
  size = 4,
  fontface = "plain",
  inherit.aes = FALSE   # Prevents it from trying to map to the groups
) +
  scale_color_manual(values = AGE) +
  scale_fill_manual(values = rep("transparent", length(unique(traits_NiN_age$age_class)))) + 
  scale_shape_manual(values = c(16, 16, 16)) +
  labs(
    title = "PCA Biplot - Plant Traits",
    color = "Age class",
    x = "PC1 (40.1%)",
    y = "PC2 (23.2%)"
  ) +
  guides(fill = "none", shape = "none") 

biplot_age
```

```{r}
#| label: Individuals only plots, to see if age class occupies different trait space

fviz_pca_ind(
  pca_prcomp,
  habillage = traits_NiN_age$age_class,
  addEllipses = TRUE,
  repel = TRUE
)+
  scale_color_manual(values = AGE)


#Trait only, to see which traits drive the PCA
fviz_pca_var(
  pca_prcomp,
  repel = TRUE
)
```

This model is not logical as NiN and age class are categorical variables. Need to figure out which model could be more suitable to answer research question.

```{r}
#| label: glm model with NiN type as resonse variable to the traits?


#testtraits <- lmerTest::lmer(calluna_NiN_type ~ SLA_scaled * LDMC_scaled * height_cm_scaled * leaf_thickness_mean_scaled * leaf_area_scaled  + (1|plant_id), data=cleaned_data, na.action = na.exclude)

```

# Presenting results wet mass

```{r}
#| label: Plots to visualize wet mass emmeans output and dataframes for other plots


emmip(emmeans_lme4_wet_mass, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_wet_mass, comparisons=TRUE) #These red arrows are specially calculated so that any overlap means "not significant" and no overlap means "significant."



#dataframe with multicomp, adding significance groups from emmeans
cld_wet_mass <- cld(
  emmeans_lme4_wet_mass,
  adjust = "tukey",
  Letters = letters
)

multicomp_wet_mass_df <- left_join(
  df_sum_wet_mass,
  cld_wet_mass %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_wet_mass)
contrasts_wet_mass_df <- as.data.frame(pairs(emmeans_lme4_wet_mass))
head(contrasts_wet_mass_df)

contrasts_wet_mass_df <- contrasts_wet_mass_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: point plot with error bars for wet mass 

point_plot_wet_mass <- ggplot(
  multicomp_wet_mass_df,
  aes(
    x = age_class,
    y = mean_wet_mass,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_wet_mass - se_wet_mass,
      ymax = mean_wet_mass + se_wet_mass
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_wet_mass + se_wet_mass,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean wet mass (g) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Wet mass (g)", fill="NiN type")

point_plot_wet_mass

# Save with specific dimensions and high DPI
ggsave("point_plot_wet_mass.png", plot = point_plot_wet_mass, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Barchart wet mass

ggplot(multicomp_wet_mass_df, 
       aes(x = age_class, y = mean_wet_mass, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_wet_mass + se_wet_mass,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean wet mass (g) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Wet mass (g)", fill="NiN type")
```

```{r}
#| label: Plot with circles and different levels of significance wet mass
#showing the size of the estimated differences between the age class and NiN type, a bit to noisy for presenting results

ggplot(contrasts_wet_mass_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )
  )+
  labs(title="Mean wet mass (g) of Calluna vulgaris across age class and NiN type")+
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  )
```

```{r}
#| label: beeswarm/ quasirandom plots for wet mass raw data with overlay of statistical tests


#Place letters above the tallest datapoint in each age class


cld_wet_mass <- cld_wet_mass %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.0075,
      age_class == "Building" ~ 0.0075,
      age_class == "Mature"   ~ 0.0075
    )
  )


ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color=calluna_NiN_type)) +
   geom_quasirandom(dodge.width = 1, 
                        alpha = 0.5,
                        size = 1.3,
                        width = 0.15)+
  geom_violin(
    aes(fill = calluna_NiN_type),
    position = position_dodge(width = 1),
    alpha = 0.1,
    trim = TRUE,
    scale = "width",
    show.legend = FALSE
  ) +
  scale_fill_manual(values = NIN) +
geom_pointrange(data=cld_wet_mass, 
                 aes(x=age_class, y=emmean, ymin=lower.CL, ymax=upper.CL, group=calluna_NiN_type), 
                color="black",
                size=0.8,
                fatten=2,
                position=position_dodge(width=1))+
   geom_text(
     data = cld_wet_mass,
     aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
     position = position_dodge(width = 1),   # Aligns letters with the pointranges
     vjust = -0.5,                             # Nudges letters slightly upward
     color = "black",
     fontface= "bold",
     show.legend = FALSE
  )+
   scale_color_manual(values = NIN) +
   labs(x="Age class", y="Wet mass (g)", title="Mean leaf wet mass (g) of C. vulgaris across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )


```

```{r}
#| label: Raincloud plot with half violins, jittered points and EMMs wet mass



ggplot(cleaned_data,
       aes(x = age_class,
           y = wet_mass,
           fill = calluna_NiN_type,
           colour = calluna_NiN_type)) +
  
  # Half violin ("cloud")
  stat_halfeye(
    adjust = 0.5,
    width = 0.4,
    justification = -0.25,   # moves violin left
    point_colour = NA,
    alpha = 0.15,
    position = position_dodge(width = 0.8)
  ) +
  
  # Raw points ("rain")
  geom_quasirandom(
    dodge.width = 0.8,
    width = 0.12,
    alpha = 0.4,
    size = 1
  ) +
  
  # EMMs
  geom_pointrange(
    data = cld_wet_mass,
    aes(
      x = age_class,
      y = emmean,
      ymin = lower.CL,
      ymax = upper.CL,
      group = calluna_NiN_type
    ),
    colour = "black",
    size = 0.7,
    fatten = 2,
    position = position_dodge(width = 0.8),
    inherit.aes = FALSE
  ) +
  
  # CLD letters
  geom_text(
    data = cld_wet_mass,
    aes(
      x = age_class,
      y = label_y,
      label = trimws(.group),
      group = calluna_NiN_type
    ),
    colour = "black",
    fontface = "bold",
    position = position_dodge(width = 0.8),
    inherit.aes = FALSE
  ) +
  
  scale_fill_manual(values = NIN) +
  scale_colour_manual(values = NIN) +
  
  labs(
    x = "Age class",
    y = "Wet mass (g)",
    title = "Mean leaf wet mass (g) of C. vulgaris across age class and NiN types",
    colour = "NiN type",
    fill = "NiN type"
  ) +
  
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )+
  guides(fill = "none")
```

Best visuals so far, easy overview, less noise

```{r}
#| label: Without violins and boxplots wet mass

#Removed cld letter display

results_ggp_wet_mass <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = wet_mass, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_wet_mass, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
#  geom_text(
 #   data = cld_wet_mass,
  #  aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
   # position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x=NULL, y="Wet mass (g)", title="Leaf wet mass (g) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    legend.justification = "center"
  ) +
facet_wrap(~ age_class)

results_ggp_wet_mass

results_export_wet_mass <- results_ggp_wet_mass +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_wet_mass.png", plot = results_export_wet_mass, width = 7, height = 4, dpi = 300)
```

```{r}

#| label: Violin + boxplot wet mass

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color=calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.5,
                   size = 1,
                   width = 0.15)+
  geom_boxplot(alpha = 0.35,
               width = 0.5,
               position = position_dodge(width=1))+
  geom_violin(
    aes(fill = calluna_NiN_type),
    position = position_dodge(width = 1),
    alpha = 0.15,
    trim = TRUE,
    scale = "width"
  ) +
  scale_fill_manual(values = NIN) +
  geom_pointrange(data=cld_wet_mass, 
                  aes(x=age_class, y=emmean, ymin=lower.CL, ymax=upper.CL, group=calluna_NiN_type), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  geom_text(
    data = cld_wet_mass,
    aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    position = position_dodge(width = 1),   # Aligns letters with the pointranges
    vjust = -0.5,                             # Nudges letters slightly upward
    color = "black",
    fontface= "bold",
    show.legend = FALSE
  )+
  scale_color_manual(values = NIN) +
  labs(x="Age class", y="Wet mass (g)", title="Mean leaf wet mass (g) of" ~ italic("C. vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

Good visuals, the violin is a nice visual to see the spread of the points, but it looks a bit messy with boxplot and violin together, the boxplot gives more information

```{r}
#| label: With Boxplot, removing pointrange


#Making a reusable plot function

ggp1 <- function(response, cld_data, ylab, title) {

  p <- ggplot(
    cleaned_data,
    aes(
      x = age_class,
      y = .data[[response]],
      color = calluna_NiN_type
    )
  ) +
      geom_quasirandom(
        dodge.width = 1,
        alpha = 0.5,
        size = 1,
        width = 0.1
      )+
    geom_boxplot(
      aes(
        fill = calluna_NiN_type,
        group = interaction(age_class, calluna_NiN_type)
      ),
      color = "black",
      alpha = 0.35,
      width = 0.5,
      position = position_dodge(width = 1)
    ) +
    geom_text(
      data = cld_data,
      aes(
        y = label_y,
        label = trimws(.group),
        group = calluna_NiN_type
      ),
      position = position_dodge(width = 1),
      vjust = -0.5,
      color = "black",
      fontface = "bold",
      show.legend = FALSE
    ) +
    scale_color_manual(values = NIN, name = "NiN type") +
    scale_fill_manual(values = NIN, name = "NiN type") +
    labs(x = "Age class", y = ylab, title = title, color = "NiN type") +
    scale_y_continuous(
      expand = expansion(mult = c(0.02, 0.15))
    ) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
}


ggp_wet_mass <- ggp1(
  "wet_mass",
  cld_wet_mass,
  "Mean leaf wet mass (g)",
  "Mean leaf wet mass (g) of" ~ italic("C. vulgaris") ~ "across age class and NiN types"
)

ggp_wet_mass

# Save with specific dimensions and high DPI
ggsave("ggp_wet_mass.png", plot = ggp_wet_mass, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: violin and boxplot without pointrange

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color=calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.35,
                   size = 1.8,
                   width = 0.15)+
  geom_boxplot(alpha = 0.35,
               width = 0.5,
               position = position_dodge(width=1))+
  geom_violin(
    aes(fill = calluna_NiN_type),
    position = position_dodge(width = 1),
    alpha = 0.15,
    trim = TRUE,
    scale = "width"
  ) +
  geom_text(
    data = cld_wet_mass,
    aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    position = position_dodge(width = 1),   # Aligns letters with the pointranges
    vjust = -0.5,                             # Nudges letters slightly upward
    color = "black",
    fontface= "bold",
    show.legend = FALSE
  )+
scale_color_manual(values = NIN, name = "NiN type") +
scale_fill_manual(values = NIN, name = "NiN type") +
  labs(x="Age class", y="Wet mass (g)", title="Mean leaf wet mass (g) of C. vulgaris across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

```{r}
#| label: Another boxplot


ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color=calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1.8,
                   width = 0.15)+
  geom_boxplot(aes(group= interaction(age_class, calluna_NiN_type)),
               alpha = 0.35,
               width = 0.5,
               position = position_dodge(width=1)
  )+
  geom_text(
    data = cld_wet_mass,
    aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    position = position_dodge(width = 1),   # Aligns letters with the pointranges
    vjust = -0.5,                             # Nudges letters slightly upward
    color = "black",
    fontface= "bold",
    show.legend = FALSE
  )+
  scale_color_manual(values = NIN) +
  labs(x="Age class", y="Wet mass (g)", title="Mean leaf wet mass (g) of C. vulgaris across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

```

This plot is the most informative, but might be too redundant. I'll stick to the simpler plot without boxplot and significance lines

```{r}
#| label: Boxplot with raw datapoints, significance lines and letters wet mass

library(ggsignif)


# 1. Finn maks-verdien for wet_mass for å styre høyden på brakettene automatisk
max_y <- max(cleaned_data$wet_mass, na.rm = TRUE)

# 2. Data frame with manual star mapping based on your p-values
signif_wet_mass <- data.frame(
  xmin = c(
    2 - 0.33, # Building: BH
    2,        # Building: FH
    1,        # FH: Pioneer to Mature
    1 + 0.33, # TH: Pioneer to Building
    2 + 0.33  # TH: Building to Mature
  ),
  xmax = c(
    2 + 0.33, # Building: TH
    2 + 0.33, # Building: TH
    3,        # FH: Mature
    2 + 0.33, # Building: TH
    3 + 0.33  # Mature: TH
  ),
  # Height tiers to stack them neatly above the letters
  y_position = c(
    max_y * 1.30, 
    max_y * 1.45, 
    max_y * 1.60, 
    max_y * 1.75, 
    max_y * 1.90  
  ),
  # UPDATED: P-values mapped directly to star ratings
  annotation = c(
    "***", # p = 0.0005 (is < 0.001)
    "***", # p = 0.0002 (is < 0.001)
    "*",   # p = 0.0109 (is < 0.05)
    "***", # p = 0.0003 (is < 0.001)
    "***"  # p = 0.0007 (is < 0.001)
  ),
  group = 1:5 
)

# 3. Build the plot
ggp2_wet_mass <- ggplot(cleaned_data, aes(x = age_class, y = wet_mass, color = calluna_NiN_type)) +
  geom_quasirandom(
    dodge.width = 1,
    alpha = 0.5,
    size = 1,
    width = 0.1
  ) +
  geom_boxplot(
    aes(
      fill = calluna_NiN_type,
      group = interaction(age_class, calluna_NiN_type)
    ),
    color = "black",
    alpha = 0.40,
    width = 0.6,
    position = position_dodge(width = 1)
  ) +
  geom_text(
    data = cld_wet_mass,
    aes(
      y = label_y,
      label = trimws(.group),
      group = calluna_NiN_type
    ),
    position = position_dodge(width = 1),
    vjust = -0.5,
    color = "black",
    fontface = "bold",
    show.legend = FALSE
  ) +
  
  # Signif Bracket Layer
  ggsignif::geom_signif(
    data = signif_wet_mass,
    aes(
      xmin = xmin, 
      xmax = xmax, 
      y_position = y_position, 
      annotations = annotation,
      group = group
    ),
    manual = TRUE,
    inherit.aes = FALSE, 
    color = "black",
    tip_length = 0.02,
    # Increased text size slightly since stars are smaller than text string expressions
    textsize = 4.2, 
    vjust = 0.4 # Centers stars vertically right on top of the bracket line
  ) +
  
  scale_color_manual(values = NIN, name = "NiN type") +
  scale_fill_manual(values = NIN, name = "NiN type") +
  labs(
    x = "Age class", 
    y = "Mean leaf wet mass (g)", 
    title = expression("Mean leaf wet mass (g) of " ~ italic("C. vulgaris") ~ " across age class and NiN types"), 
    color = "NiN type"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.10))
  ) +
  theme_classic() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  geom_segment(
    aes(x = 1.5, xend = 1.5,
        y = 0, yend = 0.0072),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )+
  geom_segment(
    aes(x = 2.5, xend = 2.5,
        y = 0, yend = 0.0072),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )

ggp2_wet_mass

# Save with specific dimensions and high DPI
ggsave("ggp2_wet_mass.png", plot = ggp2_wet_mass, width = 7, height = 4, dpi = 300)

```

# Presenting results height

Comparing results of emmeans post-hoc test, for visualizing the magnitude and direction of differences between group means

#A heatmap of estimates does not show: variability within groups, outliers, sample distributions,overlap among observations.

#Negative values → the first group has a lower mean than the second.

#Positive values → the first group has a higher mean than the second.

#Values near 0 → the groups are similar.

```{r}
#| label: Plots to visualize height emmeans output and dataframes for other plots


emmip(emmeans_height_cm, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_height_cm, comparisons=TRUE) #These red arrows are specially calculated so that any overlap means "not significant" and no overlap means "significant."

#dataframe with multicomp, adding significance groups from emmeans
cld_height <- cld(
  emmeans_height_cm,
  adjust = "tukey",
  Letters = letters
)

multicomp_height_df <- left_join(
  df_sum_height,
  cld_height %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
contrasts_height_df <- as.data.frame(pairs(emmeans_height_cm))
head(contrasts_height_df)

contrasts_height_df <- contrasts_height_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )




```

```{r}
#| label: point plot for height with error bars 

point_plot_height <- ggplot(
  multicomp_height_df,
  aes(
    x = calluna_NiN_type,
    y = mean_height_cm,
    colour = calluna_NiN_type
  )
) +
  geom_pointrange(data=cld_height, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=calluna_NiN_type),
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  scale_color_manual(values = NIN, guide = "none") +
  scale_fill_manual(values = NIN, guide = "none") +
  geom_text(data = cld_height,
    aes(
      y = upper.CL + 2,
      label = .group,
      group = calluna_NiN_type
    ),
    position = position_dodge(width = 1),
    vjust = 0,
    show.legend = FALSE
  )+ labs(title="Vegetative height (cm) of"  ~ italic("Calluna vulgaris") ~ "across age class and NiN type",
          x="NiN type", y="Height (cm)", colour = "NiN type") +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.05))) +
  theme(plot.title = element_text(hjust = 0.5), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()) +
  facet_wrap(~age_class)

point_plot_height
  
# Save with specific dimensions and high DPI
ggsave("point_plot_height.png", plot = point_plot_height, width = 7, height = 4, dpi = 300)
```

Not so easy to read as it looks like y-axis is counting the height

```{r}
#| label: Barchart height summary

ggplot(multicomp_height_df, 
       aes(x = age_class, y = mean_height_cm, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_height_cm + se_height_cm,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean vegetative height (cm) of Calluna vulgaris across age class and NiN type",
  x="Age class", y="Height (cm)", fill="NiN type")
```

```{r}
#| label: Plot with circles and different levels of significance for height

# showing the size of the estimated differences between the age class and NiN type, a bit to noisy for presenting results

ggplot(contrasts_height_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  xlab("NIN type")+
  ylab("Age class")+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )
  )+
  labs(title="Mean height (cm) of Calluna vulgaris across age class and NiN type")+
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  )
```

```{r}
#| label: ggp1 plot height

#Place letters above the tallest datapoint in each age class

cld_height <- cld_height %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 54,
      age_class == "Building" ~ 54,
      age_class == "Mature"   ~ 54
    )
  )

ggp_height <- ggplot(
    cleaned_data,
    aes(
      x = age_class,
      y = height_cm,
      color = calluna_NiN_type,
      group = calluna_NiN_type,
      fill = calluna_NiN_type,
    )
  ) +
  stat_unique(
  geom = "point",
  position = position_dodge(width = 1),
  alpha = 0.8, 
  size = 1.8
) +
    geom_boxplot(
      aes(
        group = interaction(age_class, calluna_NiN_type)
      ),
      color = "black",
      alpha = 0.35,
      width = 0.5,
      position = position_dodge(width = 1)
    ) +
    geom_text(
      data = cld_height,
      aes(
        y = label_y,
        label = trimws(.group),
        group = calluna_NiN_type
      ),
      position = position_dodge(width = 1),
      vjust = -0.5,
      color = "black",
      fontface = "bold",
      show.legend = FALSE
    ) +
    scale_color_manual(values = NIN, name = "NiN type") +
    scale_fill_manual(values = NIN, name = "NiN type") +
    labs(x = "Age class", y = "Vegetative height (cm)", title = "Vegetative height (cm) of" ~ italic("C. vulgaris") ~ "across age class and NiN type", color = "NiN type") +
    scale_y_continuous(
      expand = expansion(mult = c(0.02, 0.15))
    ) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )


ggp_height

# Save with specific dimensions and high DPI
ggsave("ggp_height.png", plot = ggp_height, width = 7, height = 4, dpi = 300)
  

```

For the report:

```{r}
#| label: Quasirandom + pointrange + EMM results plot height


results_ggp_height <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = height_cm, color = calluna_NiN_type)) +
  stat_unique(
    geom = "point",
    position = position_dodge(width = 1),
    alpha = 0.4, 
    size = 1.8
  ) +
  geom_pointrange(data=cld_height, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
  #geom_text(
   # data = cld_height,
    #aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    #position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y="Vegetative height (cm)", title="Vegetative height (cm) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_height

results_export_height <- results_ggp_height +
  theme(plot.title = element_blank())


# Save with specific dimensions and high DPI
ggsave("results_export_height.png", plot = results_export_height, width = 7, height = 4, dpi = 300)
```

# Presenting results dry mass

```{r}
#| label: Plots to visualize dry mass emmeans output and data frames for other plots


emmip(emmeans_lme4_dry_mass, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_dry_mass, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other

#dataframe with multicomp, adding significance groups from emmeans
cld_dry_mass <- cld(
  emmeans_lme4_dry_mass,
  adjust = "tukey",
  Letters = letters
)

multicomp_dry_mass_df <- left_join(
  df_sum_dry_mass,
  cld_dry_mass %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_dry_mass)
contrasts_dry_mass_df <- as.data.frame(pairs(emmeans_lme4_dry_mass))
head(contrasts_dry_mass_df)

contrasts_dry_mass_df <- contrasts_dry_mass_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )



```

```{r}
#| label: point plot with error bars dry mass

point_plot_dry_mass <- ggplot(
  multicomp_dry_mass_df,
  aes(
    x = age_class,
    y = mean_dry_mass,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_dry_mass - se_dry_mass,
      ymax = mean_dry_mass + se_dry_mass
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_dry_mass + se_dry_mass,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean dry mass (g) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Dry mass (g)", fill="NiN type")

point_plot_dry_mass

# Save with specific dimensions and high DPI
ggsave("point_plot_dry_mass.png", plot = point_plot_dry_mass, width = 7, height = 4, dpi = 300)
```

```{r}
#| label:  Barchart dry mass
ggplot(multicomp_dry_mass_df, 
       aes(x = age_class, y = mean_dry_mass, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_dry_mass + se_dry_mass,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean dry mass (g) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Dry mass (g)", fill="NiN type")
```

```{r}
#| label: Plot with circles and different levels of significance dry mass

ggplot(contrasts_dry_mass_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )) + 
  labs(title="Mean dry mass (g) of Calluna vulgaris across age class and NiN type")+
  theme(plot.title.position = "plot") +
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  )
```

```{r}
#| label: ggp plot dry mass with automatic plot function

#Place letters above the tallest datapoint in each age class

cld_dry_mass <- cld_dry_mass %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.0035,
      age_class == "Building" ~ 0.0035,
      age_class == "Mature"   ~ 0.0035
    )
  )

ggp_dry_mass <- ggp1(
  "dry_mass",
  cld_dry_mass,
  "Mean leaf dry mass (g)",
  "Mean leaf dry mass (g) of" ~ italic("C. vulgaris") ~ "across age class and NiN types"
)+
  geom_segment(
    aes(x = 1.5, xend = 1.5,
        y = 0, yend = 0.003),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )+
  geom_segment(
    aes(x = 2.5, xend = 2.5,
        y = 0, yend = 0.003),
    inherit.aes = FALSE,
    colour = "grey80",
    linewidth = 0.3
  )

ggp_dry_mass

# Save with specific dimensions and high DPI
ggsave("ggp_dry_mass.png", plot = ggp_dry_mass, width = 7, height = 4, dpi = 300)
```

This is used in the report:

```{r}
#| label: Without violins and boxplots dry mass



results_ggp_dry_mass <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = dry_mass, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_dry_mass, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
 # geom_text(
  #  data = cld_dry_mass,
   # aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    #position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x=NULL, y="Dry mass (g)", title="Leaf dry mass (g) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN types", color="NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_dry_mass

results_export_dry_mass <- results_ggp_dry_mass +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_dry_mass.png", plot = results_export_dry_mass, width = 7, height = 4, dpi = 300)
```

# Presenting results leaf area

```{r}
#| label: Plots to visualize leaf area emmeans output and data frames for other plots 

emmip(emmeans_lme4_leaf_area, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_leaf_area, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other

#dataframe with multicomp, adding significance groups from emmeans
cld_leaf_area <- cld(
  emmeans_lme4_leaf_area,
  adjust = "tukey",
  Letters = letters
)

multicomp_leaf_area_df <- left_join(
  df_sum_leaf_area,
  cld_leaf_area %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_leaf_area)
contrasts_leaf_area_df <- as.data.frame(pairs(emmeans_lme4_leaf_area))
head(contrasts_leaf_area_df)

contrasts_leaf_area_df <- contrasts_leaf_area_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: point plot with error bars leaf area

point_plot_leaf_area <- ggplot(
  multicomp_leaf_area_df,
  aes(
    x = age_class,
    y = mean_leaf_area,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_leaf_area - se_leaf_area,
      ymax = mean_leaf_area + se_leaf_area
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_leaf_area + se_leaf_area,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean leaf area (mm^2) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Leaf area (mm^2)", fill="NiN type")

point_plot_leaf_area

# Save with specific dimensions and high DPI
ggsave("point_plot_leaf_area.png", plot = point_plot_leaf_area, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Barchart leaf area
ggplot(multicomp_leaf_area_df, 
       aes(x = age_class, y = mean_leaf_area, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_leaf_area + se_leaf_area,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean leaf_area (mm^2) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Leaf area (mm^2)", fill="NiN type")
```

```{r}
#| label: Plot with circles and different levels of significance leaf area

ggplot(contrasts_leaf_area_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )) + 
  labs(title="Mean leaf area (mm^2) of Calluna vulgaris across age class and NiN type")+
  theme(plot.title.position = "plot") +
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  )
```

```{r}
#| label: Without violins and boxplots leaf area

#Place letters above the tallest datapoint in each age class

cld_leaf_area <- cld_leaf_area %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.20,
      age_class == "Building" ~ 0.20,
      age_class == "Mature"   ~ 0.20
    )
  )




results_ggp_leaf_area <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_area, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_leaf_area, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
 # geom_text(
  #  data = cld_leaf_area,
   # aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
    #position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x= NULL, y = expression(Leaf ~ area ~ (cm^2)), color = "NiN type", title=expression("Leaf area" ~ (cm^2) ~ "of" ~ italic("C. vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_leaf_area

results_export_leaf_area <- results_ggp_leaf_area +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_leaf_area.png", plot = results_export_leaf_area, width = 7, height = 4, dpi = 300)
```

# Presenting results leaf thickness

```{r}
#| label: Plots to visualize emmeans output leaf thickness


emmip(emmeans_lme4_leaf_thickness, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_leaf_thickness, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_leaf_thickness <- cld(
  emmeans_lme4_leaf_thickness,
  adjust = "tukey",
  Letters = letters
)

multicomp_leaf_thickness_df <- left_join(
  df_sum_leaf_thickness,
  cld_leaf_thickness %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_leaf_thickness)
contrasts_leaf_thickness_df <- as.data.frame(pairs(emmeans_lme4_leaf_thickness))
head(contrasts_leaf_thickness_df)

contrasts_leaf_thickness_df <- contrasts_leaf_thickness_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: point plot with error bars leaf thickness

point_plot_leaf_thickness <- ggplot(
  multicomp_leaf_thickness_df,
  aes(
    x = age_class,
    y = mean_leaf_thickness,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_line(
    colour = "grey70",
    linewidth = 0.6,
    position=position_dodge(width = 0.9)
  ) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_leaf_thickness - se_leaf_thickness,
      ymax = mean_leaf_thickness + se_leaf_thickness
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_leaf_thickness + se_leaf_thickness,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean leaf thickness (mm) of C. vulgaris across age class and NiN type",
          x="Age class", y="Leaf thickness (mm)", fill="NiN type")

point_plot_leaf_thickness


# Save with specific dimensions and high DPI
ggsave("point_plot_leaf_thickness.png", plot = point_plot_leaf_thickness, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Barchart leaf thickness

ggplot(multicomp_leaf_thickness_df, 
       aes(x = age_class, y = mean_leaf_thickness, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_leaf_thickness + se_leaf_thickness,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean leaf thickness (mm) of Calluna vulgaris across age class and NiN type",
          x="Age class", y="Leaf thickness (mm)", fill="NiN type")
```

```{r}
#| label: Plot with circles and different levels of significance leaf thickness


ggplot(contrasts_leaf_thickness_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )) + 
  labs(title="Mean leaf thickness (mm) of Calluna vulgaris across age class and NiN type")+
  theme(plot.title.position = "plot") +
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  ) 




```

```{r}
#| label: ggp plot leaf thickness with automatic plot function

#Place letters above the tallest datapoint in each age class

cld_leaf_thickness <- cld_leaf_thickness %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.83,
      age_class == "Building" ~ 0.83,
      age_class == "Mature"   ~ 0.83
    )
  )

ggp_leaf_thickness <- ggp1(
  "leaf_thickness_mean",
  cld_leaf_thickness,
  "Mean leaf thickness (mm)",
  expression("Mean leaf thickness (mm) of" ~ italic("C. vulgaris") ~ "across age class and NiN type")
)

ggp_leaf_thickness

# Save with specific dimensions and high DPI
ggsave("ggp_leaf_thickness.png", plot = ggp_leaf_thickness, width = 7, height = 4, dpi = 300)
```

This is in the report:

```{r}
#| label: Without violins and boxplots leaf thickness
#Removed cld from plot


results_ggp_leaf_thickness <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = leaf_thickness_mean, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_leaf_thickness, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
#  geom_text(
 #   data = cld_leaf_thickness,
  #  aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
   # position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
  #)+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = "Leaf thickness (mm)", color = "NiN type", title = "Mean leaf thickness (mm) of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type")+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.7),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_leaf_thickness

results_export_leaf_thickness <- results_ggp_leaf_thickness +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_leaf_thickness.png", plot = results_export_leaf_thickness, width = 7, height = 4, dpi = 300)
```

# Presenting results SLA

```{r}
#| label: Plots to visualize emmeans output for SLA

emmip(emmeans_lme4_SLA, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_SLA, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_SLA <- cld(
  emmeans_lme4_SLA,
  adjust = "tukey",
  Letters = letters
)

multicomp_SLA_df <- left_join(
  df_sum_SLA,
  cld_SLA %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_SLA)
contrasts_SLA_df <- as.data.frame(pairs(emmeans_lme4_SLA))
head(contrasts_SLA_df)

contrasts_SLA_df <- contrasts_SLA_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: point plot with error bars SLA

point_plot_SLA <- ggplot(
  multicomp_SLA_df,
  aes(
    x = age_class,
    y = mean_SLA,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_line(
    colour = "grey70",
    linewidth = 0.6,
    position=position_dodge(width = 0.9)
  ) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_SLA - se_SLA,
      ymax = mean_SLA + se_SLA
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_SLA + se_SLA,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean SLA of Calluna vulgaris across age class and NiN type",
          x="Age class", y="SLA", fill="NiN type")
point_plot_SLA


# Save with specific dimensions and high DPI
ggsave("point_plot_SLA.png", plot = point_plot_SLA, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Barchart SLA

ggplot(multicomp_SLA_df, 
       aes(x = age_class, y = mean_SLA, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_SLA + se_SLA,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean SLA of Calluna vulgaris across age class and NiN type",
          x="Age class", y="SLA", fill="NiN type")



#| label: Plot with circles and different levels of significance SLA

ggplot(contrasts_SLA_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )) + 
  labs(title="Mean SLA of Calluna vulgaris across age class and NiN type")+
  theme(plot.title.position = "plot") +
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  ) 




```

```{r}
#| label: ggp plot SLA with automatic plot function

#Place letters above the tallest datapoint in each age class

cld_SLA <- cld_SLA %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 118,
      age_class == "Building" ~ 118,
      age_class == "Mature"   ~ 118
    )
  )

ggp_SLA <- ggp1(
  "SLA",
  cld_SLA,
  expression(SLA ~ (mm^2 ~ g^"-1")),
  expression("Specific leaf area (SLA) of" ~ italic("C. vulgaris") ~ "across age class and NiN type")
)

ggp_SLA

# Save with specific dimensions and high DPI
ggsave("ggp_SLA.png", plot = ggp_SLA, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: Without violins and boxplots SLA



results_ggp_SLA <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = SLA, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_SLA, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
#  geom_text(
 #   data = cld_SLA,
  #  aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
   # position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = expression(SLA ~ (cm^2 ~ g^-1)), color = "NiN type", title = expression("Specific leaf area" ~ (cm^2 ~ g^-1) ~ "of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.95),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_SLA

results_export_SLA <- results_ggp_SLA +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_SLA.png", plot = results_export_SLA, width = 7, height = 4, dpi = 300)
```

# Presenting results LDMC

```{r}
#| label: Plots to visualize emmeans output LDMC


emmip(emmeans_lme4_LDMC, calluna_NiN_type ~ age_class, CIs=TRUE) #If confidence intervals are largely overlapping it is more likely the groups are not siginficantly different
plot(emmeans_lme4_LDMC, comparisons=TRUE) #Overlapping arrows mean the groups are likely not significantly different from each other


#dataframe with multicomp, adding significance groups from emmeans
cld_LDMC <- cld(
  emmeans_lme4_LDMC,
  adjust = "tukey",
  Letters = letters
)

multicomp_LDMC_df <- left_join(
  df_sum_LDMC,
  cld_LDMC %>%
    dplyr::select(age_class, calluna_NiN_type, .group),
  by = c("age_class", "calluna_NiN_type")
)


#data frame for estimate values from emmeans with all interactions
pairs(emmeans_lme4_LDMC)
contrasts_LDMC_df <- as.data.frame(pairs(emmeans_lme4_LDMC))
head(contrasts_LDMC_df)

contrasts_LDMC_df <- contrasts_LDMC_df |>
  separate(
    contrast,
    into = c("calluna_NiN_type", "age_class"),
    sep = " - "
  )
```

```{r}
#| label: point plot with error bars LDMC

point_plot_LDMC <- ggplot(
  multicomp_LDMC_df,
  aes(
    x = age_class,
    y = mean_LDMC,
    colour = calluna_NiN_type,
    group = calluna_NiN_type
  )
) +
  geom_point(
    position = position_dodge(0.9),
    size = 3
  ) +
  scale_color_manual(values = NIN) +
  geom_errorbar(
    aes(
      ymin = mean_LDMC - se_LDMC,
      ymax = mean_LDMC + se_LDMC
    ),
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  geom_text(
    aes(
      y = mean_LDMC + se_LDMC,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean LDMC of Calluna vulgaris across age class and NiN type",
          x="Age class", y="LDMC", fill="NiN type")
point_plot_LDMC

# Save with specific dimensions and high DPI
ggsave("point_plot_LDMC.png", plot = point_plot_LDMC, width = 7, height = 4, dpi = 300)


#| label: Barchart LDMC


ggplot(multicomp_LDMC_df, 
       aes(x = age_class, y = mean_LDMC, group= calluna_NiN_type, fill = calluna_NiN_type,
           ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC)) +
  geom_col(position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9), width=0.2) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  geom_text(
    aes(
      y = mean_LDMC + se_LDMC,
      label = .group
    ),
    position = position_dodge(0.9),
    vjust=-0.5,
    show.legend = FALSE
  )+ labs(title="Mean LDMC of Calluna vulgaris across age class and NiN type",
          x="Age class", y="LDMC", fill="NiN type")



#| label: Plot with circles and different levels of significance LDMC


ggplot(contrasts_LDMC_df,
       aes(calluna_NiN_type, age_class)) +
  geom_point(
    aes(size = abs(estimate),
        fill = estimate),
    shape = 21
  ) +
  geom_text(
    aes(label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )),
    size = 5,
    nudge_y = 0.2)+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )) + 
  labs(title="Mean LDMC of Calluna vulgaris across age class and NiN type")+
  theme(plot.title.position = "plot") +
  coord_equal()+
  scale_fill_gradient2(
    low = "brown",
    mid = "white",
    high = "darkgreen",
    midpoint=0
  ) 




```

```{r}
#| label: ggp plot LDMC with automatic plot function

#Place letters above the tallest datapoint in each age class

cld_LDMC <- cld_LDMC %>%
  mutate(
    label_y = case_when(
      age_class == "Pioneer"  ~ 0.52,
      age_class == "Building" ~ 0.52,
      age_class == "Mature"   ~ 0.52
    )
  )

ggp_LDMC <- ggp1(
  "LDMC",
  cld_LDMC,
  expression(LDMC ~ (g ~ g^"-1")),
  expression("Leaf dry matter content (LDMC) of" ~ italic("C. vulgaris") ~ "across age class and NiN type")
)

ggp_LDMC

# Save with specific dimensions and high DPI
ggsave("ggp_LDMC.png", plot = ggp_LDMC, width = 7, height = 4, dpi = 300)
```

In the report:

```{r}
#| label: Without violins and boxplots LDMC



results_ggp_LDMC <- ggplot(cleaned_data, aes(x = calluna_NiN_type, y = LDMC, color = calluna_NiN_type)) +
  geom_quasirandom(dodge.width = 1, 
                   alpha = 0.6,
                   size = 1,
                   width = 0.3)+
  geom_pointrange(data=cld_LDMC, 
                  aes(x=calluna_NiN_type, y=emmean, ymin=lower.CL, ymax=upper.CL, group=age_class), 
                  color="black",
                  size=0.8,
                  fatten=2,
                  position=position_dodge(width=1))+
#  geom_text(
 #   data = cld_LDMC,
  #  aes(y = label_y, label = trimws(.group), group = calluna_NiN_type),       
   # position = position_dodge(width = 1),   # Aligns letters with the pointranges
    #vjust = -0.5,                             # Nudges letters slightly upward
    #color = "black",
    #fontface= "bold",
    #show.legend = FALSE
 # )+
  scale_fill_manual(values = NIN) +
  scale_color_manual(values = NIN) +
  labs(x = NULL, y = expression(LDMC ~ (g ~ g^-1)), color = "NiN type", title = expression("Leaf dry matter content" ~ (g ~ g^-1) ~ "of" ~ italic("Calluna vulgaris") ~ "across age class and NiN type"))+
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.15))
  )+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        legend.justification = "center"
  ) +
  facet_wrap(~ age_class)

results_ggp_LDMC

results_export_LDMC <- results_ggp_LDMC +
  theme(plot.title = element_blank())

# Save with specific dimensions and high DPI
ggsave("results_export_LDMC.png", plot = results_export_LDMC, width = 7, height = 4, dpi = 300)
```

```{r}
#| label: SLA and LDMC relationship


ggplot(cleaned_data,
       aes(x = SLA, y = LDMC)) +
  
  # scatter points
  geom_point(aes(color = calluna_NiN_type),
             size = 1.5, alpha = 0.4) +
  
  # fitted regression line by age class
  geom_smooth(
    aes(colour = calluna_NiN_type),
    method = "lm",
    se = TRUE,
    linewidth = 1.3,
    colour = "black"
  ) +
  
  # optional: label
  labs(
    x = "SLA",
    y = "LDMC",
    title = "SLA vs LDMC",
    color = "NiN type"
  ) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )


ggplot(cleaned_data, aes(SLA, LDMC)) +
  geom_point(
    colour = "grey70",
    alpha = 0.25,
    size = 1.5
  ) +
  geom_smooth(
    aes(colour = calluna_NiN_type),
    method = "lm",
    se = FALSE,
    linewidth = 1.3
  ) +
  facet_wrap(~ age_class) +
  scale_color_manual(values = NIN) +
  scale_fill_manual(values = NIN) 

```
