---
title: "DURIN_2026_Plant_Functional_Traits -Hydraulics_Prelimenary analysis"
author: "Maiken Anabel Ransve"
format: html
editor: visual
---

# Importing data

Have to see which packages are needed later and add them.

```{r}
#| label: load libraries

# installing / loading packages 

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(readr) # Data importation functions
library(stringr) # String and character text
library(dplyr)
library(ggplot2)

```

```{r}
#| label: importing raw data of traits
#| message: false 

# Read cleaned output file
cleaned_data <- read_csv("/Users/maikenransve/Documents/Internship_DURIN/statistical analysis/data/clean_data/cleaned_data.csv")


#Creating a data frame
cleaned_data <- as.data.frame(cleaned_data)
cleaned_data$calluna_NIN_type <- factor(cleaned_data$calluna_NIN_type)
cleaned_data$age_class <- factor(cleaned_data$age_class)

```

# Wet mass(g)

```{r}
#| label: Wet mass statistics

# Data frame with mean and se

df_sum_wet_mass <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_wet_mass = mean(wet_mass, na.rm=TRUE),
            se_wet_mass = sd(wet_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

#Histograms

ggplot(cleaned_data, aes(x = wet_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lightblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Wet mass (g)", y = "Number of Calluna individuals")


#testing for normal distribution with Q-Q plot ---data appears normal

qqnorm(cleaned_data$wet_mass)
qqline(cleaned_data$wet_mass)

#testing for normality with Shapiro-Wilk test ---gives a result of non-normal data

shapiro.test(cleaned_data$wet_mass)

#Checking residuals ---Also evidence against normality

lm_wet_mass <- lm(wet_mass ~ calluna_NIN_type * age_class, data = cleaned_data)
summary(lm_dry_mass)

qqnorm(residuals(lm_wet_mass))
qqline(residuals(lm_wet_mass))

shapiro.test(residuals(lm_wet_mass))

#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Wet mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = wet_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


#Interactive plots


ggplot(df_sum_wet_mass, aes(x = age_class, y = mean_wet_mass, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wet_mass - se_wet_mass, ymax = mean_wet_mass + se_wet_mass), width = 0.1) +
  theme_minimal()


#Plotting fitted values

#Need to find another model as data is probably not normally distributed!

newd_wet_mass <- expand.grid(calluna_NIN_type = levels(cleaned_data$calluna_NIN_type),
                    age_class = levels(cleaned_data$age_class))
newd_wet_mass$pred_wet_mass <- predict(lm_wet_mass, newd_wet_mass, se.fit = TRUE)$fit

ggplot(newd_wet_mass, aes(x = age_class, y = pred_wet_mass, color = calluna_NIN_type, group = calluna_NIN_type)) +
   geom_line() + geom_point() + theme_minimal()

#glm model

glm_wet_mass <-glm(wet_mass ~ calluna_NIN_type * age_class, family = Gamma(link = "log"), data = cleaned_data)




```

# Dry mass(g)

```{r}
#| label: Dry mass statistics

# Data frame with mean and se

df_sum_dry_mass <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_dry_mass = mean(dry_mass, na.rm=TRUE),
            se_dry_mass = sd(dry_mass, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

# Histogram

ggplot(cleaned_data, aes(x = dry_mass)) +
  geom_histogram(bins = 30, color = "black", fill = "lavender") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Dry mass (g)", y = "Number of Calluna individuals")

#testing for normal distribution with Q-Q plot

qqnorm(cleaned_data$dry_mass)
qqline(cleaned_data$dry_mass)

#testing for normality with Shapiro-Wilk test ---gives a result of non-normal data

shapiro.test(cleaned_data$dry_mass)

#Linear model Dry mass

lm_dry_mass <- lm(dry_mass ~ calluna_NIN_type * age_class, data = cleaned_data)
summary(lm_dry_mass)

#Checking residuals

qqnorm(residuals(lm_dry_mass))
qqline(residuals(lm_dry_mass))

shapiro.test(residuals(lm_dry_mass))


#Boxplot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "Dry mass (g)")

#Violin plot

ggplot(cleaned_data, aes(x = age_class, y = dry_mass, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()

#Interactive plot

ggplot(df_sum_dry_mass, aes(x = age_class, y = mean_dry_mass, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line(linewidth=1) + geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme(
  text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(color = "grey90"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
) 

#swapped axis, NiN type on x-axis

ggplot(df_sum_dry_mass, aes(x = calluna_NIN_type, y = mean_dry_mass, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dry_mass - se_dry_mass, ymax = mean_dry_mass + se_dry_mass), width = 0.1) +
  theme_minimal() + labs(
    x = "Calluna NIN type",
    y = "Mean dry mass (g)",
    color = "Age class"
)



#Plotting fitted values, needs to be changed!!!

newd_dry_mass <- expand.grid(calluna_NIN_type = levels(cleaned_data$calluna_NIN_type),
                    age_class = levels(cleaned_data$age_class))
newd_dry_mass$pred_dry_mass <- predict(lm_dry_mass, newd_dry_mass, se.fit = TRUE)$fit

ggplot(newd_dry_mass, aes(x = age_class, y = pred_dry_mass, color = calluna_NIN_type, group = calluna_NIN_type)) +
   geom_line() + geom_point() + geom_line(linewidth = 1) +
geom_point(size = 3.5)+ theme_minimal()

```

# gg plots-histograms

```{r}
#| label: Checking different histograms with traits against NIN type x age class



#Height

ggplot(cleaned_data, aes(x = height_cm)) +
  geom_histogram(bins = 30, color = "black", fill = "violet") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "Height (cm)", y = "Number of Calluna individuals")


#Leaf thickness

data_thickness <- cleaned_data %>% 
  pivot_longer( # to get the separate columns reshaped
    cols = c(leaf_thickness_1, leaf_thickness_2, leaf_thickness_3),
    names_to = "replicate",
    values_to = "leaf_thickness"
  )

ggplot(data_thickness, aes(x = leaf_thickness)) +
  geom_histogram(bins = 30, color = "black", fill = "forestgreen") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "leaf thickness (mm)", y = "Number of Calluna individuals")


#leaf area

ggplot(cleaned_data, aes(x = leaf_area)) +
  geom_histogram(bins = 30, color = "black", fill = "pink") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "leaf area (mm^2))", y = "Number of Calluna individuals")


#SLA

ggplot(cleaned_data, aes(x = SLA)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "SLA", y = "Number of Calluna individuals")

#LDMC

ggplot(cleaned_data, aes(x = LDMC)) +
  geom_histogram(bins = 30, color = "black", fill = "orange") +
  facet_grid(calluna_NIN_type ~ age_class) +
  theme_minimal() +
  labs(x = "LDMC", y = "Number of Calluna individuals")
```

ggplot(df, aes(x = age_class, y = dry_mass, fill = age_class)) +

geom_violin(trim = FALSE, alpha = 0.6) +

stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +

facet_wrap(\~ calluna_NIN_type) +

theme_minimal()

# Boxplots and violin plots

```{r}
#| label: Checking boxplots and violin plots with traits against age class and NiN type




#Height

ggplot(cleaned_data, aes(x = age_class, y = height_cm, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()

#Leaf thickness

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "leaf_thickness")

ggplot(data_thickness, aes(x = age_class, y = leaf_thickness, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


#leaf area

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "leaf_area")

ggplot(cleaned_data, aes(x = age_class, y = leaf_area, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


#SLA

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "SLA")

ggplot(cleaned_data, aes(x = age_class, y = SLA, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()


#LDMC

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal() +
  labs(x = "Age class", y = "LDMC")

ggplot(cleaned_data, aes(x = age_class, y = LDMC, fill = age_class)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black") +
  facet_wrap(~ calluna_NIN_type) +
  theme_minimal()

```

# Interactive plots

```{r}
#| label: Interaction style plots



#Height

df_sum_height <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_height_cm = mean(height_cm, na.rm=TRUE),
            se_height_cm = sd(height_cm, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

ggplot(df_sum_height, aes(x = age_class, y = mean_height_cm, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  theme_minimal()

#swapped axis, NiN type on x-axis

ggplot(df_sum_height, aes(x = calluna_NIN_type, y = mean_height_cm, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_height_cm - se_height_cm, ymax = mean_height_cm + se_height_cm), width = 0.1) +
  theme_minimal()

#leaf thickness

df_sum_leaf_thickness <- data_thickness %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_leaf_thickness = mean(leaf_thickness, na.rm=TRUE),
            se_leaf_thickness = sd(leaf_thickness, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

ggplot(df_sum_leaf_thickness, aes(x = age_class, y = mean_leaf_thickness, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
  theme_minimal()

#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_thickness, aes(x = calluna_NIN_type, y = mean_leaf_thickness, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_thickness - se_leaf_thickness, ymax = mean_leaf_thickness + se_leaf_thickness), width = 0.1) +
  theme_minimal()

#leaf area

df_sum_leaf_area <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_leaf_area = mean(leaf_area, na.rm=TRUE),
            se_leaf_area = sd(leaf_area, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

ggplot(df_sum_leaf_area, aes(x = age_class, y = mean_leaf_area, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  theme_minimal()

#swapped axis, NiN type on x-axis

ggplot(df_sum_leaf_area, aes(x = calluna_NIN_type, y = mean_leaf_area, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_leaf_area - se_leaf_area, ymax = mean_leaf_area + se_leaf_area), width = 0.1) +
  theme_minimal()


#SLA

df_sum_SLA <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_SLA = mean(SLA, na.rm=TRUE),
            se_SLA = sd(SLA, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

ggplot(df_sum_SLA, aes(x = age_class, y = mean_SLA, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
  theme_minimal()

#swapped axis, NiN type on x-axis

ggplot(df_sum_SLA, aes(x = calluna_NIN_type, y = mean_SLA, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_SLA - se_SLA, ymax = mean_SLA + se_SLA), width = 0.1) +
  theme_minimal()


#LDMC

df_sum_LDMC <- cleaned_data %>%
  group_by(calluna_NIN_type, age_class) %>%
  summarise(mean_LDMC = mean(LDMC, na.rm=TRUE),
            se_LDMC = sd(LDMC, na.rm=TRUE)/sqrt(n()),
            n = n(), .groups = "drop")

ggplot(df_sum_LDMC, aes(x = age_class, y = mean_LDMC, group = calluna_NIN_type, color = calluna_NIN_type)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  theme_minimal()

#swapped axis, NiN type on x-axis

ggplot(df_sum_LDMC, aes(x = calluna_NIN_type, y = mean_LDMC, group = age_class, color = age_class)) +
  geom_line() + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_LDMC - se_LDMC, ymax = mean_LDMC + se_LDMC), width = 0.1) +
  theme_minimal()



```

# Fitted models

Don't know if this works well, the models look exactly the same as the interactive plots.

```{r}

#| label: Creating fitted models






#Linear model SLA

lm_SLA <- lm(SLA ~ calluna_NIN_type * age_class, data = cleaned_data)
summary(lm_SLA)

#Plotting fitted values

newd_SLA <- expand.grid(calluna_NIN_type = levels(cleaned_data$calluna_NIN_type),
                    age_class = levels(cleaned_data$age_class))
newd_SLA$pred_SLA <- predict(lm_SLA, newd_SLA, se.fit = TRUE)$fit

ggplot(newd_SLA, aes(x = age_class, y = pred_SLA, color = calluna_NIN_type, group = calluna_NIN_type)) +
   geom_line() + geom_point() + theme_minimal()

```
